# Getting Started with R {#Rgettingstarted}


```{r include=FALSE}
opts_chunk$set(echo=TRUE, warning=FALSE, message=FALSE)
```

The first order of business would involve installing R. 

1. Download R for your computer based on its operating system from [https://cloud.r-project.org/](https://cloud.r-project.org/) - Select base R. Mac users may have to download the version compatible with the version of the Mac OSX of your laptop.
2. Now install the software. (If using Windows, use the 64 bit version if that option is provided.) Once R is installed, open it. A console should open. Once you verify this, you can close the console and get out of the program. There will be no need to open this in the future.
3. Get an IDE. I prefer to use R-Studio desktop. This can be downloaded from [R-Studio's website](https://www.rstudio.com/products/rstudio/download/#download). Don't forget to download the version compatible with your operating system. Install the software and open it. Whenever you open R-Studio desktop, it provides access to the R-interface through one of its panels. This is the reason we don't have to open the R-console separately. (See point #2 from above.)  

## R-Studio Interface

The R-Studio interface is generally divided into 4 quadrants.

1.  **Source:** This is where you write your scripts. This is usually the top-left quadrant.
2.  **Console:** This is where you see the execution of the commands and the output. This is usually the bottom-left quadrant.
3.  **Environment/History:** This is where you can see the variables that are currently in the R environment and the history of the commands that you have executed. This is usually the top-right quadrant.
4.  **Files/Plots/Packages/Help/Viewer:** This is where you can see the files in your working directory, the plots that you have generated, the packages that are installed, and the help documentation. This is usually the bottom-right quadrant.


## Organizing projects

1. It is a good idea to organize analysis projects into separate folders of their own. Any output of the analysis and the reports you generate would also end up being saved in that folder.

2. Naming conventions for folders, files, and variables: Avoid having any special characters and or spaces for names of folders, files and variables. Variable names can have underscores between words, for example, `University_Standing` or `UniversityStanding` would be perfectly fine. Use names that can help **you** and **any reader** of the report make sense of what it is that you are studying. Graham Williams has a good style guide for how to code in R. It can be downloaded from [onepager.togaware.com](http://onepager.togaware.com).

3. After opening R-Studio, always check on the working directory and set it to the specific location within the folder intended for the project. 

## Installing Packages from CRAN

```{r eval=FALSE}
install.packages("pkgname")
install.packages(c("pkgname1","pkgname2")) # For two or more packages 
```

## Installing a package from GitHub

```{r eval=FALSE}
if (!require("devtools")) install.packages("devtools") # Checks if the devtools package is available and if not, installs it from CRAN
devtools::install_github("githubaccountname/packagerepository")
```

## Installed packages should be loaded for them to offer their functions for use

```{r eval=FALSE}
library(pkgname)
```

## Setting your working directory

This helps in organizing the inputs/outputs of a project. 

* The current working directory can be determined using `getwd()`
* A new working directory can be set using `setwd(path to directory)`
* R-Studio makes this much easier to do using their GUI using *Session -> Set Working Directory* option
* Using *Projects* in R-Studio also makes the process of switching working directories between different projects easier. 

## Reading data from different sources

Load R file

```{r eval=FALSE}
load("filename.Rda")
```

Delimited files text files

```{r eval=FALSE}
dataname=read.csv("datafilename.csv")# use ?read.csv to see options
```

Excel Files

```{r eval=FALSE}

install.packages("readxl") # https://github.com/hadley/readxl
library(readxl)
read_excel("excelfilename.xls") # reads xlsx extensions also
read_excel("excelfilename.xlsx", sheet = "nameofsheet")
read_excel("excelfilename.xlsx", sheet = 3)# for third sheet, as an example
read_excel("my-spreadsheet.xls", na = "NA") # if NAs are represented by non-blank cells, specify them with na=
```

SPSS/SAS/Stata files

```{r eval=FALSE}
install.packages("haven") # https://github.com/hadley/haven
library(haven)
read_sas("path/to/file") # SAS file
read_por("path/to/file") # SPSS por files
read_sav("path/to/file") # SPSS sav files
read_dta("path/to/file")# Stata
```

Many others

```{r eval=FALSE}
?read.table
?readClipboard
# Checkout the readr package from Hadley: https://github.com/hadley/readr
```

## Saving data

```{r eval=FALSE}
save(dataframename,file="filename.Rda") # To R data
write.csv(dataframename,file="myfilename.csv") # to csv...see also ?write.csv
write.table(dataframename,file="myfilename.tsv",sep="\t") # to tab-separated file...see also ?write.table
```

## Creating a dataframe

```{r}
Person=c("A","B","C","D","E")
Age=c(15,20,25,30,35)
Height=c(60,63,75,79,56)
page=data.frame(Person,Age,Height)
```

