# Getting Started with R {#Rgettingstarted}


```{r include=FALSE}
opts_chunk$set(echo=TRUE, warning=FALSE, message=FALSE)
```

The first order of business would involve installing R.

1. Download R for your computer based on its operating system from [https://cloud.r-project.org/](https://cloud.r-project.org/) - Select base R. Mac users may have to download the version compatible with the version of the Mac OSX of your laptop.
2. Now install the software. (If using Windows, use the 64 bit version if that option is provided.) Once R is installed, open it. A console should open. Once you verify this, you can close the console and get out of the program. There will be no need to open this in the future.
3. Get an IDE. I prefer to use RStudio desktop (now by [Posit](https://posit.co/)). This can be downloaded from [Posit's website](https://posit.co/download/rstudio-desktop/). Don't forget to download the version compatible with your operating system. Install the software and open it. Whenever you open RStudio desktop, it provides access to the R-interface through one of its panels. This is the reason we don't have to open the R-console separately. (See point #2 from above.)

## The RStudio Interface

RStudio is divided into four panes:

- **Source** (top-left): Where you write and edit R scripts and R Markdown documents.
- **Console** (bottom-left): Where R commands are executed and output is displayed.
- **Environment/History** (top-right): Shows objects in your workspace and command history.
- **Files/Plots/Packages/Help/Viewer** (bottom-right): A multi-purpose pane for file browsing, viewing plots, managing packages, and reading help documentation.

## Organizing projects

1. It is a good idea to organize analysis projects into separate folders of their own. Any output of the analysis and the reports you generate would also end up being saved in that folder.

2. Naming conventions for folders, files, and variables: Avoid having any special characters and or spaces for names of folders, files and variables. Variable names can have underscores between words, for example, `University_Standing` or `UniversityStanding` would be perfectly fine. Use names that can help **you** and **any reader** of the report make sense of what it is that you are studying.

3. After opening RStudio, always check on the working directory and set it to the specific location within the folder intended for the project.

4. **Use RStudio Projects**: The recommended way to organize your work is with RStudio Projects (`.Rproj` files). Go to *File -> New Project* to create one. Projects automatically set the working directory, keep your files organized, and integrate well with version control systems like Git.

## Installing Packages from CRAN

```{r eval=FALSE}
install.packages("pkgname")
install.packages(c("pkgname1","pkgname2")) # For two or more packages
```

### Installing the tidyverse

The tidyverse is a collection of packages that we will use extensively throughout this book. Install it with:

```{r eval=FALSE}
install.packages("tidyverse")
```

This single command installs ggplot2, dplyr, tidyr, readr, purrr, tibble, stringr, forcats, and lubridate, among others. You can also install additional useful packages:

```{r eval=FALSE}
install.packages(c("readxl", "haven", "knitr", "rmarkdown", "kableExtra"))
```

## Installing a package from GitHub

```{r eval=FALSE}
if (!require("devtools")) install.packages("devtools")
devtools::install_github("githubaccountname/packagerepository")
```

## Loading packages

Installed packages should be loaded for them to offer their functions for use.

```{r eval=FALSE}
library(pkgname)
```

Loading the tidyverse loads the core packages together:

```{r eval=FALSE}
library(tidyverse)
```

You will see a startup message listing the attached packages and any function name conflicts with base R. This is normal.

## Setting your working directory

This helps in organizing the inputs/outputs of a project.

* The current working directory can be determined using `getwd()`
* A new working directory can be set using `setwd(path to directory)`
* RStudio makes this much easier to do using their GUI using *Session -> Set Working Directory* option
* Using *Projects* in RStudio also makes the process of switching working directories between different projects easier.

## Reading data from different sources

### Base R approach

Load R file:

```{r eval=FALSE}
load("filename.Rda")
```

Delimited text files:

```{r eval=FALSE}
dataname <- read.csv("datafilename.csv") # use ?read.csv to see options
```

### Tidyverse approach (readr)

The **readr** package (part of the tidyverse) provides faster and more consistent functions for reading data. See Chapter \@ref(dataimport) for a comprehensive treatment.

```{r eval=FALSE}
library(readr)
dataname <- read_csv("datafilename.csv") # returns a tibble
```

### Excel Files

```{r eval=FALSE}
library(readxl)
read_excel("excelfilename.xls") # reads xlsx extensions also
read_excel("excelfilename.xlsx", sheet = "nameofsheet")
read_excel("excelfilename.xlsx", sheet = 3) # for third sheet
read_excel("my-spreadsheet.xls", na = "NA") # specify NA representation
```

### SPSS/SAS/Stata files

```{r eval=FALSE}
library(haven)
read_sas("path/to/file") # SAS file
read_por("path/to/file") # SPSS por files
read_sav("path/to/file") # SPSS sav files
read_dta("path/to/file") # Stata
```

### Many others

```{r eval=FALSE}
?read.table
# The readr package provides many other options: https://readr.tidyverse.org/
```

## Saving data

### Base R

```{r eval=FALSE}
save(dataframename, file = "filename.Rda") # To R data
write.csv(dataframename, file = "myfilename.csv") # to csv
write.table(dataframename, file = "myfilename.tsv", sep = "\t") # to tab-separated file
```

### Tidyverse (readr)

```{r eval=FALSE}
library(readr)
write_csv(dataframename, file = "myfilename.csv") # faster, no row names by default
write_rds(dataframename, file = "myfilename.rds") # single R object, preserves types
```

## Creating a data frame

### Base R

```{r}
Person <- c("A", "B", "C", "D", "E")
Age <- c(15, 20, 25, 30, 35)
Height <- c(60, 63, 75, 79, 56)
page <- data.frame(Person, Age, Height)
page
```

### Tidyverse (tibble)

```{r}
library(tibble)
page_tibble <- tibble(
  Person = c("A", "B", "C", "D", "E"),
  Age = c(15, 20, 25, 30, 35),
  Height = c(60, 63, 75, 79, 56)
)
page_tibble
```

You can also create small tibbles row-by-row using `tribble()`:

```{r}
page_tribble <- tribble(
  ~Person, ~Age, ~Height,
  "A",      15,    60,
  "B",      20,    63,
  "C",      25,    75,
  "D",      30,    79,
  "E",      35,    56
)
page_tribble
```

## R basics: Objects and assignment

R stores data in objects. You assign values to objects using `<-` (preferred) or `=`:

```{r}
x <- 10
y <- "hello"
z <- c(1, 2, 3, 4, 5)
```

### Common data types

```{r}
# Numeric
num <- 42.5
class(num)

# Character (string)
name <- "Gonzaga"
class(name)

# Logical (boolean)
flag <- TRUE
class(flag)

# Vector (collection of same type)
nums <- c(10, 20, 30, 40, 50)
names <- c("Alice", "Bob", "Charlie")
```

### Useful functions for exploring objects

```{r eval=FALSE}
class(x)     # What type of object?
length(x)    # How many elements?
str(x)       # Structure of the object
typeof(x)    # Internal storage type
is.numeric(x) # Is it numeric?
```

## Getting help

```{r eval=FALSE}
?mean              # Help page for a specific function
help(mean)         # Same as above
??regression       # Search help pages for a keyword
help.search("regression") # Same as above
```

In RStudio, you can also press **F1** with your cursor on a function name to open its help page.

