# Describing Data {#describedata}

In this chapter, we learn how to describe variables and datasets (a.k.a data frames). We cover both base R and tidyverse approaches.

These can expedite your understanding of the data with very little coding.

```{r include=FALSE}
library(tidyverse)
opts_chunk$set(echo=TRUE, warning=FALSE, message=FALSE)
```

## Number of rows and columns (in that sequence)

```{r}
dim(iris) # dimensions
```

## Number of rows

```{r}
nrow(iris)
```

## Number of columns

```{r}
ncol(iris)
```

## Names of variables


```{r}
names(iris) # colnames(iris) also gives that information
```

## First 6 rows

```{r}
head(iris)
```

## First 2 rows

```{r}
head(iris, 2) # alternately, can use iris[1:2,]
```

## Last 6 rows

```{r}
tail(iris) # Number of rows can be controlled, see earlier example involving the head command
```

## First row

```{r}
iris[1,]
```

## First row, first column

```{r}
iris[1,1]
```

## Name of third column

```{r}
names(iris)[3]
```

## 3 entries from third column

```{r}
head(iris[3], 3) # alternately, can use iris[1:3,3]
```

## Structure of the dataframe

There are different types of variables. For the most part, we will be concerned with `character`, `factor`, and `numeric` variables. We can broadly categorize nominal and ordinal variables as `factor` variables, few of which may be ordered (for ordinal-scaled variables). Interval and Ratio scaled variables could be classified as being `numeric`. The `str` command can be a helpful command to detect the state of the variables.


```{r}
str(iris)
```

Different levels of a factor variable can be determined using the `levels` command.

```{r}
levels(iris$Species)
```

## Summary of the dataframe


When you ask for the `summary`, the results can provide a cue to the nature of variables in the data. For factor variables, the results are the counts of different categories. For numeric variables, one can get information on the mean, median and distributional information.

```{r}
summary(iris) # for factor/categorical variables, this gives a count of all categories
```

## Summary of a variable in the dataframe

```{r}
summary(iris$Sepal.Length)
```

## Tidyverse approach: `glimpse()` and `tibble`

The tidyverse provides additional tools for exploring data. The `glimpse()` function from **dplyr** is a useful alternative to `str()` --- it shows every column in a transposed format, making it easy to scan datasets with many variables.

```{r}
glimpse(iris)
```

### Converting to a tibble

Tibbles are a modern version of data frames (see Chapter \@ref(tidyverse)). Converting data to a tibble gives you a cleaner print output:

```{r}
iris_tbl <- as_tibble(iris)
iris_tbl
```

Notice how the tibble only shows the first 10 rows and indicates the dimensions. This is much friendlier than base R's default of printing the entire data frame.

### Selecting and viewing with dplyr

You can also use **dplyr** functions to explore specific parts of the data:

```{r}
# View specific columns
iris_tbl %>% select(Species, Sepal.Length) %>% head()
```

```{r}
# Count observations per group
iris_tbl %>% count(Species)
```

```{r}
# Quick summary by group
iris_tbl %>%
  group_by(Species) %>%
  summarize(
    n = n(),
    mean_sepal_length = mean(Sepal.Length),
    mean_petal_length = mean(Petal.Length)
  )
```

### The `skimr` package

For a more comprehensive overview, the **skimr** package provides a rich summary:

```{r eval=FALSE}
# install.packages("skimr")
library(skimr)
skim(iris)
```

This produces a detailed summary with histograms for numeric variables and frequency tables for factors --- all in a single function call.


