# Describing Data {#describedata}

Getting to know your data is a lot like a first date -- but with numbers instead of awkward small talk. Before you run any fancy analysis, you need to sit down with your dataset and figure out what you are working with. How big is it? What kind of variables are in there? Are there any red flags lurking beneath the surface?

Think of this chapter as your pre-meeting prep before a big client call. You would never walk into a pitch without knowing what is in the deck, and you should never run an analysis without knowing what is in the data. The functions we cover here are your detective toolkit -- quick, low-effort commands that reveal a surprising amount about your data before you do any real work.

```{r include=FALSE}
library(tidyverse)
opts_chunk$set(echo=TRUE, warning=FALSE, message=FALSE)
```

## How big is this thing? Rows and columns

First question on any first date with data: how big are you? The `dim()` function gives you the number of rows and columns, in that order. Think of rows as individual records -- customers, transactions, survey responses -- and columns as the variables you measured about each one. If someone hands you a customer database, `dim()` tells you "we have 10,000 customers and 15 attributes about each of them." That is useful context before you do anything else.

```{r}
dim(iris) # dimensions
```

## Just the row count

Sometimes you only care about how many observations (rows) you have. Maybe you are checking if your survey got enough responses to be meaningful, or verifying that your data import did not accidentally drop half your records. The `nrow()` function is your quick headcount.

```{r}
nrow(iris)
```

## Just the column count

And sometimes you want to know how many variables (columns) are in the dataset. This is particularly useful when someone hands you a spreadsheet with "just a few fields" and it turns out to be 87 columns wide. (This happens more than you would think in corporate data.)

```{r}
ncol(iris)
```

## What are the variables called?

Before you can analyze anything, you need to know what the columns are named. The `names()` function is like reading the menu before you order -- you want to know what is available before you start making requests.

```{r}
names(iris) # colnames(iris) also gives that information
```

## Peeking at the first few rows

The `head()` function shows you the first 6 rows of your data. This is one of the most-used commands in R for good reason -- it gives you a quick sense of what the data actually looks like without dumping the entire thing into your console. It is the data equivalent of skimming the executive summary before reading a 40-page report.

```{r}
head(iris)
```

## Just a taste -- first 2 rows

If 6 rows feels like too much commitment (hey, we get it), you can specify exactly how many you want to see. Just pass a number as the second argument.

```{r}
head(iris, 2) # alternately, can use iris[1:2,]
```

## Checking the end -- last 6 rows

The `tail()` function is `head()`'s less popular but equally important sibling. It shows you the last rows of your dataset. Why bother? Because this is how you catch problems like data imports that got cut off early, or garbage rows that snuck in at the bottom of a CSV. Trust but verify, as they say.

```{r}
tail(iris) # Number of rows can be controlled, see earlier example involving the head command
```

## Grabbing a specific row

Need to look at one particular record? Maybe a customer complaint came in and you want to pull up their exact data point. You can index directly into the data frame. Row 1, all columns:

```{r}
iris[1,]
```

## A single cell -- row 1, column 1

You can drill all the way down to a single value. This is like zooming into one cell of a spreadsheet -- useful when you need to verify a specific number that does not look right in a report.

```{r}
iris[1,1]
```

## What is the third column called?

If you know a column by its position but not its name (happens more often than you would think, especially when you inherit someone else's dataset), you can look it up:

```{r}
names(iris)[3]
```

## Pulling a few entries from a specific column

Here we grab the first 3 values from the third column. Useful for a quick sanity check -- are these numbers? Text? Something weird like "N/A" spelled out as a string? Five minutes of sanity checking now saves five hours of debugging later.

```{r}
head(iris[3], 3) # alternately, can use iris[1:3,3]
```

## Understanding the structure of your data

This is where the detective work gets real. The `str()` function tells you the data type of every variable in your data frame. In the business world, you absolutely need to know the difference between a number (like revenue), a category (like customer segment), and text (like a product name) -- because R treats them very differently, and so should you.

For the most part, we will be concerned with `character`, `factor`, and `numeric` variables. Think of `factor` variables as categories -- things like customer segments, geographic regions, or satisfaction ratings on a 1-to-5 scale. `Numeric` variables are your quantitative measures -- revenue, age, order count, and so on. The `str` command is your go-to for figuring out what R thinks each variable is (which, fair warning, is not always what you think it should be).

```{r}
str(iris)
```

If you have a factor variable and want to see all its categories, the `levels()` function has you covered:

```{r}
levels(iris$Species)
```

## The quick-and-dirty summary

The `summary()` function is like asking your data to introduce itself at a networking event. For numeric variables, it gives you the mean, median, min, max, and quartile values. For factor variables, it counts how many observations fall into each category. One function call and you already know a surprising amount. If data analysis were speed dating, `summary()` would be your opening line.

```{r}
summary(iris) # for factor/categorical variables, this gives a count of all categories
```

## Summarizing a single variable

You can also focus on just one variable at a time. Imagine you only care about one KPI -- say average order value or customer satisfaction score. Here is how you would zero in on it:

```{r}
summary(iris$Sepal.Length)
```

## Tidyverse approach: `glimpse()` and `tibble`

Alright, time to upgrade. Everything we have covered so far uses base R, which gets the job done perfectly well. But the tidyverse provides some slicker, more modern ways to explore your data. The `glimpse()` function from **dplyr** is like `str()` but better-looking -- it shows every column in a transposed format, making it easy to scan datasets that have a lot of variables. Think of it as the LinkedIn profile version of `str()` -- same information, better presentation.

```{r}
glimpse(iris)
```

### Converting to a tibble

Tibbles are a modern, improved version of data frames (see Chapter \@ref(tidyverse)). They print more cleanly and do not try to dump your entire dataset into the console. Converting to a tibble is like switching from a cluttered Excel spreadsheet to a well-formatted dashboard -- same data, way easier to read.

```{r}
iris_tbl <- as_tibble(iris)
iris_tbl
```

Notice how the tibble only shows the first 10 rows and tells you the dimensions up front. Much friendlier than base R's default behavior of printing every single row like an overenthusiastic intern who CC'd the entire company on a reply-all.

### Selecting and viewing with dplyr

You can also use **dplyr** functions to explore specific slices of the data. Want to focus on just two columns? Or count how many observations fall into each group? These are exactly the kinds of questions you would ask about customer segments or product categories in a real business context:

```{r}
# View specific columns
iris_tbl %>% select(Species, Sepal.Length) %>% head()
```

```{r}
# Count observations per group
iris_tbl %>% count(Species)
```

```{r}
# Quick summary by group
iris_tbl %>%
  group_by(Species) %>%
  summarize(
    n = n(),
    mean_sepal_length = mean(Sepal.Length),
    mean_petal_length = mean(Petal.Length)
  )
```

### The `skimr` package

If you want the deluxe treatment -- a comprehensive overview with histograms and frequency tables all in one shot -- the **skimr** package is your friend. It is like getting the full executive summary instead of flipping through individual reports. One function call, and you have basically done an entire preliminary data audit.

```{r eval=FALSE}
# install.packages("skimr")
library(skimr)
skim(iris)
```

This produces a detailed summary with histograms for numeric variables and frequency tables for factors -- all in a single function call. If your boss asks "so what does this data look like?", this is the function that answers it.
