# Single variable Analysis {#singlevar}  


## Numeric Variable

### Histogram

```{r}
hist(iris$Sepal.Length,col="darkred",breaks=30,border = "white",xlab = "Sepal.Length") 
# use ?hist to know more about other parameters and how you can change the title
```

### Density plot


```{r}
plot(density(iris$Sepal.Length))
```

### Box plot

```{r}
boxplot(iris$Sepal.Length) ## box plot
```


### Test of Mean

```{r}
mean(iris$Sepal.Length)
```

If we had to test whether the mean of Sepal.Length was either = or < or >  a particular value, we could use a t-test.

There are three assumptions of a one-sample t-test.

1. Random sampling from a defined population
2. Interval or ratio scale of measurement 
3. Population is normally distributed

```{r}
t.test(iris$Sepal.Length,mu=5) # Ho: mu=5 
```


```{r}
t.test(iris$Sepal.Length,mu=5,alternative="greater",conf.level = .99) # Ho: mu=5 
```

Interpretation: Since the p-value is less than .01, we have evidence to suggest that the mean of Sepal.Length is greater than 5.



## Factor variable

### Bar plot

```{r}
plot(iris$Species)
```

### Pie Chart - FYI Do not use

```{r}
counts=table(iris$Species) ## get counts
labs=paste(iris$Species,counts)## create labels
pie(counts,labels=labs) ## plot
```


### One factor-variable Chi-square goodness-of-fit test

This is useful to determine whether a frequency distribution (of a factor variable) fits well an expected frequency distribution. To illustrate this, let us use the `survey` dataset (from an R packages named `MASS`). The data and their description have been provided in the folder for datasets. 

** Names of variables**: 

```{r}
load("data/survey.Rda")
names(survey)
```

** Structure of the dataframe**

```{r}
str(survey)

```


Plotting a single factor variable, say the smoking habits of students provided in the `Smoke` variable. 

```{r}
plot(survey$Smoke,col="lightblue")
```

Let us now suppose that we wanted to test whether in the observed frequency distribution can help us make the inference that in the population of students, 70% of students were non-smokers and the remaining 30% were equally split among the other three groups.

Null: The observed frequency distribution fits well with the expected frequency distribution.
Alternate: The observed frequency distribution does not fit well with the expected frequency distribution.

Assumptions of the test are discussed in: [http://www.basic.northwestern.edu/statguidefiles/gf-dist_ass_viol.html](http://www.basic.northwestern.edu/statguidefiles/gf-dist_ass_viol.html)

How many observations are present in each of the four levels of the `Smoker` variable? - Observed frequency distribution

```{r}
obsfreq=table(survey$Smoke)
obsfreq
```

Observed frequency distribution (in proportions)

```{r}
prop.table(obsfreq)
```

Expected frequency distribution (in proportions)

```{r}
expprop=c(.1,.7,.1,.1)
```

Chi-square test of goodness of fit test

```{r}
chisq.test(obsfreq,p=expprop)
```

Interpretation: Since the p-value is less than .01 (to yield a 99% confidence level), we find evidence to suggest that the two distributions are different. (The two distributions are the observed frequency distribution and the expected frequency distribution.) 

## One factor variable - t-test one sample proportion

```{r}
summary(survey$M.I)
```

Let us suppose we were testing a hypothesis that the proportion of students who would indicate their responses in the Imperial units (feet/inches, in contrast to the metric units - centimeters/meters) would be something other than 30%. We find that there are 68 of them who provide a response in that unit. Do we have evidence in favor of the hypothesis that there would be a value different from 30%?

```{r}
binom.test(table(survey$M.I),p=.3)
```

Conclusion: We do not have evidence that the proportion is different from .3 (why? checkout the p-value. and the confidence interval should not contain the observed probability.)


What if we were testing the hypothesis that the proportion will be greater than .3?

```{r}
binom.test(table(survey$M.I),p=.3,alternative = "greater")
```

Yet again, we don't have any evidence that the mean is greater than .3 (although the observed proportion is .325, it is not "statistically significant"
)
