# Single Variable Analysis {#singlevar}

```{r include=FALSE}
library(tidyverse)
opts_chunk$set(echo=TRUE, warning=FALSE, message=FALSE)
```

Before you start hunting for relationships between variables or building prediction models, you need to get to know each variable on its own. Think of it like the first week at a new job --- before you figure out how departments interact, you need to understand what each one actually does. Single variable analysis (the fancy term is "univariate analysis") is your get-to-know-you phase with the data.

This chapter covers analyzing one variable at a time using both base R and ggplot2 visualizations. We will look at numeric variables (histograms, density plots, box plots) and factor variables (bar charts, and yes, the dreaded pie chart). We will also dip into hypothesis testing --- the statistical machinery that lets you say "I am 95% confident that..." in a meeting and actually mean it. For the full ggplot2 experience, see Chapters \@ref(ggplot2basics) and \@ref(ggplot2customize).

## Numeric Variable

### Histogram

A histogram is the first date you should go on with any numeric variable. It chops your data into bins and shows you how many observations land in each one, giving you a quick feel for the shape of the distribution. Are values clustered tightly around the center? Spread out all over the place? Piling up on one side? A histogram tells you all of this in one glance.

In business terms, this is how you would visualize the distribution of customer order values. If most orders cluster around $25-$50 with a long tail of $500+ orders, that tells a very different story than a nice bell curve centered at $100.

**Base R:**

```{r}
hist(iris$Sepal.Length, col = "darkred", breaks = 30, border = "white", xlab = "Sepal.Length")
```

**ggplot2:**

```{r}
ggplot(iris, aes(x = Sepal.Length)) +
  geom_histogram(bins = 30, fill = "darkred", color = "white") +
  labs(x = "Sepal Length", y = "Count", title = "Distribution of Sepal Length") +
  theme_minimal()
```

### Density plot

If a histogram is a first date, a density plot is the second date --- smoother, more refined, and you start to see the real shape underneath. Instead of chunky bars, you get a sleek continuous curve. Density plots are especially handy when you want to overlay multiple groups on the same chart without it looking like a bar chart had a nervous breakdown.

**Base R:**

```{r}
plot(density(iris$Sepal.Length))
```

**ggplot2:**

```{r}
ggplot(iris, aes(x = Sepal.Length)) +
  geom_density(fill = "steelblue", alpha = 0.5) +
  labs(x = "Sepal Length", y = "Density", title = "Density of Sepal Length") +
  theme_minimal()
```

We can also overlay density plots by group --- and this is where they really earn their paycheck. Imagine comparing the distribution of deal sizes across your East, Central, and West sales regions, all on one clean chart:

```{r}
ggplot(iris, aes(x = Sepal.Length, fill = Species)) +
  geom_density(alpha = 0.4) +
  labs(x = "Sepal Length", y = "Density", title = "Density by Species") +
  theme_minimal()
```

### Box plot

The box plot is the executive summary of a distribution. In one compact visual, you get the median (the line in the middle of the box), the interquartile range (the box itself --- where the middle 50% of your data lives), the whiskers (the range of "normal" values), and any outliers (the lonely dots hanging out by themselves like the person who shows up to the meeting 30 minutes early). It is the data visualization equivalent of a one-page brief: everything important, nothing extra.

**Base R:**

```{r}
boxplot(iris$Sepal.Length)
```

**ggplot2:**

```{r}
ggplot(iris, aes(y = Sepal.Length)) +
  geom_boxplot(fill = "steelblue") +
  labs(y = "Sepal Length", title = "Boxplot of Sepal Length") +
  theme_minimal()
```

Box plots really earn a promotion when you compare groups side by side. Want to see how customer satisfaction scores differ across your three product lines? How about employee engagement by department? This is your chart:

```{r}
ggplot(iris, aes(x = Species, y = Sepal.Length, fill = Species)) +
  geom_boxplot() +
  labs(x = "Species", y = "Sepal Length", title = "Sepal Length by Species") +
  theme_minimal() +
  theme(legend.position = "none")
```


### Test of Mean

Now we are moving from "looking at the data" to "making claims about the data" --- and that is where hypothesis testing enters the chat. Visualizations are great for exploration, but when your VP asks "are you sure about that number?", you need something more rigorous than "well, the histogram looked pretty good."

```{r}
mean(iris$Sepal.Length)
```

If we had to test whether the mean of Sepal.Length was either = or < or > a particular value, we could use a one-sample t-test. In the real world, this is how you would answer questions like "Is our average customer satisfaction score really above 4.0, or did we just get lucky with this particular sample?" It is the difference between a gut feeling and a defensible business claim.

There are three assumptions of a one-sample t-test.

1. Random sampling from a defined population
2. Interval or ratio scale of measurement
3. Population is normally distributed

Here we test whether the mean is different from 5. The null hypothesis says "the true mean equals 5" and we are checking whether our data disagrees:

```{r}
t.test(iris$Sepal.Length,mu=5) # Ho: mu=5
```

Now let us get more specific. What if we want to test whether the mean is *greater* than 5, and we want to be 99% confident about it? That is a tougher bar to clear, but sometimes the stakes demand it:

```{r}
t.test(iris$Sepal.Length,mu=5,alternative="greater",conf.level = .99) # Ho: mu=5
```

Interpretation: Since the p-value is less than .01, we have evidence to suggest that the mean of Sepal.Length is greater than 5. In business terms, if this were your average customer spend and 5 was your break-even target, you could confidently walk into the CFO's office and say "we are clearing the bar." That is the power of a hypothesis test --- it turns "I think so" into "the data confirms it."



## Factor variable

### Bar plot

For categorical variables --- things like product categories, customer segments, regions, or department names --- bar charts are your go-to visualization. They answer the beautifully simple question: "How many of each do we have?" Not fancy, but enormously useful. You would be amazed how many business decisions start with just counting things.

**Base R:**

```{r}
plot(iris$Species)
```

**ggplot2:**

```{r}
ggplot(iris, aes(x = Species, fill = Species)) +
  geom_bar() +
  labs(x = "Species", y = "Count", title = "Count of Each Species") +
  theme_minimal() +
  theme(legend.position = "none")
```

### Pie Chart - FYI Do not use

We are legally obligated to show you how to make a pie chart, but we are also morally obligated to tell you not to use one. Pie charts are the cargo shorts of data visualization --- technically functional, but everyone in the analytics world judges you for wearing them. The core problem is that human brains are terrible at comparing angles and areas. Quick: is a 23% slice meaningfully different from a 27% slice? In a pie chart, good luck. In a bar chart, it is obvious. That said, your boss *will* ask for a pie chart at some point (it is an immutable law of corporate life), so here is how to make one. Then gently suggest a bar chart instead.

**Base R:**

```{r}
counts <- table(iris$Species)
labs <- paste(iris$Species, counts)
pie(counts, labels = labs)
```


### One factor-variable Chi-square goodness-of-fit test

Do not let the scary name fool you --- this test answers a very practical question: "Does what I actually see match what I expected to see?" This comes up constantly in business. You expected your customer base to be 70% retail and 30% wholesale --- does the data back that up? Your marketing team claimed the email campaign would produce a certain distribution of responses --- time to check the receipts.

To illustrate this, let us use the `survey` dataset (from an R package named `MASS`). The data and their description have been provided in the folder for datasets.

** Names of variables**:

```{r}
load("data/survey.Rda")
names(survey)
```

** Structure of the dataframe**

```{r}
str(survey)

```


Let us look at the smoking habits of students, stored in the `Smoke` variable. Think of this as a customer behavior survey --- "how frequently do customers use our premium features?" with categories like "Never," "Occasionally," "Regularly," and "Heavy."

```{r}
plot(survey$Smoke,col="lightblue")
```

Now suppose we wanted to test whether the observed distribution matches our expectation that 70% of students were non-smokers and the remaining 30% were equally split among the other three groups (10% each). In a business setting, this is like testing whether your actual customer demographics match the target profile your marketing campaign was designed around.

Null: The observed frequency distribution fits well with the expected frequency distribution. (In plain English: "Reality matches our assumptions.")
Alternate: The observed frequency distribution does not fit well with the expected frequency distribution. (In plain English: "Our assumptions were wrong and we need to rethink things.")

Assumptions of the test are discussed in: [http://www.basic.northwestern.edu/statguidefiles/gf-dist_ass_viol.html](http://www.basic.northwestern.edu/statguidefiles/gf-dist_ass_viol.html)

How many observations are present in each of the four levels of the `Smoker` variable? This is our observed frequency distribution --- what actually happened out in the real world:

```{r}
obsfreq=table(survey$Smoke)
obsfreq
```

Observed frequency distribution (in proportions):

```{r}
prop.table(obsfreq)
```

Expected frequency distribution (in proportions) --- what we thought would happen when we were being optimistic in the conference room:

```{r}
expprop=c(.1,.7,.1,.1)
```

Chi-square test of goodness of fit test --- the moment of truth:

```{r}
chisq.test(obsfreq,p=expprop)
```

Interpretation: Since the p-value is less than .01 (giving us a 99% confidence level), we find evidence to suggest that the two distributions are different. Our assumption about the proportions does not hold up to scrutiny. The actual breakdown of smoking habits is significantly different from the 70/10/10/10 split we hypothesized. If this were a business scenario, it would mean your assumed customer segments do not match reality --- which is valuable information, even if it is not what you wanted to hear.

## One factor variable - t-test one sample proportion

Sometimes instead of testing a whole distribution, you just want to test a single proportion. "Is the share of customers who prefer online checkout really 30%?" "Do more than half of our users open our emails?" That is what the one-sample proportion test is for --- it is simple, focused, and incredibly practical.

```{r}
summary(survey$M.I)
```

Let us suppose we were testing a hypothesis that the proportion of students who would indicate their responses in the Imperial units (feet/inches, in contrast to the metric units - centimeters/meters) would be something other than 30%. We find that there are 68 of them who provide a response in that unit. Do we have evidence in favor of the hypothesis that there would be a value different from 30%?

```{r}
binom.test(table(survey$M.I),p=.3)
```

Conclusion: We do not have evidence that the proportion is different from .3 (why? check out the p-value --- it is too large to reject the null, and the confidence interval contains .30). In business terms, if you claimed "more than 30% of our customers prefer in-store pickup," this result would tap you on the shoulder and say "not so fast --- your data does not back that up."


What if we were testing the hypothesis that the proportion will be greater than .3?

```{r}
binom.test(table(survey$M.I),p=.3,alternative = "greater")
```

Yet again, we do not have any evidence that the proportion is greater than .3. The observed proportion is .325, which *looks* bigger, but it is not "statistically significant." And this, friends, is one of the most important lessons in business analytics: just because a number is slightly above your target does not mean it is meaningfully above your target. Sample variation is real, and statistical tests are what keep us from making confident-sounding decisions based on noise.
