# Relationships Between Variables {#multivar}

```{r include=FALSE}
library(tidyverse)
opts_chunk$set(echo=TRUE, warning=FALSE, message=FALSE)
```

Welcome to the relationship chapter. Up until now, we have been looking at variables in isolation -- one at a time, like checking individual ingredients before cooking. Now we are asking the much more interesting question: how do these variables relate to each other?

Do customers who spend more also visit more often? Is there a difference in satisfaction scores across regions? Do the three marketing campaigns actually produce different conversion rates, or is the variation just noise? These are the kinds of questions that drive real business decisions, and this chapter gives you the tools to answer them.

We include both base R and ggplot2 visualizations throughout. For the full ggplot2 treatment, see Chapters \@ref(ggplot2basics) and \@ref(ggplot2customize).

## Between continuous variables: Correlations

Correlation answers the fundamental question: do these two things move together? When one goes up, does the other tend to go up too (positive correlation), go down (negative correlation), or just do its own thing (no correlation)?

In business, correlations are everywhere. Ad spend and revenue. Employee engagement and retention. Website load time and bounce rate. Understanding these relationships is the first step toward figuring out what levers to pull.

### Plotting

The scatterplots below show what different correlation strengths look like visually. Pay attention to how the points cluster around the trend line -- tighter clustering means stronger correlation.

```{r echo=FALSE}
# NumVar1=rnorm(n=50,mean=0,sd=10)
#
# NumVar2= -.9*NumVar1+rnorm(50,0,10)
#
# NumVar3=-.25*NumVar1+rnorm(50,0,10)
#
# NumVar4=rnorm(50,0,1)
#
# NumVar5=.25*NumVar1+rnorm(50,0,10)
#
# NumVar6=.9*NumVar1+rnorm(50,0,10)

# corrdata=data.frame(NumVar1,NumVar2,NumVar3,NumVar4,NumVar5,NumVar6)
# save(corrdata,file="corrdata.Rda")

load("data/corrdata.Rda")
#
#  cor(corrdata$NumVar1,corrdata$NumVar2)
#
#  cor(corrdata$NumVar1,corrdata$NumVar3)
#
#  cor(corrdata$NumVar1,corrdata$NumVar4)
#
#  cor(corrdata$NumVar1,corrdata$NumVar5)
#
#  cor(corrdata$NumVar1,corrdata$NumVar6)


par(mfrow=c(2,3))
plot(corrdata$NumVar1,corrdata$NumVar2,pch=18,col="red",xlab="Numeric Variable 1",ylab="Numeric Variable 2",main = "Correlation of -0.65")
abline( lm(corrdata$NumVar2~corrdata$NumVar1),col="blue",) # regression line (y~x)

plot(corrdata$NumVar1,corrdata$NumVar3,pch=18,col="red",xlab="Numeric Variable 1",ylab="Numeric Variable 2",main = "Correlation of -0.16")
abline( lm(corrdata$NumVar3~corrdata$NumVar1),col="blue",) # regression line (y~x)

plot(corrdata$NumVar1,corrdata$NumVar4,pch=18,col="red",xlab="Numeric Variable 1",ylab="Numeric Variable 2",main = "Correlation of 0.003")
abline( lm(corrdata$NumVar4~corrdata$NumVar1),col="blue",) # regression line (y~x)


plot(corrdata$NumVar1,corrdata$NumVar5,pch=18,col="red",xlab="Numeric Variable 1",ylab="Numeric Variable 2",main = "Correlation of 0.20")
abline( lm(corrdata$NumVar5~corrdata$NumVar1),col="blue",) # regression line (y~x)


plot(corrdata$NumVar1,corrdata$NumVar6,pch=18,col="red",xlab="Numeric Variable 1",ylab="Numeric Variable 2",main = "Correlation of 0.69")
abline( lm(corrdata$NumVar6~corrdata$NumVar1),col="blue",) # regression line (y~x)


```

### Correlations between two numeric variables

Before computing correlations, a few assumptions to keep in mind: your variables should be interval or ratio scaled, the relationship should be roughly linear (not curved), extreme outliers should not be dominating the picture, and ideally the variables are approximately normal. If these assumptions hold, you are good to go.

**Base R:**

```{r}
plot(iris$Sepal.Length, iris$Petal.Length, col = "red", xlab = "Sepal.Length", ylab = "Petal.Length")
abline(lm(iris$Petal.Length ~ iris$Sepal.Length), col = "blue")
```

**ggplot2:**

```{r}
ggplot(iris, aes(x = Sepal.Length, y = Petal.Length)) +
  geom_point(color = "red") +
  geom_smooth(method = "lm", se = TRUE, color = "blue") +
  labs(x = "Sepal Length", y = "Petal Length",
       title = "Sepal Length vs Petal Length") +
  theme_minimal()
```

We can color by species to reveal group structure -- a technique that is incredibly useful when you suspect that the overall trend might be hiding different patterns within subgroups (like how overall sales might be flat, but one region is booming while another is tanking):

```{r}
ggplot(iris, aes(x = Sepal.Length, y = Petal.Length, color = Species)) +
  geom_point() +
  geom_smooth(method = "lm", se = FALSE) +
  labs(x = "Sepal Length", y = "Petal Length",
       title = "Sepal Length vs Petal Length by Species") +
  theme_minimal()
```

Now let us put a number on it. The Pearson correlation coefficient ranges from -1 (perfect negative relationship) to +1 (perfect positive relationship), with 0 meaning no linear relationship at all.

```{r}
cor(iris$Sepal.Length,iris$Petal.Length)
```

That is a strong positive correlation. But is it statistically significant, or could this just be a fluke? The `cor.test()` function answers that by testing whether the observed correlation is different from zero:

```{r}
cor.test(iris$Sepal.Length,iris$Petal.Length)
```

We can also test a specific direction -- is the correlation greater than zero?

```{r}
cor.test(iris$Sepal.Length,iris$Petal.Length,alternative = "greater")
```



### Paired Samples - comparing two means

Sometimes you have two measurements on the same subjects and you want to know if they differ. Think of before-and-after scenarios: did the training program actually improve test scores? Did the new packaging increase purchase intent? A paired t-test handles this by focusing on the difference within each pair.

```{r}
summary(survey[,c(2,3)])
```

```{r}
t.test(survey$Wr.Hnd,survey$NW.Hnd,paired = TRUE)
```


### Relationships between more than two numeric variables

When you have several numeric variables, it helps to look at all the pairwise relationships at once. The `pairs()` function creates a scatterplot matrix -- every variable plotted against every other variable. It is like a correlation speed-dating event.

```{r}
pairs(iris[,1:4])

```


```{r}
library(car)
scatterplotMatrix(iris[,1:4])
```

And here is the actual correlation matrix -- the table of numbers behind all those scatterplots:

```{r}
cor(iris[,1:4])
```

Want to know which of those correlations are statistically significant? The `rcorr()` function from the **Hmisc** package gives you the correlation coefficients and their p-values in one shot:

```{r}
# Correlations with significance levels
library(Hmisc)
rcorr(as.matrix(iris[,1:4]), type="pearson") # type can be pearson or spearman

```

## Relationships between factor variables

What about the relationship between two categorical variables? Think: is there a relationship between gender and product preference? Between region and brand loyalty? Between job title and preferred communication channel?

Let us use the `survey` dataset and focus on the `Sex` and `Exer` (exercise frequency) variables.

```{r}
str(survey)
```

```{r}
summary(survey$Sex)
```

```{r}
summary(survey$Exer)
```

A cross-tabulation (or cross-tab) is the go-to tool here. It shows the joint frequencies of two variables -- basically a contingency table. If you have ever built a pivot table in Excel, this is the R equivalent.

```{r}
crosstab <- table(survey$Exer, survey$Sex)
crosstab
```

### Tidyverse cross-tabulation

The tidyverse can produce the same result in a more readable pipe-based workflow:

```{r}
survey %>%
  filter(!is.na(Sex), !is.na(Exer)) %>%
  count(Sex, Exer) %>%
  pivot_wider(names_from = Sex, values_from = n, values_fill = 0)
```

We can visualize this with bar charts -- either side-by-side (dodged) or stacked.

**Base R:**

```{r}
barplot(crosstab, beside = TRUE, legend = levels(unique(survey$Exer)),
        args.legend = list(x = 8, y = 70))
```

```{r}
barplot(crosstab, legend = levels(unique(survey$Exer)),
        args.legend = list(x = 1.75, y = 150, horiz = TRUE))
```

**ggplot2:**

```{r}
survey %>%
  filter(!is.na(Sex), !is.na(Exer)) %>%
  ggplot(aes(x = Sex, fill = Exer)) +
  geom_bar(position = "dodge") +
  labs(title = "Exercise Frequency by Sex", y = "Count", fill = "Exercise") +
  theme_minimal()
```

```{r}
survey %>%
  filter(!is.na(Sex), !is.na(Exer)) %>%
  ggplot(aes(x = Sex, fill = Exer)) +
  geom_bar(position = "fill") +
  labs(title = "Proportion of Exercise Frequency by Sex", y = "Proportion", fill = "Exercise") +
  scale_y_continuous(labels = scales::percent) +
  theme_minimal()
```

The visual is suggestive, but we need a formal test. The chi-square test of independence tells us whether the two variables are actually related or whether any apparent pattern is just random variation.

The null hypothesis is that the two variables are independent (no relationship). The alternative is that they are related.

```{r}
chisq.test(crosstab)
```

Interpretation: The p-value of .057 is above the typical .05 threshold, so we do not have strong enough evidence to claim a relationship between Sex and Exercise. They appear to be independent. Sometimes the data just says "these things are not connected," and that is a perfectly valid finding.



## Between factor and continuous variables

This is where things get really practical for business analysis. You want to know: does the average differ across groups? Do customers in different segments spend different amounts? Do employees in different departments have different satisfaction scores?

### Two levels of a factor level and one continuous variable - t-test

Let us compare the mean height of different genders in the `survey` data. First, the numbers:

**Base R:**

```{r}
aggregate(survey$Height, list(survey$Sex), mean, na.rm = T)
```

**Tidyverse:**

```{r}
survey %>%
  filter(!is.na(Sex), !is.na(Height)) %>%
  group_by(Sex) %>%
  summarize(mean_height = mean(Height))
```

Now let us visualize the distributions:

**Base R:**

```{r}
plot(survey$Height ~ survey$Sex)
```

**ggplot2:**

```{r}
survey %>%
  filter(!is.na(Sex), !is.na(Height)) %>%
  ggplot(aes(x = Sex, y = Height, fill = Sex)) +
  geom_boxplot() +
  geom_jitter(width = 0.2, alpha = 0.3) +
  labs(title = "Height Distribution by Sex", y = "Height (cm)") +
  theme_minimal() +
  theme(legend.position = "none")
```

Males appear to be taller than females in this sample, but "appears to be" is not good enough for a business decision. Is the difference statistically significant, or could it just be random variation in our sample? An independent samples t-test gives us the answer.

```{r}
t.test(survey$Height ~ survey$Sex, alternative = 'less', conf.level = .99, var.equal = FALSE)
```

Note that the mean height of females is lower than the mean height of males, which is why the test statistic is negative. The result suggests that the mean height of females is indeed less than the mean height of males. (The p-value is less than .01 and falls within the 99% confidence interval.) In business terms, this is the equivalent of confirming that Campaign A really did outperform Campaign B -- it is not just noise.

### More than two levels of a factor variable and one continuous variable - One-Way ANOVA

What happens when you have more than two groups to compare? You cannot just run a bunch of t-tests (well, you can, but you should not -- it inflates your error rate). Instead, you use ANOVA, which stands for Analysis of Variance. Despite the name, it is really about comparing group averages.

Let us check the mean height across different exercise levels:

**Base R:**

```{r}
aggregate(survey$Height, list(survey$Exer), mean, na.rm = T)
```

**Tidyverse:**

```{r}
survey %>%
  filter(!is.na(Exer), !is.na(Height)) %>%
  group_by(Exer) %>%
  summarize(
    n = n(),
    mean_height = mean(Height),
    sd_height = sd(Height)
  )
```

**Base R box plot:**

```{r}
plot(survey$Height ~ survey$Exer)
```

**ggplot2:**

```{r}
survey %>%
  filter(!is.na(Exer), !is.na(Height)) %>%
  ggplot(aes(x = Exer, y = Height, fill = Exer)) +
  geom_boxplot() +
  geom_jitter(width = 0.2, alpha = 0.3) +
  labs(title = "Height by Exercise Frequency",
       x = "Exercise Level", y = "Height (cm)") +
  theme_minimal() +
  theme(legend.position = "none")
```

The pattern is interesting: those who exercise frequently appear taller than those who exercise sometimes, who in turn appear taller than those who never exercise. But are these differences statistically significant, or are we just seeing noise?

Since we have more than two groups, we use ANOVA. The null hypothesis is that all group means are the same. The alternative is that at least one mean is different from the others.

```{r}
results <- aov(survey$Height ~ survey$Exer)
summary(results)
```

The p-value of .00354 tells us that at least one of the three group means is different from the others. That is statistically significant at the .01 level. Great -- but ANOVA only tells us "something is different." It does not tell us which specific groups differ from each other. It is like your sales report saying "one region underperformed" without telling you which one.

### OK, but WHICH groups are different? -- Tukey HSD

That is where the Tukey HSD (Honestly Significant Difference) test comes in. It performs pairwise comparisons between every combination of groups and tells you exactly where the significant differences are. Think of ANOVA as the alarm going off, and Tukey HSD as the detective work that follows.

```{r}
TukeyHSD(results, conf.level = .99)
```
