# Relationships Between Variables {#multivar}

```{r include=FALSE}
library(tidyverse)
opts_chunk$set(echo=TRUE, warning=FALSE, message=FALSE)
```

Now we are getting to the good stuff. Analyzing one variable at a time is like judging a basketball player by their height alone --- useful, but you are missing the whole picture. The real magic of data analysis happens when you start asking "how do these variables relate to each other?" This is where you go from describing your data to actually understanding it, and it is where the questions that keep business leaders up at night finally get answers:

- Does higher marketing spend actually lead to more sales, or are we just throwing money into the void? (Correlation)
- Do male and female employees earn different salaries? (t-test)
- Does customer satisfaction differ across our three service tiers? (ANOVA)
- Is there a relationship between customer region and product preference? (Chi-square test of independence)

This chapter covers bivariate and multivariate analysis --- fancy words for "looking at two or more variables at the same time and seeing what shakes out." We include both base R and ggplot2 visualizations throughout. For the full ggplot2 experience, see Chapters \@ref(ggplot2basics) and \@ref(ggplot2customize).

## Between continuous variables: Correlations

Correlation is the workhorse of "do these two things move together?" It produces a single number between -1 and +1 that tells you the strength and direction of a linear relationship. Here is the cheat sheet:

- **+1** = perfect positive relationship. When one goes up, the other goes up in lockstep. (Think: hours studied and exam scores.)
- **-1** = perfect negative relationship. When one goes up, the other goes down like a seesaw. (Think: price of a product and quantity demanded --- Econ 101 vibes.)
- **0** = no linear relationship whatsoever. These two variables could not care less about each other. (Think: your shoe size and your GPA.)

In business, correlations are everywhere. A marketing analyst might correlate ad spend with website traffic. An operations manager might correlate warehouse staffing levels with order fulfillment speed. A finance analyst might correlate interest rates with loan applications. The correlation coefficient gives you a quick read on whether two things are related --- and how strongly --- before you invest time in deeper analysis.

### Plotting

The best way to start is with scatterplots. Below are several examples showing what different correlation values actually look like in practice. Notice how the dots cluster tighter around the blue line as the correlation gets stronger (closer to -1 or +1), and scatter into a shapeless cloud as it approaches 0. A correlation of 0.003 is basically "I drew dots at random and drew a line through them for fun":

```{r echo=FALSE}
# NumVar1=rnorm(n=50,mean=0,sd=10)
#
# NumVar2= -.9*NumVar1+rnorm(50,0,10)
#
# NumVar3=-.25*NumVar1+rnorm(50,0,10)
#
# NumVar4=rnorm(50,0,1)
#
# NumVar5=.25*NumVar1+rnorm(50,0,10)
#
# NumVar6=.9*NumVar1+rnorm(50,0,10)

# corrdata=data.frame(NumVar1,NumVar2,NumVar3,NumVar4,NumVar5,NumVar6)
# save(corrdata,file="corrdata.Rda")

load("data/corrdata.Rda")
#
#  cor(corrdata$NumVar1,corrdata$NumVar2)
#
#  cor(corrdata$NumVar1,corrdata$NumVar3)
#
#  cor(corrdata$NumVar1,corrdata$NumVar4)
#
#  cor(corrdata$NumVar1,corrdata$NumVar5)
#
#  cor(corrdata$NumVar1,corrdata$NumVar6)


par(mfrow=c(2,3))
plot(corrdata$NumVar1,corrdata$NumVar2,pch=18,col="red",xlab="Numeric Variable 1",ylab="Numeric Variable 2",main = "Correlation of -0.65")
abline( lm(corrdata$NumVar2~corrdata$NumVar1),col="blue",) # regression line (y~x)

plot(corrdata$NumVar1,corrdata$NumVar3,pch=18,col="red",xlab="Numeric Variable 1",ylab="Numeric Variable 2",main = "Correlation of -0.16")
abline( lm(corrdata$NumVar3~corrdata$NumVar1),col="blue",) # regression line (y~x)

plot(corrdata$NumVar1,corrdata$NumVar4,pch=18,col="red",xlab="Numeric Variable 1",ylab="Numeric Variable 2",main = "Correlation of 0.003")
abline( lm(corrdata$NumVar4~corrdata$NumVar1),col="blue",) # regression line (y~x)


plot(corrdata$NumVar1,corrdata$NumVar5,pch=18,col="red",xlab="Numeric Variable 1",ylab="Numeric Variable 2",main = "Correlation of 0.20")
abline( lm(corrdata$NumVar5~corrdata$NumVar1),col="blue",) # regression line (y~x)


plot(corrdata$NumVar1,corrdata$NumVar6,pch=18,col="red",xlab="Numeric Variable 1",ylab="Numeric Variable 2",main = "Correlation of 0.69")
abline( lm(corrdata$NumVar6~corrdata$NumVar1),col="blue",) # regression line (y~x)


```

### Correlations between two numeric variables

Before running a correlation, there are a few assumptions to keep in mind: the variables should be on an interval or ratio scale (so, actual numbers --- not categories), the relationship between them should be roughly linear (not curved like a roller coaster), extreme outliers should not be hijacking the result, and ideally both variables are approximately normally distributed. Do not lose sleep over these for now, but keep them in the back of your mind like a mental sticky note.

Let us plot Sepal.Length against Petal.Length. In a business context, think of this as plotting "years of experience" against "salary" or "social media followers" against "monthly revenue."

**Base R:**

```{r}
plot(iris$Sepal.Length, iris$Petal.Length, col = "red", xlab = "Sepal.Length", ylab = "Petal.Length")
abline(lm(iris$Petal.Length ~ iris$Sepal.Length), col = "blue")
```

**ggplot2:**

```{r}
ggplot(iris, aes(x = Sepal.Length, y = Petal.Length)) +
  geom_point(color = "red") +
  geom_smooth(method = "lm", se = TRUE, color = "blue") +
  labs(x = "Sepal Length", y = "Petal Length",
       title = "Sepal Length vs Petal Length") +
  theme_minimal()
```

We can color by species to reveal group structure --- and this is a genuinely important technique. Sometimes the overall trend masks completely different patterns within subgroups. Imagine overall company revenue is flat, but when you break it down, the East region is booming while the West is tanking. The aggregate number hid the real story. Same idea here:

```{r}
ggplot(iris, aes(x = Sepal.Length, y = Petal.Length, color = Species)) +
  geom_point() +
  geom_smooth(method = "lm", se = FALSE) +
  labs(x = "Sepal Length", y = "Petal Length",
       title = "Sepal Length vs Petal Length by Species") +
  theme_minimal()
```

Now let us put an actual number on it. The Pearson correlation coefficient:

```{r}
cor(iris$Sepal.Length,iris$Petal.Length)
```

That is a strong positive correlation. But here is the question that separates a good analyst from a great one: is it *statistically significant*, or could we have gotten a number like this purely by chance? The `cor.test()` function answers that. Testing whether the observed correlation coefficient is different from 0:

```{r}
cor.test(iris$Sepal.Length,iris$Petal.Length)
```

We can also do a one-sided test --- testing whether the correlation is *greater* than 0. This is useful when you already have a directional hypothesis going in, like "I expect ad spend and revenue to be positively correlated" (if they were negatively correlated, your marketing department has some explaining to do):

```{r}
cor.test(iris$Sepal.Length,iris$Petal.Length,alternative = "greater")
```



### Paired Samples - comparing two-means

Sometimes you want to compare two measurements taken on the *same* subjects. This is the classic "before and after" scenario, and it shows up everywhere in business. Did the training program actually improve test scores? Did the website redesign change time-on-page? Did the new pricing strategy affect average basket size? The key word here is *same* subjects, measured twice.

In our data, we have measurements of writing-hand span and non-writing-hand span for the same students. Think of this as comparing an employee's performance review score before and after a development program --- same people, two time points:

```{r}
summary(survey[,c(2,3)])
```

A paired t-test checks whether the average difference between the two measurements is statistically significant. It is smarter than just comparing two separate groups because it accounts for person-to-person variability:

```{r}
t.test(survey$Wr.Hnd,survey$NW.Hnd,paired = TRUE)
```


### Relationships between more than two numeric variables

When you have a bunch of numeric variables and you want to see all their pairwise relationships at once, a scatterplot matrix is your best friend. It is like speed-dating for variables --- every possible pair gets a quick look, and you can spot which ones have chemistry at a glance:

```{r}
pairs(iris[,1:4])

```

For a fancier version with added density curves and correlation information, the `car` package has you covered. Think of it as the premium upgrade --- same information, but presented more elegantly:

```{r}
library(car)
scatterplotMatrix(iris[,1:4])
```

Now let us look at the actual numbers. The correlation matrix is the "at-a-glance" table that lets you quickly spot which pairs of variables are strongly related. In a business setting, you might run this on metrics like revenue, profit margin, customer count, marketing spend, and employee count to see which ones move together:

```{r}
cor(iris[,1:4])
```

A correlation matrix is great, but how do you know which of those numbers are statistically significant versus just noise? The `rcorr()` function from the **Hmisc** package gives you the correlation coefficients *and* their p-values side by side. This is the "show me the receipts" version --- essential when you need to back up your findings in a report:

```{r}
# Correlations with significance levels
library(Hmisc)
rcorr(as.matrix(iris[,1:4]), type="pearson") # type can be pearson or spearman

```

## Relationships between factor variables

When both your variables are categorical --- think "customer region" and "product preference," or "department" and "preferred communication channel" --- you cannot compute a correlation. (You cannot take the average of "Northeast" and "Midwest." Well, you can try, but R will not be happy about it.) Instead, you use cross-tabulations and the chi-square test of independence to figure out whether the two categories are related or just doing their own thing.

Let us use the `survey` dataset and focus on the `Sex` and `Exer` (exercise frequency) variables. In a business context, this is exactly the kind of analysis you would do to determine whether customer demographics (gender, age group, income bracket) are related to behavioral variables (purchase channel, product preference, loyalty program participation).

```{r}
str(survey)
```

```{r}
summary(survey$Sex)
```

```{r}
summary(survey$Exer)
```

A cross-tab (or cross-tabulation) is the bread and butter of categorical analysis. It shows the joint frequencies of two variables in a neat table --- basically a contingency table. If you have ever built a pivot table in Excel (and if you are a business student, you have built approximately one thousand of them), this is the R equivalent:

```{r}
crosstab <- table(survey$Exer, survey$Sex)
crosstab
```

### Tidyverse cross-tabulation

The tidyverse way of building a cross-tab uses `count()` and `pivot_wider()`. It reads more like a recipe than a math equation, which is always a win for readability:

```{r}
survey %>%
  filter(!is.na(Sex), !is.na(Exer)) %>%
  count(Sex, Exer) %>%
  pivot_wider(names_from = Sex, values_from = n, values_fill = 0)
```

We could also plot that data in the form of stacked or dodged bar graphs. These are the kinds of charts that show up in every quarterly business review presentation, so get comfortable making them.

**Base R:**

```{r}
barplot(crosstab, beside = TRUE, legend = levels(unique(survey$Exer)),
        args.legend = list(x = 8, y = 70))
```

```{r}
barplot(crosstab, legend = levels(unique(survey$Exer)),
        args.legend = list(x = 1.75, y = 150, horiz = TRUE))
```

**ggplot2:**

Dodged bar chart --- bars side by side for easy comparison. This is the "which group is bigger?" view:

```{r}
survey %>%
  filter(!is.na(Sex), !is.na(Exer)) %>%
  ggplot(aes(x = Sex, fill = Exer)) +
  geom_bar(position = "dodge") +
  labs(title = "Exercise Frequency by Sex", y = "Count", fill = "Exercise") +
  theme_minimal()
```

Proportional (filled) bar chart --- great for when you care about the *percentage* breakdown rather than raw counts. This answers "what share of each group falls into each category?":

```{r}
survey %>%
  filter(!is.na(Sex), !is.na(Exer)) %>%
  ggplot(aes(x = Sex, fill = Exer)) +
  geom_bar(position = "fill") +
  labs(title = "Proportion of Exercise Frequency by Sex", y = "Proportion", fill = "Exercise") +
  scale_y_continuous(labels = scales::percent) +
  theme_minimal()
```

OK, so the charts look like there might be a pattern. But "looks like" is not good enough for a business decision. We need a formal test. The chi-square test of independence tells us whether the two variables are actually related or whether any apparent pattern in the chart is just random noise doing random noise things.

An HR analyst might use this exact test to determine whether participation in the company wellness program differs by department. A marketer might test whether response to a campaign varies by customer segment.

The null hypothesis is that the two variables are independent (no relationship). The alternative is that they are related:

```{r}
chisq.test(crosstab)
```

Interpretation: the p-value of .057 is above the typical .05 threshold, which means we do not have strong enough evidence to claim a relationship between sex and exercise frequency. The variables appear to be independent of each other. In business terms, if "sex" were "customer segment" and "exercise" were "product preference," you would conclude that product preference does not significantly differ across segments --- at least not based on this data. And that is a perfectly valid finding. Sometimes the most useful answer is "nope, those two things are not connected."



## Between factor and continuous variables

This is arguably the most common analytical scenario in the business world: you have groups (departments, product lines, regions, customer segments, pricing tiers) and you want to know if a numeric outcome (revenue, satisfaction score, response time, conversion rate) differs across those groups. Are some groups genuinely outperforming others, or is the variation just statistical noise?

### Two levels of a factor level and one continuous variable - t-test

When you have exactly two groups and one numeric variable, the independent samples t-test is your tool. This is the "are these two groups *really* different, or does it just look that way?" test. It shows up everywhere:

- An HR analyst asks: "Do employees with MBAs earn more than those without?"
- A product manager asks: "Do users on the annual plan have higher engagement than month-to-month users?"
- A marketing analyst asks: "Did Campaign A produce higher average order values than Campaign B?"

Let us check the mean height of different genders in the `survey` data.

**Base R:**

```{r}
aggregate(survey$Height, list(survey$Sex), mean, na.rm = T)
```

**Tidyverse:**

```{r}
survey %>%
  filter(!is.na(Sex), !is.na(Height)) %>%
  group_by(Sex) %>%
  summarize(mean_height = mean(Height))
```

The numbers look different, but --- and this is critical --- "looks different" and "is statistically different" are not the same thing. (This distinction has saved many analysts from embarrassing PowerPoint corrections.) Let us visualize the distributions to get a better feel.

**Base R:**

```{r}
plot(survey$Height ~ survey$Sex)
```

**ggplot2:**

```{r}
survey %>%
  filter(!is.na(Sex), !is.na(Height)) %>%
  ggplot(aes(x = Sex, y = Height, fill = Sex)) +
  geom_boxplot() +
  geom_jitter(width = 0.2, alpha = 0.3) +
  labs(title = "Height Distribution by Sex", y = "Height (cm)") +
  theme_minimal() +
  theme(legend.position = "none")
```

Males appear taller than females in the charts, but "appear" is a weasel word. Is the difference statistically significant, or could it be due to the luck of the draw in our particular sample? Time to bring in the t-test. We are testing whether the mean height of females is less than the mean height of males, with a 99% confidence level (because we want to be really sure):

```{r}
t.test(survey$Height ~ survey$Sex, alternative = 'less', conf.level = .99, var.equal = FALSE)
```

Note that the mean height of females is lower than the mean height of males. Hence the test-statistic is negative. The result suggests that mean height of females is indeed less than the mean height of males. (The p-value is tiny --- way below .01 --- and the t-value falls within the 99% confidence interval.) In business terms, this level of confidence means you can walk into the meeting and say "yes, these two groups are genuinely different" without anyone being able to poke holes in your analysis. That is a good feeling.

### More than two levels of a factor variable and one continuous variable - One-Way ANOVA

But what happens when you have *more* than two groups? Maybe you have three customer segments, four regional offices, or five product lines, and you want to know if the average outcome differs across them. You might be tempted to just run a bunch of t-tests --- segment A vs. B, A vs. C, B vs. C --- but statisticians will send you strongly worded emails about something called "inflated Type I error rates." (Translation: you will find "significant" differences that are actually just noise.)

Instead, you use ANOVA --- Analysis of Variance. Despite the name (which sounds like it is about variance), it is really about comparing group means. ANOVA answers one clear question: "Is at least one group's average different from the others?" It does not tell you *which* group is different --- just that a difference exists somewhere. Think of it as a smoke detector: it tells you there is a fire, but not which room it is in.

Let us check the mean height of people who engage in different levels of exercising.

**Base R:**

```{r}
aggregate(survey$Height, list(survey$Exer), mean, na.rm = T)
```

**Tidyverse:**

```{r}
survey %>%
  filter(!is.na(Exer), !is.na(Height)) %>%
  group_by(Exer) %>%
  summarize(
    n = n(),
    mean_height = mean(Height),
    sd_height = sd(Height)
  )
```

**Base R box plot:**

```{r}
plot(survey$Height ~ survey$Exer)
```

**ggplot2:**

```{r}
survey %>%
  filter(!is.na(Exer), !is.na(Height)) %>%
  ggplot(aes(x = Exer, y = Height, fill = Exer)) +
  geom_boxplot() +
  geom_jitter(width = 0.2, alpha = 0.3) +
  labs(title = "Height by Exercise Frequency",
       x = "Exercise Level", y = "Height (cm)") +
  theme_minimal() +
  theme(legend.position = "none")
```

Interestingly, we find that those who engage in frequent exercise have a higher mean height than those who engage in some exercise. In a similar vein, those who engage in some exercise seem to have a higher mean height than those who do not engage in any exercise. But, are these differences statistically significant? Since we have more than two groups, it is ANOVA time.

The null hypothesis is that the means of all groups are the same (boring world, no differences). The alternative is that at least one mean is different from the others (interesting world, something is going on):

```{r}
results <- aov(survey$Height ~ survey$Exer)
summary(results)
```

The findings suggest that we have evidence that at least one of the three means is different from the others. (The p-value of .00354 indicates significance at the .01 level --- this is not a fluke.) In a business context, if these three groups were pricing strategies and the outcome was revenue, you would now know that pricing strategy *matters* --- at least one approach is producing meaningfully different results. **But ANOVA does not tell you which specific groups differ.** It just sounds the alarm.

So naturally, the next question is: OK, *which* groups are different from each other? Is "Frequent" different from "None"? Is "Some" different from "Frequent"? Are all three different from each other? For that detective work, we use Tukey's HSD (Honestly Significant Difference) test. It compares every possible pair of group means and tells you exactly where the statistically significant differences live:

```{r}
TukeyHSD(results, conf.level = .99)
```
