#	Relationships between variables {#multivar}

##	Between continuous variables: Correlations

### Plotting

```{r echo=FALSE}
# NumVar1=rnorm(n=50,mean=0,sd=10)
# 
# NumVar2= -.9*NumVar1+rnorm(50,0,10)
# 
# NumVar3=-.25*NumVar1+rnorm(50,0,10)
# 
# NumVar4=rnorm(50,0,1)
# 
# NumVar5=.25*NumVar1+rnorm(50,0,10)
# 
# NumVar6=.9*NumVar1+rnorm(50,0,10)

# corrdata=data.frame(NumVar1,NumVar2,NumVar3,NumVar4,NumVar5,NumVar6)
# save(corrdata,file="corrdata.Rda")

load("data/corrdata.Rda")
# 
#  cor(corrdata$NumVar1,corrdata$NumVar2)
#  
#  cor(corrdata$NumVar1,corrdata$NumVar3)
#  
#  cor(corrdata$NumVar1,corrdata$NumVar4)
#  
#  cor(corrdata$NumVar1,corrdata$NumVar5)
#  
#  cor(corrdata$NumVar1,corrdata$NumVar6)


par(mfrow=c(2,3))
plot(corrdata$NumVar1,corrdata$NumVar2,pch=18,col="red",xlab="Numeric Variable 1",ylab="Numeric Variable 2",main = "Correlation of -0.65")
abline( lm(corrdata$NumVar2~corrdata$NumVar1),col="blue",) # regression line (y~x) 

plot(corrdata$NumVar1,corrdata$NumVar3,pch=18,col="red",xlab="Numeric Variable 1",ylab="Numeric Variable 2",main = "Correlation of -0.16")
abline( lm(corrdata$NumVar3~corrdata$NumVar1),col="blue",) # regression line (y~x) 

plot(corrdata$NumVar1,corrdata$NumVar4,pch=18,col="red",xlab="Numeric Variable 1",ylab="Numeric Variable 2",main = "Correlation of 0.003")
abline( lm(corrdata$NumVar4~corrdata$NumVar1),col="blue",) # regression line (y~x) 


plot(corrdata$NumVar1,corrdata$NumVar5,pch=18,col="red",xlab="Numeric Variable 1",ylab="Numeric Variable 2",main = "Correlation of 0.20")
abline( lm(corrdata$NumVar5~corrdata$NumVar1),col="blue",) # regression line (y~x) 


plot(corrdata$NumVar1,corrdata$NumVar6,pch=18,col="red",xlab="Numeric Variable 1",ylab="Numeric Variable 2",main = "Correlation of 0.69")
abline( lm(corrdata$NumVar6~corrdata$NumVar1),col="blue",) # regression line (y~x) 


```

### Correlations between two numeric variables

Assumptions of interval/ratio scaled variables, presence of linear relationships, absence of extreme outliers, and normality of the variables are made when conducting these tests. 

```{r}
plot(iris$Sepal.Length,iris$Petal.Length,col="red",xlab="Sepal.Length",ylab="Petal.Length")
abline(lm(iris$Petal.Length~iris$Sepal.Length),col="blue")
```

Pearson Correlation coefficient

```{r}
cor(iris$Sepal.Length,iris$Petal.Length)
```

Testing whether the observed correlation coefficient is different from 0

```{r}
cor.test(iris$Sepal.Length,iris$Petal.Length)
```

Testing whether the observed correlation coefficient is greater than 0

```{r}
cor.test(iris$Sepal.Length,iris$Petal.Length,alternative = "greater")
```



### Paired Samples - comparing two-means

```{r}
summary(survey[,c(2,3)])
```

```{r}
t.test(survey$Wr.Hnd,survey$NW.Hnd,paired = TRUE)
```


### Relationships between more than two numeric variables

```{r}
pairs(iris[,1:4])

```


```{r}
library(car)
scatterplotMatrix(iris[,1:4])
```

The correlations matrix and test of significance

```{r}
cor(iris[,1:4])
```


```{r}
# Correlations with significance levels
library(Hmisc)
rcorr(as.matrix(iris[,1:4]), type="pearson") # type can be pearson or spearman

```

## Relationships between factor variables

Let us revert back to our use of the `survey` dataset and focus on the `Sex` and `Exer`variables. Below, we look at the structure of the dataset and summarize each of the three variables.

```{r}
str(survey)
```

```{r}
summary(survey$Sex)
```

```{r}
summary(survey$Exer)
```

A cross-tab (or a cross-tabulation) presents joint-frequencies of two or more variables. For example, if we were interested in finding how many of the 118 Females engaged in different levels of Exercise, we could use a cross-tab to answer that.

```{r}
crosstab=table(survey$Exer,survey$Sex)
crosstab
```

We could also plot that data in the form of stacked or dodged bar graphs.

```{r}
barplot(crosstab,beside=TRUE, legend=levels(unique(survey$Exer)),args.legend=list(
      x= 8,y=70))
```


```{r}
barplot(crosstab,legend=levels(unique(survey$Exer)),args.legend=list(
      x= 1.75,y=150,horiz=TRUE))
```


If we now wanted to check if the tendency to exercise was independent of the gender, we could use the Chi-square test of independence.

The null hypothesis is that the two variables are independent and the alternate is that they are related.

```{r}
chisq.test(crosstab)
```

Interpretation: the p-value of .057 suggests that there is no evidence of the presence of a relationship between Sex and Exercise. Thus, it is likely that the variables are independent of each other.



##	Between factor and continuous variables

In such cases, one may be interested in studying how the mean scores of the continuous variable differs across multiple levels of the factor variable

### Two levels of a factor level and one continuous variable - t-test

Let us check how the mean height of different genders in the `survey` data.

```{r}
aggregate(survey$Height,list(survey$Sex),mean, na.rm=T)
```

The distribution of Height across the two levels can be studied using a box plot.

```{r}
plot(survey$Height~survey$Sex)
```


Although males seem to be taller than females, it is unclear if the difference being observed is statistically significant. In such a case, we could use an independent samples t-test to answer the question.

```{r}
t.test(survey$Height~survey$Sex, alternative='less', conf.level=.99, var.equal=FALSE)
```

Note that the mean height of females is lower than the mean height of males. Hence the test-statistic is negative. The result suggests that mean height of females is less than the mean height of men. (the p-value is less than .01 and the t-value observed lies in the 99 percent confidence interval.)

### More than two levels of a factor variable and one continuous variable.- One-Way ANOVA

Let us check the mean height of people who engage in different levels of Exercising.


```{r}
aggregate(survey$Height,list(survey$Exer),mean, na.rm=T)
```

```{r}
plot(survey$Height~survey$Exer)
```

Interestingly, we find that those who engage in frequent exercise have a higher mean height than those who engage in some exercise. In a similar vein, those who engage in some exercise seem to have a higher mean height than those who do not engage in any exercise. But, are these differences statistically significant?

Since we have more than two levels of the factor variable, one will have to use an ANOVA (Analysis of Variance). 

The Null hypothesis is that the means of all groups is the same and the alternate is that at least one mean is different.

```{r}
results=aov(survey$Height ~ survey$Exer)
summary(results)
```

The findings suggest that we have evidence that at least one of the three means is different from the others. (The p-value of .00354 indicates significance at the .01 level.) **Note that the earlier test does not tell us whether the differences between consecutive means is statistically significant.**

We can do a multiple-comparison of means to test that.

```{r}
TukeyHSD(results,conf.level = .99)
```
