# The Tidyverse Ecosystem {#tidyverse}

```{r include=FALSE}
library(knitr)
opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE,
               comment = "#>", collapse = TRUE)
```

If base R is a Swiss Army knife, the tidyverse is the entire Williams-Sonoma catalog --- beautifully designed, everything matches, and once you learn where things go, cooking (er, analyzing data) becomes a joy instead of a chore.

This chapter gives you the 30,000-foot overview before we dive into individual packages later. Think of it as the orientation session before your first day at the office.

## What is the Tidyverse?

The **tidyverse** is a collection of R packages built for data science. They were created primarily by Hadley Wickham and the team at [Posit](https://posit.co/) (formerly RStudio), and they all follow the same design philosophy. That means once you learn one package, the others feel familiar --- like walking into a new franchise location and already knowing the menu.

The "tidy" in tidyverse boils down to two big ideas:

1. **Tidy data.** Your data should be organized so that each variable is a column, each observation is a row, and each cell has one value. (More on this soon --- it matters more than you think.)

2. **Tidy code.** Your code should read like a sentence, not an encrypted ransom note. Tidyverse functions have verb-based names --- `filter()`, `select()`, `mutate()`, `summarize()` --- so a pipeline of code practically narrates what you are doing.

The tidyverse is not a replacement for base R. It is an opinionated, well-organized toolkit that most people find easier to learn and read for everyday data analysis. Base R is still essential for programming, package development, and plenty of specialized tasks.

## Installing and Loading the Tidyverse

One install, many packages:

```{r eval=FALSE}
install.packages("tidyverse")
```

One library call to load the core set:

```{r}
library(tidyverse)
```

When you run `library(tidyverse)`, R prints two things. First, the **core packages** it just loaded. Second, any **conflicts** --- cases where a tidyverse function has the same name as one already loaded. You will usually see:

- `dplyr::filter()` masks `stats::filter()`
- `dplyr::lag()` masks `stats::lag()`

Do not panic. This just means that typing `filter()` now calls the dplyr version (which is what you want 99% of the time). If you ever need the base R version, use `stats::filter()`.

You can also load packages individually if you prefer a lighter footprint:

```{r eval=FALSE}
library(dplyr)
library(ggplot2)
library(tidyr)
```

## Core Tidyverse Packages

Each core package handles a specific job in your data analysis workflow. Here is the lineup --- think of them as departments in a well-run company.

### ggplot2 --- Data Visualization

**ggplot2** lets you build plots layer by layer instead of picking from a fixed menu of chart types. It is the tidyverse's crown jewel, and you will spend a lot of quality time with it. Chapters \@ref(ggplot2basics) and \@ref(ggplot2customize) go deep.

```{r fig.cap="A ggplot2 scatterplot of the iris dataset."}
ggplot(iris, aes(x = Sepal.Length, y = Sepal.Width, color = Species)) +
  geom_point(size = 2) +
  labs(title = "Sepal Dimensions by Species",
       x = "Sepal Length (cm)", y = "Sepal Width (cm)") +
  theme_minimal()
```

### dplyr --- Data Manipulation

**dplyr** is your data wrangling workhorse. It gives you a handful of verbs that handle most of what you need: `filter()` rows, `select()` columns, `mutate()` to create new variables, `arrange()` to sort, `summarize()` to aggregate, and `group_by()` to do all of the above within groups. Chapter \@ref(dplyr) covers it in detail.

```{r}
mtcars %>%
  group_by(cyl) %>%
  summarize(
    n = n(),
    avg_mpg = mean(mpg),
    avg_hp = mean(hp)
  )
```

### tidyr --- Data Tidying

**tidyr** reshapes your data between "wide" and "long" formats with `pivot_longer()` and `pivot_wider()`. If you have ever wrestled with a spreadsheet where the column headers are actually data values, tidyr is your new best friend. Chapter \@ref(tidyr) has the details.

### readr --- Data Import

**readr** reads rectangular data from CSV, TSV, and other delimited files. Its `read_csv()` is faster than base R's `read.csv()`, produces tibbles, and has better defaults (no surprise factor conversions). Chapter \@ref(dataimport) covers data import.

### purrr --- Functional Programming

**purrr** gives you `map()` functions that apply an operation to every element of a list or vector. It is like VLOOKUP's cooler, more powerful cousin. Chapter \@ref(purrr) goes deeper.

```{r}
map_dbl(iris[, 1:4], mean)
```

### tibble --- Modern Data Frames

**Tibbles** are data frames that learned manners. They print neatly, never silently convert your strings to factors, and have stricter rules that prevent subtle bugs. We will talk more about them later in this chapter.

### stringr --- String Manipulation

**stringr** handles text data. Every function starts with `str_` and takes a character vector first, so it plays nicely with pipes. Chapter \@ref(stringr) covers it.

```{r}
fruits <- c("apple", "banana", "cherry", "date")
str_detect(fruits, "an")
str_to_upper(fruits)
```

### forcats --- Working with Factors

**forcats** makes categorical variables less painful. Reorder levels, collapse categories, recode labels --- all the things you wish Excel could do cleanly. (The name is an anagram of "factors." Nerdy, but fun.) Chapter \@ref(forcats) has more.

### lubridate --- Dates and Times

**lubridate** makes dates and times behave. Parse them, extract components, do arithmetic --- without the headaches that dates usually cause. Chapter \@ref(lubridate) covers it.

```{r}
today_date <- ymd("2024-09-15")
today_date
year(today_date)
month(today_date, label = TRUE)
today_date + days(30)
```

### Beyond the Core

The broader tidyverse ecosystem includes packages like **readxl** (Excel files), **haven** (SPSS/Stata/SAS), **httr2** (web APIs), **rvest** (web scraping), and **glue** (string interpolation). These are not loaded by `library(tidyverse)` but can be installed individually when you need them.

## The Pipe Operator {#pipe}

If there is one thing you take away from this chapter, let it be the pipe. It is the single most important concept in the tidyverse, and once it clicks, you will never want to go back.

### What Does the Pipe Do?

The pipe takes the thing on its left and feeds it as the **first argument** to the function on its right. That is it. That is the whole idea.

The magrittr pipe `%>%` (loaded with the tidyverse):

> `x %>% f(y)` is the same as `f(x, y)`

The base R pipe `|>` (available since R 4.1):

> `x |> f(y)` is the same as `f(x, y)`

For everyday work, they are interchangeable. This book uses `%>%` because it is the established tidyverse convention, but `|>` works fine too.

### Why the Pipe Changes Everything

Think of the pipe as an assembly line. Raw materials go in on one end, each station does one job, and the finished product comes out the other end. Without the pipe, your code looks like either a Russian nesting doll or a cluttered warehouse of temporary variables. Neither is fun.

Let us say you want to take the `mtcars` dataset, keep only 6-cylinder cars, pick three columns, and sort by fuel economy.

**Without pipes --- nested calls (read inside-out, good luck):**

```{r}
arrange(select(filter(mtcars, cyl == 6), mpg, hp, wt), mpg)
```

Reading this is like trying to understand a sentence that was written backwards. The first thing that happens (`filter`) is buried in the middle.

**Without pipes --- intermediate variables (clutters your workspace):**

```{r}
step1 <- filter(mtcars, cyl == 6)
step2 <- select(step1, mpg, hp, wt)
step3 <- arrange(step2, mpg)
step3
```

Readable, but now you have three temporary variables floating around like Post-it notes you forgot to throw away.

**With the pipe (read top-to-bottom, like a recipe):**

```{r}
mtcars %>%
  filter(cyl == 6) %>%
  select(mpg, hp, wt) %>%
  arrange(mpg)
```

Now it reads like plain English: "Take `mtcars`, **then** filter for 6 cylinders, **then** select three columns, **then** arrange by mpg." Clean. Clear. No junk variables. This is why people love the pipe.

### More Pipe Examples

**Summarizing data by group:**

```{r}
iris %>%
  group_by(Species) %>%
  summarize(
    mean_petal_length = mean(Petal.Length),
    sd_petal_length = sd(Petal.Length),
    count = n()
  )
```

**Piping straight into a chart:**

```{r fig.cap="Average horsepower by number of cylinders."}
mtcars %>%
  group_by(cyl) %>%
  summarize(avg_hp = mean(hp)) %>%
  ggplot(aes(x = factor(cyl), y = avg_hp)) +
  geom_col(fill = "steelblue") +
  labs(x = "Cylinders", y = "Average Horsepower",
       title = "Average Horsepower by Cylinder Count") +
  theme_minimal()
```

One thing to watch: when piping into ggplot2, you switch from `%>%` to `+` after the `ggplot()` call. The `+` is how ggplot2 layers work --- it is not a pipe.

### The Dot Placeholder

Sometimes you need the piped value somewhere other than the first argument. The magrittr pipe lets you use `.` as a placeholder:

```{r}
# Pipe into a non-first argument position using the dot
c(3, 7, 1, 9, 2) %>% paste("The number is", .)
```

```{r}
# Using the dot in a model formula context
mtcars %>%
  lm(mpg ~ wt + hp, data = .) %>%
  summary() %>%
  .$r.squared
```

Here, `mtcars` goes into the `data` argument of `lm()` (not the first argument), because we told it exactly where with the dot.

## Tidy Data Principles {#tidydata}

"Tidy data" is a specific way of organizing data that makes everything downstream --- analysis, visualization, modeling --- dramatically easier. The three rules:

1. **Each variable forms a column.**
2. **Each observation forms a row.**
3. **Each value occupies a single cell.**

Sounds obvious, right? And yet, a shocking amount of real-world business data violates these rules. Let us look at a classic example.

### The Messy Sales Spreadsheet

Imagine your regional sales manager emails you this quarter's numbers. It looks like a perfectly normal spreadsheet:

```{r}
untidy_sales <- tibble(
  product = c("Widget A", "Widget B", "Widget C"),
  Q1 = c(150, 200, 180),
  Q2 = c(170, 210, 160),
  Q3 = c(190, 195, 200),
  Q4 = c(210, 230, 220)
)
untidy_sales
```

Looks fine, right? Wrong. This is **untidy** because Q1, Q2, Q3, and Q4 are not variable names --- they are **values** of a variable called "quarter." The actual data has been smeared across multiple columns.

Here is the tidy version, created with `pivot_longer()`:

```{r}
tidy_sales <- untidy_sales %>%
  pivot_longer(cols = Q1:Q4, names_to = "quarter", values_to = "sales")
tidy_sales
```

Now every row is one product in one quarter. Every column is a variable. Every cell holds one value. *That* is tidy.

### Why Should You Care?

Because tidy data plays beautifully with the tidyverse. Want a grouped bar chart comparing products by quarter? With tidy data, it is one pipeline:

```{r fig.cap="Quarterly sales by product, created easily from tidy data."}
tidy_sales %>%
  ggplot(aes(x = quarter, y = sales, fill = product)) +
  geom_col(position = "dodge") +
  labs(title = "Quarterly Sales by Product",
       x = "Quarter", y = "Sales (units)") +
  theme_minimal()
```

With the messy version, you would need a lot more manual reshaping. Tidy data saves you time and headaches. Get your data tidy first, and everything else falls into place.

## Tibbles vs Data Frames {#tibbles}

Short version: **tibbles are data frames that learned manners.** They are technically still data frames under the hood, but they behave more predictably. Here is what you need to know.

**They print nicely.** Tibbles show just the first 10 rows and fit columns to your screen, instead of dumping 10,000 rows into your console like a base R data frame would.

```{r}
# A tibble prints neatly
as_tibble(mtcars)
```

**No sneaky partial matching.** Base R lets you type `df$cat` and it will guess you meant `df$category`. Tibbles require the exact name. Fewer bugs, fewer "wait, what just happened?" moments.

**No surprise factor conversions.** Tibbles keep your text as text. No silent conversions.

**Consistent subsetting.** Pulling one column from a tibble with `[` always gives you a tibble back. Base R data frames sometimes give you a vector, sometimes a data frame. Tibbles are predictable.

In practice, you rarely need to think about this distinction. Tidyverse functions return tibbles automatically. If older code needs a base data frame, just use `as.data.frame()`.

## The Tidyverse Workflow {#workflow}

The tidyverse packages are not random --- they map to stages of a data analysis pipeline:

1. **Import** your data (readr, readxl, haven)
2. **Tidy** it into a clean structure (tidyr)
3. **Transform** --- filter, create variables, summarize (dplyr)
4. **Visualize** patterns (ggplot2)
5. **Model** relationships (tidymodels ecosystem)
6. **Communicate** findings (R Markdown, Quarto)

Let us walk through a quick example with the `mpg` dataset (fuel economy for 38 car models, 1999--2008). Think of this as a mini consulting engagement.

### Import and Explore

Since `mpg` is built in, we skip the file import. In real life, you would use `read_csv()`. Let us see what we are working with:

```{r}
glimpse(mpg)
```

### Transform

Which manufacturers have the best highway fuel economy? Let us find the top 10:

```{r}
top_manufacturers <- mpg %>%
  group_by(manufacturer) %>%
  summarize(
    avg_hwy = mean(hwy),
    n_models = n(),
    .groups = "drop"
  ) %>%
  arrange(desc(avg_hwy)) %>%
  slice_head(n = 10)

top_manufacturers
```

### Visualize

Now let us turn those numbers into a chart your boss would actually want to see:

```{r fig.cap="Top 10 manufacturers by average highway fuel economy."}
top_manufacturers %>%
  mutate(manufacturer = fct_reorder(manufacturer, avg_hwy)) %>%
  ggplot(aes(x = avg_hwy, y = manufacturer)) +
  geom_col(fill = "steelblue") +
  geom_text(aes(label = round(avg_hwy, 1)), hjust = -0.1, size = 3) +
  labs(
    title = "Top 10 Manufacturers by Highway Fuel Economy",
    x = "Average Highway MPG",
    y = NULL
  ) +
  theme_minimal() +
  xlim(0, 40)
```

And here is a complete pipeline --- raw data to finished chart in one unbroken chain:

```{r fig.cap="Distribution of highway fuel economy by vehicle class, ordered by median."}
mpg %>%
  mutate(class = str_to_title(class)) %>%
  mutate(class = fct_reorder(class, hwy, .fun = median)) %>%
  ggplot(aes(x = class, y = hwy, fill = class)) +
  geom_boxplot(show.legend = FALSE, alpha = 0.7) +
  coord_flip() +
  labs(
    title = "Highway Fuel Economy by Vehicle Class",
    subtitle = "Ordered by median MPG",
    x = NULL,
    y = "Highway MPG"
  ) +
  theme_minimal() +
  theme(plot.title = element_text(face = "bold"))
```

This pipeline uses stringr, forcats, dplyr, and ggplot2 all in one chain. That is the tidyverse at its best: small, focused functions that snap together like LEGO bricks.

## Getting Help {#tidyverse-help}

You *will* get stuck. Everyone does. Here is how to get unstuck fast.

### Cheat Sheets (Your New Best Friends)

Posit publishes free, two-page PDF cheat sheets for most tidyverse packages. Print them out. Tape them to your wall. Seriously. Find them at [https://posit.co/resources/cheatsheets/](https://posit.co/resources/cheatsheets/) or in RStudio under **Help > Cheat Sheets**. The dplyr, ggplot2, and tidyr cheat sheets alone will save you hours.

### Googling Errors Like a Pro

When you get an error, copy the error message into Google (minus your specific variable names) and add "R" or "tidyverse." Nine times out of ten, someone on Stack Overflow has already asked and answered your exact question. Tag your own questions with `[r]` and the package name (e.g., `[dplyr]`, `[ggplot2]`).

### Built-In Help

```{r eval=FALSE}
# Help on a specific function
?filter

# Browse all functions in a package
help(package = "dplyr")

# Long-form tutorials (vignettes)
browseVignettes("dplyr")
```

### Books (Free Online)

- **R for Data Science** (2nd ed.) --- [https://r4ds.hadley.nz/](https://r4ds.hadley.nz/). The definitive tidyverse guide. If you read one book, make it this one.
- **ggplot2: Elegant Graphics for Data Analysis** (3rd ed.) --- [https://ggplot2-book.org/](https://ggplot2-book.org/).

### RStudio Tips

- **Tab completion**: Start typing a function name and hit Tab. RStudio will show you matches and arguments.
- **F1 key**: Put your cursor on any function name and press F1 to open its help page instantly.
- **The Help pane**: Search and browse documentation without leaving RStudio.

## Summary

Here is the executive summary (see what we did there?):

- The **tidyverse** is a collection of R packages that share a common design philosophy. Learn one, and the others feel familiar.
- The **pipe operator** (`%>%` or `|>`) chains operations into readable, left-to-right pipelines. It is the single most important concept in this chapter.
- **Tidy data** --- each variable a column, each observation a row, each value a cell --- is the organizing principle that makes everything work smoothly.
- **Tibbles** are friendlier data frames with safer defaults.
- The **tidyverse workflow** (Import, Tidy, Transform, Visualize, Model, Communicate) gives you a structured approach to any data analysis project.
- **When you get stuck**, use cheat sheets, Google your error messages, and check Stack Overflow. You are not the first person to hit that error, and you will not be the last.

In the chapters ahead, we will roll up our sleeves and explore each of these packages in depth. Onward.
