# Data Import {#dataimport}

```{r setup-dataimport, include=FALSE}
library(knitr)
opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE,
               comment = "#>", collapse = TRUE)
```

Before you can analyze anything, you have to get data into R. That sounds obvious, but it trips people up more often than you'd think. Your boss emails you a CSV. A client sends over an Excel workbook with six tabs. Someone from the research team hands you a mysterious `.sav` file from SPSS. This chapter shows you how to handle all of it.

We'll use three packages: **readr** (for CSV and other text files), **readxl** (for Excel), and **haven** (for SPSS/Stata/SAS files). The first one loads automatically with the tidyverse; the other two you load separately.

```{r}
library(tidyverse)
library(readxl)
library(haven)
```


## Reading CSV Files {#csv}

CSV files are the lingua franca of data. If someone says "I'll send you the data," there's a 90% chance it's a CSV. The function you want is `read_csv()` from the readr package.

### Why read_csv() Instead of read.csv()?

Base R has `read.csv()`. It works, but `read_csv()` is better in almost every way:

- It's **faster**, especially on large files (like that 500,000-row sales export).
- It returns a **tibble** instead of a plain data frame, so it prints nicely.
- It **doesn't silently convert your text to factors**. (If you've ever had "Q1 Revenue" turn into a mystery number, you know the pain.)
- It **tells you how it parsed each column**, so you can catch problems early.

Let's see it in action. `read_csv()` can even read data from an inline string, which is handy for quick demos.

```{r}
sample_data <- read_csv(
  "name,age,score,enrolled
   Alice,25,92.5,TRUE
   Bob,30,88.0,TRUE
   Charlie,22,95.3,FALSE
   Diana,28,79.1,TRUE
   Eve,35,85.6,FALSE"
)

sample_data
```

See that message? It's telling you it read `name` as text, `age` and `score` as numbers, and `enrolled` as TRUE/FALSE. That's the kind of transparency you want.

### The Arguments You'll Actually Use

**col_names** -- By default, the first row is treated as column headers. If your file doesn't have headers (looking at you, legacy systems), set `col_names = FALSE` or supply your own names.

```{r}
read_csv(
  "1,2,3\n4,5,6\n7,8,9",
  col_names = c("x", "y", "z")
)
```

**na** -- Different systems represent missing data differently. Some use blanks, some use "N/A", some use "." because... reasons. Tell R what to look for.

```{r}
read_csv(
  "name,value\nAlice,10\nBob,N/A\nCharlie,.",
  na = c("", "NA", "N/A", ".")
)
```

**skip** -- If the file has junk at the top (report headers, timestamps, disclaimers), skip those lines.

```{r}
read_csv(
  "This is a comment line\nAnother comment\nname,value\nAlice,10\nBob,20",
  skip = 2
)
```

**n_max** -- Got a 2-million-row file and just want to peek at it? Read only the first few rows.

```{r}
read_csv(
  "name,value\nAlice,10\nBob,20\nCharlie,30\nDiana,40",
  n_max = 2
)
```

### Specifying Column Types

Sometimes R guesses wrong. The classic example: zip codes. If a zip code starts with 0 (like 02134 for Boston), R will read it as the number 2134 and silently drop the leading zero. Not great.

```{r}
# Problem: leading zero gets eaten
problem_data <- read_csv(
  "id,zipcode,amount\n1,02134,100\n2,10001,200\n3,90210,300"
)

problem_data
```

The fix: tell R that `zipcode` is text, not a number.

```{r}
fixed_data <- read_csv(
  "id,zipcode,amount\n1,02134,100\n2,10001,200\n3,90210,300",
  col_types = cols(zipcode = col_character())
)

fixed_data
```

You can also use a compact shorthand where each letter represents a type: `c` = character, `i` = integer, `d` = double, `l` = logical, `_` = skip the column entirely.

```{r}
read_csv(
  "id,name,score,active\n1,Alice,92.5,TRUE\n2,Bob,88.0,FALSE",
  col_types = "icdc"
)
```

If you only need certain columns from a massive file, `cols_only()` lets you cherry-pick.

```{r}
read_csv(
  "id,name,score,active\n1,Alice,92.5,TRUE\n2,Bob,88.0,FALSE",
  col_types = cols_only(
    name = col_character(),
    score = col_double()
  )
)
```


## Other Delimited Files

Not everything is comma-separated. Some files use tabs, pipes, or semicolons. readr has you covered.

```{r}
# Tab-separated
read_tsv(
  "name\tage\tscore\nAlice\t25\t92.5\nBob\t30\t88.0"
)
```

```{r}
# Pipe-delimited (sometimes you'll see this from older database exports)
read_delim(
  "name|age|score\nAlice|25|92.5\nBob|30|88.0",
  delim = "|"
)
```

If you're working with European-format CSVs (semicolons as delimiters, commas as decimal marks), use `read_csv2()`. This comes up more than you'd expect in international business.

```{r}
read_csv2(
  "name;age;score\nAlice;25;92,5\nBob;30;88,0"
)
```


## Reading Excel Files {#excel}

Let's be honest: in business, Excel is king. Your quarterly reports, your budget forecasts, your customer lists -- they're probably all in `.xlsx` files. The **readxl** package handles these beautifully.

### The Basics

`read_excel()` auto-detects whether a file is `.xls` or `.xlsx` and just works.

```{r}
# readxl comes with example files for practice
datasets_path <- readxl_example("datasets.xlsx")
datasets_data <- read_excel(datasets_path)
datasets_data
```

### Working with Sheets

Excel workbooks usually have multiple sheets -- maybe "Q1 Sales," "Q2 Sales," "Summary." You can pick which sheet to read.

```{r}
# See what sheets are available
excel_sheets(datasets_path)
```

```{r}
# Read a specific sheet by name
mtcars_data <- read_excel(datasets_path, sheet = "mtcars")
head(mtcars_data)
```

```{r}
# Or by position (the 3rd sheet)
quakes_data <- read_excel(datasets_path, sheet = 3)
head(quakes_data)
```

### Reading Specific Cell Ranges

Sometimes you don't want the whole sheet. Maybe the data you need is in cells B2:F50, and there's a bunch of formatting garbage around it. The `range` argument uses Excel-style cell references, so it feels familiar.

```{r}
# Read a specific cell range
read_excel(datasets_path, range = "mtcars!B1:D5")
```

```{r}
# Read specific rows
read_excel(datasets_path, sheet = "mtcars", range = cell_rows(1:6))
```

```{r}
# Read specific columns
read_excel(datasets_path, sheet = "mtcars", range = cell_cols("A:C"))
```

Just like `read_csv()`, you can use `na`, `skip`, `col_names`, and `n_max` to handle messy Excel files.

```{r}
read_excel(
  datasets_path,
  sheet = "chickwts",
  na = c("", "NA"),
  skip = 0,
  n_max = 5
)
```


## SPSS, Stata, and SAS Files

If someone hands you a `.sav` file (SPSS), a `.dta` file (Stata), or a `.sas7bdat` file (SAS), don't panic. The **haven** package reads all of them.

```{r eval=FALSE}
# SPSS
spss_data <- read_sav("path/to/datafile.sav")

# Stata
stata_data <- read_dta("path/to/datafile.dta")

# SAS
sas_data <- read_sas("path/to/datafile.sas7bdat")
```

These files often have special "value labels" (like 1 = "Male", 2 = "Female"). Haven preserves those as a special `labelled` class. To convert them to regular R factors, use `as_factor()`.

```{r eval=FALSE}
spss_data <- read_sav("path/to/datafile.sav") %>%
  mutate(across(where(is.labelled), as_factor))
```

You probably won't use these every day, but it's nice to know they exist when you need them.


## R's Own File Formats

R has two native formats for saving data: `.RData` and `.rds`. Here's the quick version:

- **`.RData`** files can store multiple objects. You load them with `load()`, and they pop back into your environment with their original names. The catch? They can overwrite things without warning.
- **`.rds`** files store a single object. You read them with `readRDS()` and assign the result to whatever name you want. Much safer.

```{r eval=FALSE}
# Saving and loading .RData
x <- 1:10
y <- data.frame(a = 1:3, b = letters[1:3])
save(x, y, file = "my_objects.RData")
load("my_objects.RData")  # x and y are restored

# Saving and loading .rds (preferred)
my_data <- tibble(name = c("Alice", "Bob"), score = c(92.5, 88.0))
saveRDS(my_data, file = "my_data.rds")
restored_data <- readRDS("my_data.rds")
```

For day-to-day work, prefer `.rds`. It's explicit, it's safe, and it won't accidentally overwrite your variables.


## Writing Data Out

You can also go the other direction -- save your clean data for someone else (or for future you).

```{r eval=FALSE}
my_data <- tibble(
  name = c("Alice", "Bob", "Charlie"),
  score = c(92.5, 88.0, 95.3),
  date = as.Date(c("2025-01-15", "2025-02-20", "2025-03-10"))
)

# Write a CSV
write_csv(my_data, "my_data.csv")

# If someone will open it in Excel, use this version
# (it handles special characters better)
write_excel_csv(my_data, "my_data_excel.csv")

# Save as an R object (preserves column types perfectly)
write_rds(my_data, "my_data.rds")
```

One thing to watch out for: when you save to CSV, you lose type information. Factors become plain text, dates become strings. If you need to preserve everything exactly, use `write_rds()`.


## Handling Common Import Problems

Real-world data is messy. Here are the problems you'll hit most often and how to fix them.

**Missing values in disguise.** Different systems use different codes for "no data." Check your file and tell R what to look for.

```{r}
read_csv(
  "name,value\nAlice,10\nBob,\nCharlie,NA\nDiana,N/A\nEve,-99",
  na = c("", "NA", "N/A", "-99")
)
```

**Junk at the top of the file.** Reports exported from enterprise systems love to put timestamps and headers above the actual data. Use `skip` or `comment`.

```{r}
messy_file <- read_csv(
  "Report generated on 2025-01-15\nSource: Internal Database\n\nname,score\nAlice,92\nBob,88\nCharlie,95",
  skip = 3
)

messy_file
```

**Reading from a URL.** All readr functions can read directly from web URLs. Very handy for publicly available datasets.

```{r eval=FALSE}
url_data <- read_csv("https://raw.githubusercontent.com/tidyverse/readr/main/inst/extdata/mtcars.csv")
```

**Encoding problems.** If your data looks like garbled nonsense (especially with accented characters or non-English text), it's probably an encoding issue.

```{r eval=FALSE}
# Try specifying the encoding
latin_data <- read_csv("data_latin1.csv", locale = locale(encoding = "Latin1"))

# Or let R guess
guess_encoding("mystery_file.csv")
```


## Reading Multiple Files at Once

In the real world, you'll often have data split across multiple files -- monthly sales reports, regional data dumps, weekly survey exports. You need to read them all and stack them together.

The easiest approach (readr 2.0+) is to hand `read_csv()` a vector of file paths directly.

```{r eval=FALSE}
csv_files <- list.files("data/", pattern = "\\.csv$", full.names = TRUE)
all_data <- read_csv(csv_files, id = "source_file")
```

The `id` argument adds a column telling you which file each row came from. Very useful for debugging.

Here's a self-contained example showing the pattern:

```{r}
file1 <- "name,score\nAlice,92\nBob,88"
file2 <- "name,score\nCharlie,95\nDiana,79"
file3 <- "name,score\nEve,85\nFrank,91"

files <- list(file1, file2, file3)

combined_data <- files %>%
  set_names(paste0("file_", 1:3)) %>%
  map(read_csv) %>%
  list_rbind(names_to = "file_number")

combined_data
```


## Summary

Here's your cheat sheet for getting data into R:

- **CSV files**: `read_csv()` from readr. This is your bread and butter.
- **Excel files**: `read_excel()` from readxl. Supports sheets, cell ranges, all the good stuff.
- **SPSS/Stata/SAS**: `read_sav()`, `read_dta()`, `read_sas()` from haven. There when you need them.
- **R native files**: `readRDS()` for `.rds`, `load()` for `.RData`.
- **Writing data**: `write_csv()` for CSVs, `write_rds()` to preserve R types.
- **Multiple files**: Pass a vector of paths to `read_csv()`, or use `map()` with `list_rbind()`.

The most important habit: always inspect your data right after importing it. Check the column types, look for unexpected `NA` values, make sure nothing got mangled. Five minutes of checking now saves five hours of debugging later.
