# Data Wrangling with dplyr {#dplyr}

```{r include=FALSE}
library(tidyverse)
library(knitr)
opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

Data wrangling --- the process of cleaning, transforming, and reshaping data --- is where analysts spend the majority of their time. Getting data into the right shape for analysis or visualization is often the most labor-intensive step of any project. The **dplyr** package, a core member of the tidyverse, provides an elegant and efficient grammar for tackling these tasks. This chapter covers dplyr in depth, walking through every major function with extensive examples.

## Introduction to dplyr

The dplyr package was created by Hadley Wickham and is now maintained by the tidyverse team. It provides a consistent set of "verbs" --- functions that correspond to the most common data manipulation tasks. Rather than learning dozens of unrelated base R functions with inconsistent argument orders and return types, dplyr gives you a small vocabulary of well-designed functions that combine to accomplish virtually any transformation.

The key advantages of dplyr over base R include:

- **Readability**: Code written with dplyr reads almost like English. A chain of operations like "take the data, filter the rows, select some columns, then summarize" translates directly into dplyr code.
- **Consistency**: Every dplyr function takes a data frame as its first argument and returns a data frame. This consistency makes functions easy to learn and easy to chain together with the pipe operator.
- **Speed**: dplyr is implemented with performance in mind. For large datasets, operations are substantially faster than their base R equivalents.
- **Database support**: The same dplyr code can operate on data frames in memory, database tables, or even Apache Spark clusters with minimal modification.

The dplyr package is loaded as part of the tidyverse, so running `library(tidyverse)` gives you access to all dplyr functions.


## The Core Verbs

The dplyr package is organized around six core verbs. Nearly every data manipulation task can be expressed as some combination of these operations:

| Verb | Purpose |
|:-------------|:----------------------------------------------|
| `filter()` | Select rows based on conditions |
| `select()` | Select columns by name or position |
| `mutate()` | Create new columns or modify existing ones |
| `summarize()` | Collapse many rows into summary statistics |
| `arrange()` | Reorder rows |
| `group_by()` | Define groups for grouped operations |

Each of these verbs follows a common pattern:

1. The first argument is a data frame.
2. Subsequent arguments describe what to do, using unquoted column names.
3. The result is a new data frame.

Because the first argument is always a data frame and the result is always a data frame, these verbs chain together naturally using the pipe operator (`|>` or `%>%`). We will explore each verb in detail.


## filter() - Selecting Rows {#filter}

The `filter()` function selects rows that satisfy one or more conditions. It is the dplyr equivalent of subsetting rows in base R with `[`, but with cleaner syntax.

### Comparison Operators

The most basic filters use comparison operators to test column values.

```{r}
# Rows where Sepal.Length is greater than 7
filter(iris, Sepal.Length > 7)
```

```{r}
# Rows where Species is exactly "setosa"
filter(iris, Species == "setosa")
```

```{r}
# Cars with exactly 6 cylinders
filter(mtcars, cyl == 6)
```

```{r}
# Cars with mpg not equal to 21
filter(mtcars, mpg != 21)
```

The full set of comparison operators available in R is: `==` (equal), `!=` (not equal), `>` (greater than), `<` (less than), `>=` (greater than or equal), and `<=` (less than or equal).

### Logical Operators

Multiple conditions can be combined using logical operators. The `&` operator means "and" (both conditions must be true), `|` means "or" (at least one condition must be true), and `!` negates a condition.

```{r}
# Setosa flowers with Sepal.Length greater than 5
filter(iris, Species == "setosa" & Sepal.Length > 5)
```

When you supply multiple arguments to `filter()`, they are combined with `&` automatically. The following is equivalent to the code above:

```{r}
filter(iris, Species == "setosa", Sepal.Length > 5)
```

```{r}
# Cars with 4 cylinders OR more than 30 mpg
filter(mtcars, cyl == 4 | mpg > 30)
```

```{r}
# Flowers that are NOT setosa
filter(iris, !(Species == "setosa"))
```

### The %in% Operator

When you want to filter for rows where a column matches any value in a set, use `%in%`.

```{r}
# Setosa or virginica flowers
filter(iris, Species %in% c("setosa", "virginica"))
```

```{r}
# Cars with 4 or 8 cylinders
filter(mtcars, cyl %in% c(4, 8))
```

### between() and near()

The `between()` helper tests whether values fall within a range (inclusive on both ends). The `near()` function is useful for comparing floating-point numbers, which can have rounding issues with `==`.

```{r}
# Sepal.Length between 5.0 and 6.0 (inclusive)
filter(iris, between(Sepal.Length, 5.0, 6.0))
```

```{r}
# near() handles floating point comparison safely
# This may miss values due to floating point precision:
# filter(iris, Sepal.Length == 5.1)
# near() is safer:
filter(iris, near(Sepal.Length, 5.1))
```

### Filtering Missing Values

Missing values (`NA`) require special handling. You cannot test for `NA` with `==`; instead, use `is.na()`.

```{r}
# Create a small data frame with missing values
df_with_na <- tibble(
  x = c(1, 2, NA, 4, 5),
  y = c("a", NA, "c", "d", NA)
)

# Filter to rows where x is not missing
filter(df_with_na, !is.na(x))
```

```{r}
# Filter to rows where neither x nor y is missing
filter(df_with_na, !is.na(x), !is.na(y))
```

### Complex Filter Examples

Real-world filtering often involves combining several conditions.

```{r}
# High-performance cars: more than 200 horsepower and under 15 seconds quarter mile
filter(mtcars, hp > 200, qsec < 15)
```

```{r}
# Efficient cars: 4 cylinders with above-average mpg
filter(mtcars, cyl == 4, mpg > mean(mpg))
```

```{r}
# Large setosa or small virginica flowers
filter(iris,
       (Species == "setosa" & Sepal.Length > 5.5) |
       (Species == "virginica" & Sepal.Length < 6.0))
```


## select() - Selecting Columns {#select}

The `select()` function picks columns from a data frame. It is especially useful when working with wide datasets that have many columns, only a few of which are relevant for the current task.

### Selecting by Name

```{r}
# Select specific columns by name
select(iris, Sepal.Length, Species)
```

```{r}
# Select a range of columns
select(iris, Sepal.Length:Petal.Length)
```

```{r}
# Exclude columns using the minus sign
select(iris, -Species)
```

```{r}
# Exclude a range of columns
select(iris, -(Sepal.Length:Petal.Length))
```

### Selecting by Position

```{r}
# Select the first and last columns
select(iris, 1, 5)
```

### Helper Functions

The `select()` function provides several helper functions that match columns based on patterns in their names. These are particularly useful when working with data that has many similarly named columns.

```{r}
# Columns that start with "Sepal"
select(iris, starts_with("Sepal"))
```

```{r}
# Columns that end with "Width"
select(iris, ends_with("Width"))
```

```{r}
# Columns that contain "pal"
select(iris, contains("pal"))
```

```{r}
# Columns that match a regular expression
select(iris, matches("^S.*th$"))
```

The `everything()` helper selects all remaining columns and is most useful for reordering.

```{r}
# Move Species to the front
select(iris, Species, everything())
```

The `last_col()` helper selects the last column, with an optional offset.

```{r}
# Select the last column
select(iris, last_col())
```

```{r}
# Select the second-to-last column
select(iris, last_col(offset = 1))
```

The `where()` helper selects columns based on a function that returns `TRUE` or `FALSE`.

```{r}
# Select only numeric columns
select(iris, where(is.numeric))
```

```{r}
# Select only factor columns
select(iris, where(is.factor))
```

### Renaming with select()

You can rename columns as you select them by using the syntax `new_name = old_name`.

```{r}
select(iris, sepal_length = Sepal.Length, species = Species)
```

Note that `select()` drops all columns you do not mention. If you want to rename columns without dropping any, use `rename()` instead (covered in a later section).

### Reordering Columns with select()

Using `select()` together with `everything()` is a convenient way to bring specific columns to the front while keeping all others.

```{r}
# Bring Petal columns to the front, keep everything else after
select(iris, starts_with("Petal"), everything())
```


## mutate() - Creating New Columns {#mutate}

The `mutate()` function adds new columns to a data frame or modifies existing columns. The original data frame is preserved; `mutate()` returns a new data frame with the additions.

### Creating New Variables

```{r}
# Add a column for the sepal area (approximation)
iris_modified <- mutate(iris, Sepal.Area = Sepal.Length * Sepal.Width)
select(iris_modified, Sepal.Length, Sepal.Width, Sepal.Area) |> head()
```

```{r}
# You can create multiple columns in a single mutate() call
# Columns created earlier can be referenced by later columns
mtcars_modified <- mutate(mtcars,
  weight_kg = wt * 453.592,
  hp_per_kg = hp / weight_kg
)
select(mtcars_modified, wt, weight_kg, hp, hp_per_kg) |> head()
```

### Modifying Existing Variables

If you assign to a column name that already exists, `mutate()` will overwrite it.

```{r}
# Convert wt from 1000 lbs to actual lbs
mutate(mtcars, wt = wt * 1000) |> select(wt) |> head()
```

### Useful Functions Inside mutate()

Many different functions can be used within `mutate()`. Here are some of the most common categories.

**Arithmetic operations**: `+`, `-`, `*`, `/`, `^`, `%%` (modulo), `%/%` (integer division).

```{r}
mutate(iris,
  Sepal.Sum = Sepal.Length + Sepal.Width,
  Petal.Ratio = Petal.Length / Petal.Width
) |> head()
```

**Logarithmic and mathematical functions**: `log()`, `log2()`, `log10()`, `sqrt()`, `abs()`.

```{r}
mutate(mtcars, log_hp = log(hp), sqrt_mpg = sqrt(mpg)) |>
  select(hp, log_hp, mpg, sqrt_mpg) |> head()
```

**Lead and lag**: `lead()` and `lag()` shift a column forward or backward by a specified number of positions. These are invaluable for computing differences between consecutive rows.

```{r}
example_df <- tibble(
  day = 1:7,
  sales = c(100, 120, 90, 150, 130, 170, 160)
)
mutate(example_df,
  previous_day_sales = lag(sales),
  next_day_sales = lead(sales),
  daily_change = sales - lag(sales)
)
```

**Cumulative functions**: `cumsum()`, `cumprod()`, `cummax()`, `cummin()`, `cummean()`.

```{r}
mutate(example_df,
  cumulative_sales = cumsum(sales),
  running_average = cummean(sales)
)
```

### Conditional Transformations with if_else() and case_when()

The `if_else()` function provides a vectorized if-else statement. It is stricter than base R's `ifelse()` because it checks that the true and false values have the same type.

```{r}
mutate(mtcars,
  efficiency = if_else(mpg > 20, "efficient", "inefficient")
) |> select(mpg, efficiency) |> head(10)
```

For multiple conditions, `case_when()` is more readable than nested `if_else()` calls. Each condition is written as a formula with the condition on the left of `~` and the result on the right. The `.default` argument handles values that do not match any condition.

```{r}
mutate(mtcars,
  mpg_category = case_when(
    mpg >= 30 ~ "excellent",
    mpg >= 20 ~ "good",
    mpg >= 15 ~ "moderate",
    .default = "poor"
  )
) |> select(mpg, mpg_category) |> head(10)
```

```{r}
# Classify iris flowers by sepal size
mutate(iris,
  sepal_size = case_when(
    Sepal.Length > 6.5 ~ "large",
    Sepal.Length > 5.5 ~ "medium",
    .default = "small"
  )
) |> count(sepal_size, Species)
```

### Using across() Inside mutate()

The `across()` function applies a transformation to multiple columns at once. This avoids repetitive code when you need to apply the same operation to many columns.

```{r}
# Round all numeric columns to 1 decimal place
mutate(iris, across(where(is.numeric), \(x) round(x, 1))) |> head()
```

```{r}
# Log-transform specific columns
mutate(mtcars, across(c(hp, wt, disp), log)) |>
  select(hp, wt, disp) |> head()
```

### transmute()

The `transmute()` function works like `mutate()` but keeps only the newly created columns, dropping all others.

```{r}
transmute(mtcars,
  car_name = rownames(mtcars),
  hp_per_ton = hp / wt,
  kpl = mpg * 0.425144
) |> head()
```


## summarize() - Summarizing Data {#summarize}

The `summarize()` function (also spelled `summarise()`) collapses a data frame into a single row of summary statistics. It is most powerful when combined with `group_by()`, but can also be used on its own.

### Basic Summary Statistics

```{r}
summarize(iris,
  mean_sepal_length = mean(Sepal.Length),
  sd_sepal_length = sd(Sepal.Length),
  median_sepal_length = median(Sepal.Length)
)
```

```{r}
summarize(mtcars,
  avg_mpg = mean(mpg),
  avg_hp = mean(hp),
  min_wt = min(wt),
  max_wt = max(wt)
)
```

### Useful Summary Functions

Here is a comprehensive list of functions commonly used inside `summarize()`:

| Function | Purpose |
|:---------------|:---------------------------------------|
| `mean()` | Arithmetic mean |
| `median()` | Median |
| `sd()` | Standard deviation |
| `IQR()` | Interquartile range |
| `min()` | Minimum value |
| `max()` | Maximum value |
| `sum()` | Sum of values |
| `n()` | Count of rows |
| `n_distinct()` | Count of unique values |
| `first()` | First value |
| `last()` | Last value |
| `nth()` | nth value |

```{r}
summarize(iris,
  n_rows = n(),
  n_species = n_distinct(Species),
  iqr_sepal_length = IQR(Sepal.Length),
  first_sepal = first(Sepal.Length),
  last_sepal = last(Sepal.Length),
  third_sepal = nth(Sepal.Length, 3)
)
```

### Multiple Summaries with across()

Rather than typing out a summary for every column, you can use `across()` to apply summary functions to multiple columns at once.

```{r}
# Mean of all numeric columns
iris |>
  summarize(across(where(is.numeric), mean))
```

```{r}
# Multiple summary functions at once using a named list
iris |>
  summarize(across(
    where(is.numeric),
    list(mean = mean, sd = sd),
    .names = "{.col}_{.fn}"
  ))
```

The `.names` argument controls the naming of the output columns. The `{.col}` placeholder is replaced by the column name and `{.fn}` by the function name.

```{r}
# Summarize selected columns with multiple functions
mtcars |>
  summarize(across(
    c(mpg, hp, wt),
    list(avg = mean, minimum = min, maximum = max),
    .names = "{.fn}_{.col}"
  ))
```


## arrange() - Sorting Rows {#arrange}

The `arrange()` function reorders rows. By default, it sorts in ascending order.

```{r}
# Sort mtcars by mpg (ascending)
arrange(mtcars, mpg) |> head()
```

### Descending Order

Use `desc()` to sort in descending order.

```{r}
# Sort by mpg, highest first
arrange(mtcars, desc(mpg)) |> head()
```

### Sorting by Multiple Columns

When you sort by multiple columns, ties in the first column are broken by the second column, and so on.

```{r}
# Sort by number of cylinders, then by mpg within each cylinder group
arrange(mtcars, cyl, desc(mpg)) |> head(12)
```

```{r}
# Sort iris by Species, then by Sepal.Length within species
arrange(iris, Species, desc(Sepal.Length)) |> head(10)
```

Note that `NA` values are always sorted to the end, regardless of whether you use ascending or descending order.


## group_by() - Grouped Operations {#groupby}

The `group_by()` function does not change the data itself. Instead, it adds grouping metadata so that subsequent operations (especially `summarize()`, `mutate()`, and `filter()`) are performed separately for each group.

### Grouped Summaries

This is the most common use of `group_by()`: computing summary statistics for each group.

```{r}
iris |>
  group_by(Species) |>
  summarize(
    mean_sepal_length = mean(Sepal.Length),
    mean_sepal_width = mean(Sepal.Width),
    mean_petal_length = mean(Petal.Length),
    mean_petal_width = mean(Petal.Width),
    count = n()
  )
```

```{r}
mtcars |>
  group_by(cyl) |>
  summarize(
    avg_mpg = mean(mpg),
    avg_hp = mean(hp),
    n_cars = n()
  )
```

### Multiple Grouping Variables

You can group by more than one variable. The summary will be computed for every combination of the grouping variables.

```{r}
mtcars |>
  group_by(cyl, am) |>
  summarize(
    avg_mpg = mean(mpg),
    n_cars = n(),
    .groups = "drop"
  )
```

The `.groups` argument in `summarize()` controls what happens to the grouping after summarization. Setting `.groups = "drop"` removes all grouping. The default behavior (`.groups = "drop_last"`) drops only the last grouping variable, leaving the remaining groups in place.

### Grouped Mutate

When `mutate()` is applied to grouped data, calculations are performed within each group. This is useful for computing group-relative statistics.

```{r}
# Compute each flower's deviation from its species mean
iris |>
  group_by(Species) |>
  mutate(
    sepal_deviation = Sepal.Length - mean(Sepal.Length)
  ) |>
  select(Species, Sepal.Length, sepal_deviation) |>
  head(10)
```

```{r}
# Rank cars within each cylinder group by mpg
mtcars |>
  group_by(cyl) |>
  mutate(mpg_rank = min_rank(desc(mpg))) |>
  select(cyl, mpg, mpg_rank) |>
  arrange(cyl, mpg_rank) |>
  head(12)
```

### Grouped Filter

A grouped `filter()` applies the condition within each group. This lets you write filters relative to group statistics.

```{r}
# Keep only cars with above-average mpg within their cylinder group
mtcars |>
  group_by(cyl) |>
  filter(mpg > mean(mpg)) |>
  select(cyl, mpg) |>
  arrange(cyl, desc(mpg))
```

### Removing Grouping with ungroup()

Grouping persists until explicitly removed. If you forget to ungroup, subsequent operations may produce unexpected results. Use `ungroup()` to remove grouping.

```{r}
iris_grouped <- iris |> group_by(Species)

# Check the groups
group_vars(iris_grouped)

# Remove grouping
iris_ungrouped <- iris_grouped |> ungroup()
group_vars(iris_ungrouped)
```


## Combining Operations {#combining}

The real power of dplyr emerges when you chain multiple verbs together using the pipe operator. Each step flows naturally into the next, creating readable analysis pipelines.

### Example: Analyzing Fuel Efficiency

```{r}
mtcars |>
  mutate(car_name = rownames(mtcars)) |>
  filter(cyl %in% c(4, 6)) |>
  select(car_name, mpg, cyl, hp, wt) |>
  group_by(cyl) |>
  mutate(mpg_rank = min_rank(desc(mpg))) |>
  arrange(cyl, mpg_rank) |>
  filter(mpg_rank <= 3) |>
  ungroup()
```

This pipeline answers the question: "What are the top 3 most fuel-efficient cars in the 4-cylinder and 6-cylinder groups?"

### Example: Species Comparison

```{r}
iris |>
  group_by(Species) |>
  summarize(
    across(everything(), list(mean = mean, sd = sd)),
    .groups = "drop"
  ) |>
  mutate(across(where(is.numeric), \(x) round(x, 2)))
```

### Example: Car Performance Profile

```{r}
mtcars |>
  mutate(
    transmission = if_else(am == 1, "manual", "automatic"),
    engine = if_else(vs == 1, "straight", "V-shaped")
  ) |>
  group_by(transmission, engine) |>
  summarize(
    n = n(),
    avg_mpg = round(mean(mpg), 1),
    avg_hp = round(mean(hp), 1),
    avg_qsec = round(mean(qsec), 1),
    .groups = "drop"
  ) |>
  arrange(desc(avg_mpg))
```


## Joining Data {#joins}

In practice, data is often spread across multiple tables that need to be combined. The dplyr package provides a family of join functions modeled after SQL joins.

### Example Datasets for Joins

Let us create two small data frames to illustrate joins clearly.

```{r}
# A table of students and their majors
students <- tibble(
  student_id = c(1, 2, 3, 4, 5),
  name = c("Alice", "Bob", "Carol", "David", "Eve"),
  major = c("Statistics", "CS", "Statistics", "Math", "CS")
)
students

# A table of students and their GPAs (note: not all students appear)
grades <- tibble(
  student_id = c(1, 2, 3, 6, 7),
  gpa = c(3.8, 3.5, 3.9, 3.2, 3.7),
  year = c(2023, 2023, 2024, 2023, 2024)
)
grades
```

Notice that students 4 and 5 appear only in `students`, while students 6 and 7 appear only in `grades`. Students 1, 2, and 3 appear in both.

### inner_join()

`inner_join()` keeps only rows that have matching keys in **both** tables.

```{r}
inner_join(students, grades, by = "student_id")
```

Only students 1, 2, and 3 appear because they are the only ones present in both tables.

### left_join()

`left_join()` keeps all rows from the **left** (first) table and matches from the right. Unmatched rows from the right table are filled with `NA`.

```{r}
left_join(students, grades, by = "student_id")
```

All five students from the `students` table are kept. David and Eve have `NA` for gpa and year because they had no match in `grades`.

### right_join()

`right_join()` keeps all rows from the **right** (second) table and matches from the left.

```{r}
right_join(students, grades, by = "student_id")
```

All rows from `grades` are kept. Students 6 and 7 have `NA` for name and major.

### full_join()

`full_join()` keeps all rows from **both** tables, filling in `NA` where there is no match.

```{r}
full_join(students, grades, by = "student_id")
```

### semi_join()

`semi_join()` returns rows from the left table that have a match in the right table, but does **not** add columns from the right table. Think of it as a filter using the right table.

```{r}
semi_join(students, grades, by = "student_id")
```

### anti_join()

`anti_join()` returns rows from the left table that do **not** have a match in the right table. It is the inverse of `semi_join()`.

```{r}
anti_join(students, grades, by = "student_id")
```

David and Eve are returned because they have no matching records in `grades`.

### Joining with Different Column Names

When the key columns have different names in the two tables, use the `join_by()` helper or a named vector in the `by` argument.

```{r}
# Suppose the grades table used "id" instead of "student_id"
grades_renamed <- grades |> rename(id = student_id)

# Using join_by()
left_join(students, grades_renamed, by = join_by(student_id == id))
```

### Joining on Multiple Keys

You can join on multiple columns by specifying several names in the `by` argument.

```{r}
# Create example tables with composite keys
orders <- tibble(
  customer = c("A", "A", "B", "B"),
  product = c("X", "Y", "X", "Z"),
  quantity = c(10, 5, 8, 12)
)

prices <- tibble(
  customer = c("A", "A", "B", "C"),
  product = c("X", "Y", "X", "X"),
  price = c(2.50, 3.00, 2.75, 2.50)
)

inner_join(orders, prices, by = c("customer", "product"))
```


## Other Useful Functions {#other-functions}

Beyond the core verbs, dplyr includes many additional functions that handle common tasks.

### The slice() Family

The `slice()` functions select rows by position rather than by condition.

```{r}
# Select rows 5 through 10
slice(iris, 5:10)
```

```{r}
# First 5 rows
slice_head(iris, n = 5)
```

```{r}
# Last 3 rows
slice_tail(mtcars, n = 3)
```

```{r}
# Random sample of rows
set.seed(42)
slice_sample(iris, n = 5)
```

```{r}
# Random fraction of rows
set.seed(42)
slice_sample(iris, prop = 0.05)
```

```{r}
# Rows with the 3 smallest values of mpg
slice_min(mtcars, mpg, n = 3)
```

```{r}
# Rows with the 3 largest values of hp
slice_max(mtcars, hp, n = 3)
```

### distinct()

The `distinct()` function removes duplicate rows. You can specify columns to consider; if a subset of columns is given, only the first occurrence of each unique combination is kept.

```{r}
# Unique combinations of cyl and am
distinct(mtcars, cyl, am)
```

```{r}
# Keep all columns while de-duplicating by cyl
distinct(mtcars, cyl, .keep_all = TRUE)
```

### count() and add_count()

The `count()` function is a shortcut for `group_by()` followed by `summarize(n = n())`. It is extremely convenient for quick frequency tables.

```{r}
count(iris, Species)
```

```{r}
# Count with sorting
count(mtcars, cyl, sort = TRUE)
```

```{r}
# Count by multiple variables
count(mtcars, cyl, am)
```

The `add_count()` function does the same but adds the count as a new column rather than collapsing the data frame.

```{r}
# Add a column with the count of each species
iris |>
  add_count(Species, name = "species_count") |>
  select(Species, species_count) |>
  distinct()
```

### pull()

The `pull()` function extracts a single column as a vector rather than a data frame. This is useful when you need a vector for further computation.

```{r}
# Extract mpg as a vector
mpg_values <- pull(mtcars, mpg)
mean(mpg_values)
```

```{r}
# Pull the last column
pull(iris, last_col()) |> head(10)
```

### rename() and rename_with()

While `select()` can rename columns, it drops unmentioned columns. The `rename()` function renames columns while keeping all other columns intact.

```{r}
rename(iris, sepal_length = Sepal.Length, sepal_width = Sepal.Width) |> head(3)
```

The `rename_with()` function applies a renaming function to selected columns.

```{r}
# Convert all column names to lowercase
rename_with(iris, tolower) |> head(3)
```

```{r}
# Replace dots with underscores in column names
rename_with(iris, \(x) gsub("\\.", "_", x)) |> head(3)
```

```{r}
# Apply tolower only to columns that start with "Sepal"
rename_with(iris, tolower, starts_with("Sepal")) |> head(3)
```

### relocate()

The `relocate()` function moves columns to a new position. By default, it moves the specified columns to the front.

```{r}
# Move Species to the front
relocate(iris, Species) |> head(3)
```

```{r}
# Move Petal columns after Species
relocate(iris, starts_with("Petal"), .after = Species) |> head(3)
```

```{r}
# Move numeric columns before factor columns
relocate(iris, where(is.numeric), .before = where(is.factor)) |> head(3)
```

### bind_rows() and bind_cols()

These functions combine data frames by stacking rows or appending columns.

```{r}
# Bind rows: stack data frames vertically
df1 <- tibble(x = 1:3, y = c("a", "b", "c"))
df2 <- tibble(x = 4:6, y = c("d", "e", "f"))
bind_rows(df1, df2)
```

```{r}
# bind_rows() can add an .id column to track the source
bind_rows(first = df1, second = df2, .id = "source")
```

```{r}
# Bind columns: combine data frames side by side
df3 <- tibble(z = c(10, 20, 30))
bind_cols(df1, df3)
```

Note that `bind_cols()` requires the data frames to have the same number of rows.


## Working with across() {#across}

The `across()` function is a powerful tool for applying transformations to multiple columns simultaneously. It can be used inside `mutate()`, `summarize()`, and other dplyr verbs.

### Basic Syntax

The basic syntax is `across(.cols, .fns)`, where `.cols` specifies which columns to operate on and `.fns` specifies the function(s) to apply.

```{r}
# Scale all numeric columns to have mean 0 and sd 1
iris |>
  mutate(across(where(is.numeric), scale)) |>
  head()
```

### Column Selection Methods

You can select columns for `across()` using any of the tidyselect helpers.

```{r}
# Using where() to select by type
iris |>
  summarize(across(where(is.numeric), mean))
```

```{r}
# Using starts_with()
iris |>
  summarize(across(starts_with("Sepal"), list(mean = mean, sd = sd)))
```

```{r}
# Using everything() - all columns (works best when all are numeric)
mtcars |>
  summarize(across(everything(), mean)) |>
  round(2)
```

```{r}
# Using specific column names
mtcars |>
  summarize(across(c(mpg, hp, wt), median))
```

### Applying Multiple Functions

Pass a named list to `.fns` to apply multiple functions. The `.names` argument controls the naming of the resulting columns.

```{r}
iris |>
  group_by(Species) |>
  summarize(
    across(
      starts_with("Petal"),
      list(
        mean = \(x) mean(x),
        sd = \(x) sd(x),
        min = \(x) min(x),
        max = \(x) max(x)
      ),
      .names = "{.col}__{.fn}"
    ),
    .groups = "drop"
  )
```

### Using across() with mutate()

```{r}
# Convert all numeric columns from raw values to percentiles (within species)
iris |>
  group_by(Species) |>
  mutate(across(
    where(is.numeric),
    \(x) percent_rank(x),
    .names = "{.col}_pctile"
  )) |>
  select(Species, ends_with("pctile")) |>
  head(10)
```

```{r}
# Apply different functions to different columns using multiple across() calls
mtcars |>
  mutate(
    across(c(mpg, hp), \(x) round(x, 0)),
    across(c(wt, qsec), \(x) round(x, 1))
  ) |>
  select(mpg, hp, wt, qsec) |>
  head()
```

### Combining across() with case_when()

```{r}
# Categorize multiple numeric columns
iris |>
  mutate(across(
    where(is.numeric),
    \(x) case_when(
      x > mean(x) + sd(x) ~ "high",
      x < mean(x) - sd(x) ~ "low",
      .default = "medium"
    ),
    .names = "{.col}_cat"
  )) |>
  select(Species, ends_with("_cat")) |>
  head(10)
```


## Summary

The dplyr package provides a complete and coherent grammar for data manipulation. Its six core verbs --- `filter()`, `select()`, `mutate()`, `summarize()`, `arrange()`, and `group_by()` --- cover the vast majority of data wrangling tasks. Combined with the pipe operator, these verbs enable you to build readable, step-by-step analysis pipelines. The join functions handle the common need to combine data from multiple sources, and helper functions like `across()`, `count()`, `slice()`, and `relocate()` provide additional convenience for everyday tasks.

The consistent design of dplyr --- every function takes a data frame first and returns a data frame --- means that once you learn one verb, the others follow naturally. As you gain experience, you will find that even complex transformations can be expressed as a clear sequence of simple steps. The next chapter on tidyr will complement your dplyr skills by covering how to reshape data between wide and long formats.
