# Data Wrangling with dplyr {#dplyr}

```{r include=FALSE}
library(tidyverse)
library(knitr)
opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

If there's one chapter in this book that will directly make you better at your job, it's this one. Data wrangling -- filtering, sorting, summarizing, reshaping -- is what analysts spend most of their time doing. And dplyr makes it almost enjoyable.

Think of dplyr as a set of power tools for your data. Instead of clicking around in Excel or writing complicated formulas, you write clear, readable code that says exactly what you want. "Show me only the rows where revenue is above $10,000." "Add a column that calculates profit margin." "Give me average sales by region." That's dplyr in a nutshell.

The best part? Every dplyr function takes a data frame in and gives a data frame back. So you can chain them together like an assembly line. But we'll get to that.


## The Six Verbs You Need to Know {#core-verbs}

Almost everything you'll do with dplyr comes down to six functions. Here they are, in plain English:

| Verb | What It Does | Excel Equivalent |
|:-------------|:-----------------------------------------|:--------------------------|
| `filter()` | Show me only the rows I care about | Filtering rows |
| `select()` | Show me only the columns I care about | Hiding columns |
| `mutate()` | Add a new column (or change one) | Adding a formula column |
| `summarize()` | Give me summary stats | SUM, AVERAGE, COUNT |
| `arrange()` | Sort my data | Sort A-Z / Z-A |
| `group_by()` | Do it separately for each group | Pivot tables! |

Every one of these functions works the same way:

1. First argument: your data.
2. Next arguments: what you want to do, using column names directly (no quotes needed).
3. Result: a new data frame.

Let's dig in.


## filter() -- Show Me Only the Rows I Care About {#filter}

You know how in Excel you can filter a column to show only certain values? `filter()` does the same thing, but with way more power.

### Basic Comparisons

```{r}
# Cars that get more than 30 miles per gallon
filter(mtcars, mpg > 30)
```

```{r}
# Only setosa flowers
filter(iris, Species == "setosa")
```

```{r}
# Cars with exactly 6 cylinders
filter(mtcars, cyl == 6)
```

The comparison operators are what you'd expect: `==` (equals), `!=` (not equal), `>`, `<`, `>=`, `<=`.

One gotcha: it's `==` for comparison, not `=`. A single `=` is for assigning values. This trips up everyone at least once.

### Combining Conditions

Need multiple conditions? Use `&` (and) or `|` (or).

```{r}
# Setosa flowers with long sepals
filter(iris, Species == "setosa" & Sepal.Length > 5)
```

Here's a nice shortcut: when you separate conditions with commas inside `filter()`, R treats them as "and" automatically.

```{r}
# Same thing, cleaner syntax
filter(iris, Species == "setosa", Sepal.Length > 5)
```

```{r}
# Cars with 4 cylinders OR great gas mileage
filter(mtcars, cyl == 4 | mpg > 30)
```

### The %in% Operator

When you want to match against a list of values, `%in%` is your friend. Think of it as "is this value in this set?"

```{r}
# Setosa or virginica (but not versicolor)
filter(iris, Species %in% c("setosa", "virginica"))
```

```{r}
# 4-cylinder or 8-cylinder cars
filter(mtcars, cyl %in% c(4, 8))
```

### Filtering Ranges

The `between()` helper is perfect for "give me everything from X to Y" situations.

```{r}
# Sepal length between 5.0 and 6.0
filter(iris, between(Sepal.Length, 5.0, 6.0))
```

### Dealing with Missing Values

Missing data (`NA`) is a fact of life in business. Someone didn't fill out the survey field, a sensor malfunctioned, a record got corrupted. You can't test for `NA` with `==`; you need `is.na()`.

```{r}
df_with_na <- tibble(
  x = c(1, 2, NA, 4, 5),
  y = c("a", NA, "c", "d", NA)
)

# Keep only rows where x is not missing
filter(df_with_na, !is.na(x))
```

```{r}
# Keep rows where BOTH x and y are present
filter(df_with_na, !is.na(x), !is.na(y))
```

### Putting It Together

Real filtering usually combines several conditions. Here's the kind of thing you'll actually write at work:

```{r}
# Efficient cars: 4 cylinders with above-average mpg
filter(mtcars, cyl == 4, mpg > mean(mpg))
```


## select() -- Show Me Only the Columns I Care About {#select}

Your dataset has 47 columns but you only need 5 of them? `select()` strips it down.

### Picking Columns

```{r}
# Just these two columns, please
select(iris, Sepal.Length, Species)
```

```{r}
# A range of columns
select(iris, Sepal.Length:Petal.Length)
```

```{r}
# Everything EXCEPT Species
select(iris, -Species)
```

### Smart Selection Helpers

When your data has lots of similarly named columns (like `Q1_Revenue`, `Q2_Revenue`, `Q3_Revenue`...), these helpers save tons of typing.

```{r}
# All columns that start with "Sepal"
select(iris, starts_with("Sepal"))
```

```{r}
# All columns that end with "Width"
select(iris, ends_with("Width"))
```

```{r}
# All columns that contain "pal"
select(iris, contains("pal"))
```

```{r}
# Only numeric columns
select(iris, where(is.numeric))
```

### Reordering and Renaming

Want to move a column to the front? Use `everything()` to keep the rest.

```{r}
# Put Species first, keep everything else
select(iris, Species, everything())
```

You can rename as you select:

```{r}
select(iris, sepal_length = Sepal.Length, species = Species)
```

If you want to rename without dropping columns, use `rename()`:

```{r}
rename(iris, sepal_length = Sepal.Length, sepal_width = Sepal.Width) |> head(3)
```


## mutate() -- Add a New Column {#mutate}

This is like adding a formula column in Excel. You define a new column based on existing ones.

### Creating New Columns

```{r}
# Approximate sepal area
iris_modified <- mutate(iris, Sepal.Area = Sepal.Length * Sepal.Width)
select(iris_modified, Sepal.Length, Sepal.Width, Sepal.Area) |> head()
```

```{r}
# You can create multiple columns at once
# Later columns can reference earlier ones (like chained formulas)
mtcars_modified <- mutate(mtcars,
  weight_kg = wt * 453.592,
  hp_per_kg = hp / weight_kg
)
select(mtcars_modified, wt, weight_kg, hp, hp_per_kg) |> head()
```

### Overwriting Existing Columns

If you assign to a name that already exists, it overwrites.

```{r}
# Convert wt from thousands of pounds to actual pounds
mutate(mtcars, wt = wt * 1000) |> select(wt) |> head()
```

### Useful Functions Inside mutate()

**Arithmetic** -- the usual suspects: `+`, `-`, `*`, `/`.

```{r}
mutate(iris,
  Sepal.Sum = Sepal.Length + Sepal.Width,
  Petal.Ratio = Petal.Length / Petal.Width
) |> head()
```

**Lag and lead** -- great for "compared to last month" calculations.

```{r}
example_df <- tibble(
  day = 1:7,
  sales = c(100, 120, 90, 150, 130, 170, 160)
)
mutate(example_df,
  previous_day_sales = lag(sales),
  daily_change = sales - lag(sales)
)
```

**Cumulative functions** -- running totals, running averages.

```{r}
mutate(example_df,
  cumulative_sales = cumsum(sales),
  running_average = cummean(sales)
)
```

### if_else() and case_when() -- Conditional Logic

`if_else()` is your basic "if this, then that" tool.

```{r}
mutate(mtcars,
  efficiency = if_else(mpg > 20, "efficient", "inefficient")
) |> select(mpg, efficiency) |> head(10)
```

`case_when()` handles multiple conditions. Think of it as a series of "if... else if... else if..." checks. Way more readable than nesting a dozen IF statements in Excel.

```{r}
mutate(mtcars,
  mpg_category = case_when(
    mpg >= 30 ~ "excellent",
    mpg >= 20 ~ "good",
    mpg >= 15 ~ "moderate",
    .default = "poor"
  )
) |> select(mpg, mpg_category) |> head(10)
```

This is incredibly useful for bucketing data. Customer segments, performance tiers, risk categories -- `case_when()` handles them all.

```{r}
mutate(iris,
  sepal_size = case_when(
    Sepal.Length > 6.5 ~ "large",
    Sepal.Length > 5.5 ~ "medium",
    .default = "small"
  )
) |> count(sepal_size, Species)
```


## summarize() -- Give Me the Summary Stats {#summarize}

`summarize()` (or `summarise()` if you prefer British spelling) collapses your data down to summary numbers. On its own, it gives you one row of results. Pair it with `group_by()` and it becomes a pivot table machine.

### The Basics

```{r}
summarize(mtcars,
  avg_mpg = mean(mpg),
  avg_hp = mean(hp),
  min_wt = min(wt),
  max_wt = max(wt)
)
```

### Common Summary Functions

Here's your toolkit:

| Function | What It Does |
|:---------------|:---------------------------------------|
| `mean()` | Average |
| `median()` | Median |
| `sd()` | Standard deviation |
| `min()`, `max()` | Smallest / largest value |
| `sum()` | Total |
| `n()` | Count of rows |
| `n_distinct()` | Count of unique values |

```{r}
summarize(iris,
  n_rows = n(),
  n_species = n_distinct(Species),
  iqr_sepal_length = IQR(Sepal.Length)
)
```

### Summarizing Multiple Columns with across()

Instead of writing out a summary for every single column, `across()` lets you apply the same function to multiple columns at once.

```{r}
# Mean of all numeric columns
iris |>
  summarize(across(where(is.numeric), mean))
```

```{r}
# Multiple stats at once
iris |>
  summarize(across(
    where(is.numeric),
    list(mean = mean, sd = sd),
    .names = "{.col}_{.fn}"
  ))
```


## arrange() -- Sort My Data {#arrange}

Sorting. Simple but essential.

```{r}
# Sort by mpg (lowest first)
arrange(mtcars, mpg) |> head()
```

### Descending Order

Wrap the column in `desc()` to flip the sort.

```{r}
# Best gas mileage first
arrange(mtcars, desc(mpg)) |> head()
```

### Sorting by Multiple Columns

Ties in the first column get broken by the second.

```{r}
# Sort by cylinders, then by mpg within each group
arrange(mtcars, cyl, desc(mpg)) |> head(12)
```

Note: `NA` values always get sorted to the end, no matter what.


## group_by() -- The Pivot Table of R {#groupby}

This is where dplyr starts to feel magical. `group_by()` tells R: "Hey, for the next operation, do it separately for each group." It's exactly like creating a pivot table in Excel, but more flexible.

### Grouped Summaries

This is the most common combo: `group_by()` + `summarize()`.

```{r}
iris |>
  group_by(Species) |>
  summarize(
    mean_sepal_length = mean(Sepal.Length),
    mean_petal_length = mean(Petal.Length),
    count = n()
  )
```

```{r}
mtcars |>
  group_by(cyl) |>
  summarize(
    avg_mpg = mean(mpg),
    avg_hp = mean(hp),
    n_cars = n()
  )
```

Just like that, you've got a pivot table. Average MPG and horsepower, broken out by cylinder count, with a count of how many cars are in each group.

### Multiple Grouping Variables

You can group by more than one thing. R computes summaries for every combination.

```{r}
mtcars |>
  group_by(cyl, am) |>
  summarize(
    avg_mpg = mean(mpg),
    n_cars = n(),
    .groups = "drop"
  )
```

The `.groups = "drop"` at the end removes the grouping when you're done. Good habit to get into -- leftover grouping can cause surprises later.

### Grouped Mutate

When you use `mutate()` on grouped data, calculations happen within each group. This is great for "compared to the group average" analysis.

```{r}
# How does each flower compare to its species average?
iris |>
  group_by(Species) |>
  mutate(
    sepal_deviation = Sepal.Length - mean(Sepal.Length)
  ) |>
  select(Species, Sepal.Length, sepal_deviation) |>
  head(10)
```

### Grouped Filter

You can also filter within groups. "Show me only the above-average performers in each department."

```{r}
# Cars with above-average mpg for their cylinder class
mtcars |>
  group_by(cyl) |>
  filter(mpg > mean(mpg)) |>
  select(cyl, mpg) |>
  arrange(cyl, desc(mpg))
```

### Don't Forget to ungroup()

Grouping is sticky -- it persists until you explicitly remove it. If your results look weird, check whether there's leftover grouping.

```{r}
iris_grouped <- iris |> group_by(Species)
group_vars(iris_grouped)

iris_ungrouped <- iris_grouped |> ungroup()
group_vars(iris_ungrouped)
```


## Combining Operations -- Where the Magic Happens {#combining}

Here's the payoff. Everything we've learned chains together with the pipe operator (`|>`). Each step feeds into the next, creating a readable analysis pipeline. Read these out loud and they almost sound like English.

### Example: Top Fuel-Efficient Cars by Class

"Take the car data. Add car names. Keep only 4- and 6-cylinder cars. Select the columns I care about. Within each cylinder group, rank them by mpg. Show me the top 3 in each group."

```{r}
mtcars |>
  mutate(car_name = rownames(mtcars)) |>
  filter(cyl %in% c(4, 6)) |>
  select(car_name, mpg, cyl, hp, wt) |>
  group_by(cyl) |>
  mutate(mpg_rank = min_rank(desc(mpg))) |>
  arrange(cyl, mpg_rank) |>
  filter(mpg_rank <= 3) |>
  ungroup()
```

That pipeline answers a real question in eight clean lines. Try doing that in Excel without a mess of helper columns.

### Example: Performance Report

"Take the car data. Label the transmissions and engine types with readable names. Group by transmission and engine. Compute averages. Sort by fuel efficiency."

```{r}
mtcars |>
  mutate(
    transmission = if_else(am == 1, "manual", "automatic"),
    engine = if_else(vs == 1, "straight", "V-shaped")
  ) |>
  group_by(transmission, engine) |>
  summarize(
    n = n(),
    avg_mpg = round(mean(mpg), 1),
    avg_hp = round(mean(hp), 1),
    avg_qsec = round(mean(qsec), 1),
    .groups = "drop"
  ) |>
  arrange(desc(avg_mpg))
```

This is exactly the kind of summary table your manager wants to see. Clean, grouped, sorted, with readable labels.

### Example: Species Comparison with Full Stats

```{r}
iris |>
  group_by(Species) |>
  summarize(
    across(everything(), list(mean = mean, sd = sd)),
    .groups = "drop"
  ) |>
  mutate(across(where(is.numeric), \(x) round(x, 2)))
```


## Joining Data -- Connecting Your Tables {#joins}

In any real business, data lives in multiple tables. You've got a customer table and an orders table. An employee table and a payroll table. A product table and an inventory table. Joins let you connect them.

### Setting Up the Example

Let's use a scenario you'd actually encounter: customers and their orders.

```{r}
customers <- tibble(
  customer_id = c(1, 2, 3, 4, 5),
  name = c("Alice", "Bob", "Carol", "David", "Eve"),
  region = c("East", "West", "East", "South", "West")
)
customers

orders <- tibble(
  customer_id = c(1, 2, 3, 6, 7),
  revenue = c(500, 320, 890, 150, 410),
  quarter = c("Q1", "Q1", "Q2", "Q1", "Q2")
)
orders
```

Notice: customers 4 and 5 haven't placed orders. Customers 6 and 7 placed orders but aren't in our customer table (maybe they're new and haven't been entered yet).

### left_join() -- The One You'll Use 90% of the Time

`left_join()` keeps everything from the left table and pulls in matching data from the right table. No match? You get `NA`. This is almost always what you want.

```{r}
left_join(customers, orders, by = "customer_id")
```

All five customers are here. David and Eve show `NA` for revenue and quarter because they had no orders. Customers 6 and 7 from the orders table are excluded because they weren't in the customer table.

This is your go-to join. "Start with my main table, and bring in extra info where available."

### inner_join() -- Only the Matches

`inner_join()` keeps only rows that appear in both tables.

```{r}
inner_join(customers, orders, by = "customer_id")
```

Only Alice, Bob, and Carol survive -- they're the only ones in both tables.

### Other Joins (Quick Reference)

**right_join()** -- like `left_join()` but keeps everything from the right table instead.

```{r}
right_join(customers, orders, by = "customer_id")
```

**full_join()** -- keeps everything from both tables. Missing data gets `NA`.

```{r}
full_join(customers, orders, by = "customer_id")
```

**anti_join()** -- "show me customers who DON'T have orders." Great for finding gaps in your data.

```{r}
anti_join(customers, orders, by = "customer_id")
```

### When Column Names Don't Match

Sometimes the key column has different names in each table. Use `join_by()`.

```{r}
orders_renamed <- orders |> rename(id = customer_id)

left_join(customers, orders_renamed, by = join_by(customer_id == id))
```

### Joining on Multiple Columns

You can join on more than one column.

```{r}
order_details <- tibble(
  customer = c("A", "A", "B", "B"),
  product = c("X", "Y", "X", "Z"),
  quantity = c(10, 5, 8, 12)
)

prices <- tibble(
  customer = c("A", "A", "B", "C"),
  product = c("X", "Y", "X", "X"),
  price = c(2.50, 3.00, 2.75, 2.50)
)

inner_join(order_details, prices, by = c("customer", "product"))
```


## Other Handy Functions {#other-functions}

Here are some extras that come up all the time in practice.

### count() -- Quick Frequency Tables

`count()` is a shortcut for "group by this, then count the rows." You'll use it constantly.

```{r}
count(iris, Species)
```

```{r}
# Sort by frequency
count(mtcars, cyl, sort = TRUE)
```

```{r}
# Count by multiple variables
count(mtcars, cyl, am)
```

### The slice() Family -- Rows by Position

```{r}
# First 5 rows
slice_head(iris, n = 5)
```

```{r}
# Top 3 by horsepower
slice_max(mtcars, hp, n = 3)
```

```{r}
# Bottom 3 by mpg
slice_min(mtcars, mpg, n = 3)
```

```{r}
# Random sample
set.seed(42)
slice_sample(iris, n = 5)
```

### distinct() -- Remove Duplicates

```{r}
# Unique combinations of cylinders and transmission
distinct(mtcars, cyl, am)
```

### pull() -- Extract a Column as a Vector

Sometimes you need a plain vector, not a data frame.

```{r}
mpg_values <- pull(mtcars, mpg)
mean(mpg_values)
```

### relocate() -- Move Columns Around

```{r}
# Move Species to the front
relocate(iris, Species) |> head(3)
```

### bind_rows() -- Stack Data Frames

When you have data in separate tables with the same columns, stack them vertically.

```{r}
df1 <- tibble(x = 1:3, y = c("a", "b", "c"))
df2 <- tibble(x = 4:6, y = c("d", "e", "f"))
bind_rows(df1, df2)
```

```{r}
# Track which table each row came from
bind_rows(first = df1, second = df2, .id = "source")
```


## Summary

Here's the bottom line: dplyr gives you six verbs that handle 95% of data wrangling tasks.

- **`filter()`** -- pick the rows you want
- **`select()`** -- pick the columns you want
- **`mutate()`** -- add or change columns
- **`summarize()`** -- compute summary statistics
- **`arrange()`** -- sort your data
- **`group_by()`** -- do any of the above by group (pivot tables!)

Chain them together with the pipe (`|>`), and you can build analysis pipelines that are powerful and readable. Add in `left_join()` to connect your tables, `count()` for quick frequency tables, and `case_when()` for conditional logic, and you've got a toolkit that handles just about anything a business analyst needs.

The next chapter on tidyr will show you how to reshape data between wide and long formats -- the final piece of the data wrangling puzzle.
