# Tidy Data with tidyr {#tidyr}

```{r include=FALSE}
library(tidyverse)
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

Let's talk about the single most important concept that will save you hours of frustration: **tidy data**. Tidy data is just your data in the format that R, ggplot2, and basically every tidyverse function *expects*. Get your data tidy first, and everything downstream -- charts, summaries, models -- becomes dramatically easier.

The **tidyr** package is your toolkit for reshaping messy data into tidy data. And trust me, you will use it constantly. Every CSV export from a CRM, every spreadsheet your boss emails you, every quarterly report pulled from a database -- they all need some tidying.

## What is Tidy Data?

Three rules. That's it:

1. **Each variable gets its own column.** Revenue is a column. Region is a column. Quarter is a column.
2. **Each observation gets its own row.** One row per thing you measured.
3. **Each cell contains one value.** No cramming "Q1/2024" into a single cell.

Here's what tidy data looks like:

```{r}
tidy_example <- tribble(
  ~country,     ~year, ~cases, ~population,
  "Afghanistan", 2000,    745,    20000000,
  "Afghanistan", 2001,   2666,    20595360,
  "Brazil",      2000,  37737,   172006362,
  "Brazil",      2001,  80488,   174504898,
  "China",       2000, 212258,  1280428583,
  "China",       2001, 213766,  1272915272
)

tidy_example
```

Each variable (`country`, `year`, `cases`, `population`) is a column. Each row is one observation (a country in a given year). Each cell has exactly one value. Clean, simple, ready to analyze.

## Common Ways Data Gets Messy

Real-world data violates these rules all the time. Here are the patterns you will see over and over again.

### Column Headers That Are Actually Values

This is the number-one offender. Picture this: your boss hands you a spreadsheet where the column headers are months, or years, or product names. The columns themselves are *values* of a variable, not variable names.

```{r}
messy_wide <- tribble(
  ~country,      ~`2000`, ~`2001`, ~`2002`,
  "Afghanistan",     745,    2666,    2900,
  "Brazil",        37737,   80488,   90125,
  "China",        212258,  213766,  220000
)

messy_wide
```

See the problem? `2000`, `2001`, and `2002` aren't variable names -- they're values of a `year` variable. The actual measurement (cases) is smeared across three columns. We need to **pivot** this into a longer format.

### Two Variables Jammed into One Column

Sometimes someone thought it was a good idea to cram two values into a single cell with a slash or a dash.

```{r}
messy_combined <- tribble(
  ~country,     ~year, ~rate,
  "Afghanistan", 2000, "745/20000000",
  "Afghanistan", 2001, "2666/20595360",
  "Brazil",      2000, "37737/172006362",
  "Brazil",      2001, "80488/174504898"
)

messy_combined
```

That `rate` column is hiding both `cases` and `population`. We need to **separate** it.

### Values That Should Be Column Names Stuck in Rows

Sometimes what should be separate columns is listed as categories in a single column.

```{r}
messy_both <- tribble(
  ~country,     ~year, ~type,    ~count,
  "Afghanistan", 2000, "cases",       745,
  "Afghanistan", 2000, "population",  20000000,
  "Afghanistan", 2001, "cases",       2666,
  "Afghanistan", 2001, "population",  20595360,
  "Brazil",      2000, "cases",       37737,
  "Brazil",      2000, "population",  172006362
)

messy_both
```

The `type` column contains what should be variable names. We need to **pivot wider** to give `cases` and `population` their own columns.

## pivot_longer() -- Fixing Wide Spreadsheets

This is the function you will use the most. When your boss gives you a spreadsheet with months as columns, quarters as columns, or product lines as columns -- `pivot_longer()` is your fix.

### Basic Usage

Let's take that wide dataset and make it tidy:

```{r}
messy_wide
```

```{r}
messy_wide %>%
  pivot_longer(
    cols = `2000`:`2002`,
    names_to = "year",
    values_to = "cases"
  )
```

Three arguments to remember:

- **`cols`**: Which columns to pivot (the ones that are really values, not variables).
- **`names_to`**: What to call the new column that holds the old column names.
- **`values_to`**: What to call the new column that holds the values.

Notice that `year` came in as character text. You can fix that automatically:

```{r}
messy_wide %>%
  pivot_longer(
    cols = `2000`:`2002`,
    names_to = "year",
    values_to = "cases",
    names_transform = list(year = as.integer)
  )
```

### Selecting Which Columns to Pivot

You have several ways to tell `pivot_longer()` which columns to grab. Here is a dataset to demonstrate:

```{r}
student_scores <- tribble(
  ~student, ~math_midterm, ~math_final, ~english_midterm, ~english_final,
  "Alice",            88,          92,               78,             85,
  "Bob",              76,          81,               90,             88,
  "Carol",            95,          97,               82,             91
)

student_scores
```

**Pivot everything except the ID column** (this is the most common pattern):

```{r}
student_scores %>%
  pivot_longer(
    cols = -student,
    names_to = "exam",
    values_to = "score"
  )
```

**Use `starts_with()` to grab columns by prefix:**

```{r}
student_scores %>%
  pivot_longer(
    cols = starts_with("math"),
    names_to = "math_exam",
    values_to = "math_score"
  )
```

### A Business Example: Quarterly Revenue

Here is a scenario you will definitely encounter. Quarterly revenue in wide format:

```{r}
quarterly_revenue <- tribble(
  ~company, ~Q1_revenue, ~Q2_revenue, ~Q3_revenue, ~Q4_revenue,
  "Acme",          100,         120,         130,         150,
  "Beta",          200,         210,         250,         270
)

quarterly_revenue %>%
  pivot_longer(
    cols = -company,
    names_to = "quarter",
    values_to = "revenue",
    names_pattern = "Q(\\d)_revenue"
  )
```

The `names_pattern` uses a regular expression to extract just the quarter number. Clean and ready for plotting.

### Dropping NAs While Pivoting

Got sparse data with lots of blanks? Use `values_drop_na = TRUE`:

```{r}
sparse_data <- tribble(
  ~id, ~week1, ~week2, ~week3,
  "A",     10,     NA,     30,
  "B",     NA,     25,     35,
  "C",     15,     20,     NA
)

sparse_data %>%
  pivot_longer(
    cols = -id,
    names_to = "week",
    values_to = "measurement",
    values_drop_na = TRUE
  )
```

### Pivoting Column Names with Multiple Pieces

Sometimes column names encode more than one thing. Our student scores have both subject and exam type baked in:

```{r}
student_scores %>%
  pivot_longer(
    cols = -student,
    names_to = c("subject", "exam"),
    names_sep = "_",
    values_to = "score"
  )
```

The `names_sep = "_"` splits each column name at the underscore and puts the pieces into two new columns. Pretty slick.

## pivot_wider() -- Going the Other Way

Sometimes you need to spread data out, not collapse it. Maybe you are building a summary table for a presentation, or a function expects a matrix-like format. That is where `pivot_wider()` comes in.

### Basic Usage

Remember our messy dataset where variable names were stuck in a column?

```{r}
messy_both
```

```{r}
messy_both %>%
  pivot_wider(
    names_from = type,
    values_from = count
  )
```

Two arguments:

- **`names_from`**: The column whose values become the new column names.
- **`values_from`**: The column whose values fill those new columns.

### Filling in Missing Combinations

When some combinations don't exist, you will get NAs. Use `values_fill` to replace them:

```{r}
weekly_sales <- tribble(
  ~store,   ~product,  ~sales,
  "North",  "Apples",     100,
  "North",  "Bananas",    150,
  "South",  "Apples",     200,
  "South",  "Cherries",    80,
  "East",   "Bananas",    120
)

weekly_sales %>%
  pivot_wider(
    names_from = product,
    values_from = sales,
    values_fill = 0
  )
```

Without `values_fill = 0`, the East store would show `NA` for Apples and Cherries. With it, you get zeros -- much better for a sales report.

### When to Use pivot_wider()

You will reach for this when you need to:

- **Build summary tables** for a slide deck or report
- **Create cross-tabulations** (e.g., survey responses by question)
- **Match a required input format** for a specific function

Here is a survey data example:

```{r}
survey <- tribble(
  ~respondent, ~question,     ~answer,
  "R1",        "satisfaction", "High",
  "R1",        "recommend",    "Yes",
  "R2",        "satisfaction", "Medium",
  "R2",        "recommend",    "No",
  "R3",        "satisfaction", "High",
  "R3",        "recommend",    "Yes"
)

survey %>%
  pivot_wider(
    names_from = question,
    values_from = answer
  )
```

Now each respondent is one row with their answers spread across columns. Way easier to work with.

## separate() and unite() -- Quick Splitting and Combining

### Splitting a Column Apart

When a column has two things mashed together, `separate()` splits them:

```{r}
locations <- tribble(
  ~name,          ~city_state,
  "Alice",        "Portland_Oregon",
  "Bob",          "Austin_Texas",
  "Carol",        "Denver_Colorado",
  "David",        "Miami_Florida"
)

locations %>%
  separate(city_state, into = c("city", "state"), sep = "_")
```

Works with dates too:

```{r}
dates_data <- tribble(
  ~event,     ~date,
  "Meeting",  "2024-01-15",
  "Workshop", "2024-03-22",
  "Summit",   "2024-07-10"
)

dates_data %>%
  separate(date, into = c("year", "month", "day"), sep = "-", convert = TRUE)
```

The `convert = TRUE` automatically turns the resulting columns into numbers instead of leaving them as text.

The newer `separate_wider_delim()` does the same thing with slightly more explicit syntax:

```{r}
locations %>%
  separate_wider_delim(city_state, delim = "_", names = c("city", "state"))
```

### Combining Columns with unite()

`unite()` does the reverse -- it glues columns together:

```{r}
date_parts <- tribble(
  ~year, ~month, ~day,
  2024,      1,   15,
  2024,      3,   22,
  2024,      7,   10
)

date_parts %>%
  unite(col = "date", year, month, day, sep = "-")
```

Handy for building full addresses, combined keys, or file names from parts.

## Handling Missing Values

Missing data is a fact of life in business. Customer didn't fill out a field? Sensor went offline? Store didn't report that week? Here are your tools.

### drop_na() -- Remove Rows with Missing Values

```{r}
customer_data <- tribble(
  ~customer, ~revenue, ~satisfaction, ~nps_score,
  "C1",          1200,            8,        9.0,
  "C2",            NA,            7,        8.5,
  "C3",          1500,           NA,        7.2,
  "C4",           980,            9,         NA,
  "C5",          1100,            8,        8.8
)

# Drop rows missing ANY value
customer_data %>%
  drop_na()
```

```{r}
# Drop only rows missing revenue
customer_data %>%
  drop_na(revenue)
```

### fill() -- Carry Values Down (or Up)

This is a lifesaver for spreadsheets where someone only typed the year once at the top of each group:

```{r}
quarterly_reports <- tribble(
  ~year, ~quarter, ~revenue,
  2023,        1,    500,
  NA,          2,    520,
  NA,          3,    530,
  NA,          4,    610,
  2024,        1,    580,
  NA,          2,    600,
  NA,          3,    650,
  NA,          4,    700
)

quarterly_reports %>%
  fill(year)
```

By default it fills downward. You can also fill upward with `.direction = "up"`.

### replace_na() -- Swap NAs for Specific Values

```{r}
customer_data %>%
  replace_na(list(revenue = 0, satisfaction = 0, nps_score = 0))
```

### complete() -- Make Hidden Missing Data Visible

Sometimes data is missing entire rows, not just individual cells. `complete()` generates all combinations and makes the gaps obvious:

```{r}
experiment <- tribble(
  ~subject, ~treatment, ~outcome,
  "S1",     "A",            5.2,
  "S1",     "B",            6.1,
  "S2",     "A",            4.8,
  "S3",     "B",            7.0,
  "S3",     "A",            5.5
)

experiment %>%
  complete(subject, treatment, fill = list(outcome = 0))
```

Subject S2 with treatment B was completely missing from the original data. Now you can see it.

## Nesting and Unnesting

Nesting lets you collapse groups of rows into sub-data-frames stored inside a single column. It is an advanced technique primarily used with the **purrr** package for fitting models or running functions group-by-group. If that sounds intense, don't worry -- you can skip this for now and come back later when you need it.

## Putting It All Together

Let's walk through a realistic example. You get a spreadsheet of student test scores from a school district. It looks like something a well-meaning administrator exported from Excel:

```{r}
messy_scores <- tribble(
  ~school,      ~`math_2022_male`, ~`math_2022_female`, ~`math_2023_male`, ~`math_2023_female`, ~`reading_2022_male`, ~`reading_2022_female`, ~`reading_2023_male`, ~`reading_2023_female`,
  "Lincoln",                   78,                  82,                80,                  85,                   72,                    80,                   75,                    83,
  "Washington",                85,                  88,                87,                  90,                   80,                    85,                   82,                    87,
  "Jefferson",                 70,                  75,                73,                  78,                   68,                    72,                   71,                    76
)

messy_scores
```

Eight columns of scores with subject, year, and gender all crammed into the column names. Yikes.

One pipeline to tidy it all:

```{r}
tidy_scores <- messy_scores %>%
  pivot_longer(
    cols = -school,
    names_to = c("subject", "year", "gender"),
    names_sep = "_",
    values_to = "avg_score"
  ) %>%
  mutate(year = as.integer(year)) %>%
  arrange(school, subject, year, gender)

tidy_scores
```

Now we can actually *do* things with this data. Average score by subject and year?

```{r}
tidy_scores %>%
  group_by(subject, year) %>%
  summarize(
    mean_score = mean(avg_score),
    .groups = "drop"
  )
```

A quick visualization?

```{r}
tidy_scores %>%
  group_by(subject, year, gender) %>%
  summarize(mean_score = mean(avg_score), .groups = "drop") %>%
  ggplot(aes(x = year, y = mean_score, color = gender)) +
  geom_line() +
  geom_point(size = 2) +
  facet_wrap(~subject) +
  labs(
    title = "Average Test Scores by Subject, Year, and Gender",
    x = "Year",
    y = "Average Score",
    color = "Gender"
  ) +
  theme_minimal()
```

That whole analysis became possible because we invested a few lines in tidying the data first. The messier the original spreadsheet, the more valuable this step becomes.

## Summary

Here is your tidyr cheat sheet:

| Function | What It Does |
|:---------|:--------|
| `pivot_longer()` | Wide to long (the one you will use most) |
| `pivot_wider()` | Long to wide (for summary tables and reports) |
| `separate()` / `separate_wider_delim()` | Split one column into multiple columns |
| `unite()` | Glue multiple columns into one |
| `drop_na()` | Remove rows with missing values |
| `fill()` | Fill missing values with the previous (or next) value |
| `replace_na()` | Replace NAs with a value you choose |
| `complete()` | Make hidden missing rows visible |

The bottom line: getting your data into tidy form -- where each variable is a column, each observation is a row, and each value is a cell -- is the single best investment you can make before any analysis. Tidy first, analyze second. Your future self will thank you.
