# Data Visualization with ggplot2 {#ggplot2basics}

```{r include=FALSE}
library(tidyverse)
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

> **Note:** This chapter provides the R foundations for ggplot2. For deeper coverage of visualization design principles, chart type selection, and visual perception, refer to the [Visual Analytics and Data Storytelling](http://patilv.com/VizAnalytics4Biz/) course.

ggplot2 is the tidyverse's data visualization package. It is one of the most popular R packages ever created and has fundamentally changed how people think about building statistical graphics. In this chapter, we will learn the grammar that powers ggplot2 and build a wide range of visualizations from the ground up.

## The Grammar of Graphics

ggplot2 is based on *The Grammar of Graphics*, a landmark book by Leland Wilkinson published in 1999. Wilkinson's key insight was that statistical graphics, much like natural language, can be decomposed into a set of fundamental components governed by rules of composition. Rather than thinking of charts as fixed types --- "this is a bar chart," "that is a scatterplot" --- the grammar encourages you to think of every visualization as a combination of building blocks.

Hadley Wickham, the creator of ggplot2, adapted Wilkinson's grammar into a **layered grammar of graphics** with seven key components:

1. **Data**: The dataset being visualized.
2. **Aesthetics** (`aes`): Mappings from variables in the data to visual properties (position, color, size, shape, etc.).
3. **Geometric objects** (`geom`): The visual marks that represent data points (points, lines, bars, etc.).
4. **Facets**: Subplots that split the data by one or more categorical variables.
5. **Statistical transformations** (`stat`): Summaries or computations performed on the data before plotting (counts, means, smoothing lines, etc.).
6. **Coordinate systems** (`coord`): The system that maps the position aesthetics to the 2D plane of the plot (Cartesian, polar, flipped, etc.).
7. **Themes**: Non-data visual elements that control the overall appearance (fonts, background, gridlines, etc.).

Every ggplot2 plot is built by specifying some or all of these layers. The beauty of this approach is its composability --- you can mix and match layers to create virtually any chart you can imagine.

The template for building a ggplot can be expressed as:

```
ggplot(data = <DATA>, mapping = aes(<MAPPINGS>)) +
  <GEOM_FUNCTION>(
    stat = <STAT>,
    position = <POSITION>
  ) +
  <COORDINATE_FUNCTION> +
  <FACET_FUNCTION> +
  <THEME>
```

You rarely need to specify every component explicitly. ggplot2 provides sensible defaults for most layers, so in practice you build plots incrementally, adding only the layers you need.


## Your First ggplot

Let us build our first visualization step by step using the `mpg` dataset, which comes bundled with ggplot2. This dataset contains fuel economy data for 38 popular car models from 1999 to 2008.

```{r}
glimpse(mpg)
```

### Step 1: Initialize the plot

The `ggplot()` function creates a coordinate system that you can add layers to. By itself, it produces a blank canvas:

```{r}
ggplot(data = mpg)
```

### Step 2: Map variables to aesthetics

Next, we tell ggplot2 which variables should be mapped to the x-axis and y-axis using the `aes()` function. This still does not draw anything visible because we have not specified *how* to represent the data:

```{r}
ggplot(data = mpg, mapping = aes(x = displ, y = hwy))
```

Notice that ggplot2 has now set up the axis scales based on the data, but there are no points or lines yet.

### Step 3: Add a geometric layer

Finally, we add a `geom` layer to specify *how* the data should be displayed. For a scatterplot, we use `geom_point()`:

```{r}
ggplot(data = mpg, mapping = aes(x = displ, y = hwy)) +
  geom_point()
```

This plot reveals a clear negative relationship: as engine displacement (`displ`) increases, highway fuel economy (`hwy`) tends to decrease. Notice also the cluster of outliers in the upper right --- these turn out to be two-seater sports cars with efficient highway gearing.

The `+` operator is essential in ggplot2. It adds layers to the plot. You always start with `ggplot()` and add components using `+`.

> **Tip:** A common mistake is to put the `+` at the beginning of a new line instead of at the end of the previous line. The `+` must come at the end of the line that precedes the next layer.


## Aesthetic Mappings

Aesthetics are the visual properties of the objects in your plot. The `aes()` function maps variables in your data to these visual properties. Common aesthetics include:

| Aesthetic  | Description                       |
|------------|-----------------------------------|
| `x`        | Position on the x-axis            |
| `y`        | Position on the y-axis            |
| `color`    | Color of points, lines, outlines  |
| `fill`     | Fill color of areas               |
| `size`     | Size of points or thickness of lines |
| `shape`    | Shape of points                   |
| `alpha`    | Transparency (0 = invisible, 1 = opaque) |
| `linetype` | Type of line (solid, dashed, etc.) |
| `group`    | Grouping variable for connecting observations |

### Mapping vs. Setting

There is a critical distinction between **mapping** an aesthetic (placing it inside `aes()`) and **setting** it (placing it outside `aes()`).

**Mapping** connects a variable in the data to a visual property. ggplot2 automatically creates a scale and a legend:

```{r}
ggplot(data = mpg, aes(x = displ, y = hwy, color = class)) +
  geom_point()
```

Here, each car class is assigned a distinct color, and a legend is generated automatically.

**Setting** assigns a fixed visual property to all points. The value is not connected to any variable:

```{r}
ggplot(data = mpg, aes(x = displ, y = hwy)) +
  geom_point(color = "steelblue", size = 3, alpha = 0.6)
```

All points are now steel blue, size 3, and 60% opaque, regardless of any variable in the data.

### Examples of different aesthetics

**Shape aesthetic:**

```{r}
ggplot(data = mpg, aes(x = displ, y = hwy, shape = drv)) +
  geom_point(size = 2)
```

**Size aesthetic:**

```{r}
ggplot(data = mpg, aes(x = displ, y = hwy, size = cyl)) +
  geom_point(alpha = 0.5)
```

**Alpha (transparency) aesthetic:**

```{r}
ggplot(data = mpg, aes(x = displ, y = hwy, alpha = cyl)) +
  geom_point(size = 3, color = "darkblue")
```

**Multiple aesthetics combined:**

```{r}
ggplot(data = mpg, aes(x = displ, y = hwy, color = class, shape = drv)) +
  geom_point(size = 2.5)
```

When you map a continuous variable to `color`, ggplot2 uses a gradient scale instead of discrete colors:

```{r}
ggplot(data = mpg, aes(x = displ, y = hwy, color = cty)) +
  geom_point(size = 2)
```


## Geometric Objects (Geoms)

Geoms are the visual building blocks of your plot. Each geom function creates a different type of visual representation. You can layer multiple geoms on the same plot.

### One Variable: Continuous

**Histogram** --- shows the distribution of a single continuous variable by dividing it into bins:

```{r}
ggplot(data = mpg, aes(x = hwy)) +
  geom_histogram(binwidth = 2, fill = "steelblue", color = "white")
```

**Density plot** --- a smoothed version of the histogram:

```{r}
ggplot(data = mpg, aes(x = hwy)) +
  geom_density(fill = "steelblue", alpha = 0.5)
```

**Frequency polygon** --- similar to a histogram but uses lines instead of bars, making it easier to compare distributions:

```{r}
ggplot(data = mpg, aes(x = hwy, color = drv)) +
  geom_freqpoly(binwidth = 2, linewidth = 1)
```

**Dotplot** --- each observation is a dot, stacked along the axis:

```{r}
ggplot(data = mpg, aes(x = hwy)) +
  geom_dotplot(binwidth = 1, fill = "steelblue")
```

### One Variable: Discrete

**Bar chart** --- counts the number of observations in each category:

```{r}
ggplot(data = mpg, aes(x = class)) +
  geom_bar(fill = "steelblue")
```

Note that `geom_bar()` computes the count automatically via `stat_count()`. You provide only the x variable.

### Distribution by Group

**Boxplot** --- displays the five-number summary (minimum, Q1, median, Q3, maximum) and outliers:

```{r}
ggplot(data = mpg, aes(x = class, y = hwy)) +
  geom_boxplot(fill = "lightblue", color = "navy")
```

**Violin plot** --- similar to a boxplot but shows the full distribution shape:

```{r}
ggplot(data = mpg, aes(x = class, y = hwy)) +
  geom_violin(fill = "lightblue", color = "navy", alpha = 0.7)
```

You can layer a boxplot on top of a violin plot for added detail:

```{r}
ggplot(data = mpg, aes(x = class, y = hwy)) +
  geom_violin(fill = "lightblue", alpha = 0.5) +
  geom_boxplot(width = 0.2, fill = "white")
```

### Two Variables: Continuous x and Continuous y

**Scatterplot** with `geom_point()`:

```{r}
ggplot(data = mpg, aes(x = displ, y = hwy)) +
  geom_point()
```

**Jittered scatterplot** --- adds small random noise to reveal overlapping points:

```{r}
ggplot(data = mpg, aes(x = displ, y = hwy)) +
  geom_jitter(width = 0.2, height = 0.5, alpha = 0.6)
```

**Line plot** --- connects observations in order of the x variable. Especially useful for time series:

```{r}
ggplot(data = economics, aes(x = date, y = unemploy)) +
  geom_line(color = "steelblue")
```

**Smooth line** --- fits a smoothed conditional mean:

```{r}
ggplot(data = mpg, aes(x = displ, y = hwy)) +
  geom_point(alpha = 0.4) +
  geom_smooth()
```

By default, `geom_smooth()` uses LOESS for small datasets and GAM for large ones. You can specify the method explicitly:

```{r}
ggplot(data = mpg, aes(x = displ, y = hwy)) +
  geom_point(alpha = 0.4) +
  geom_smooth(method = "lm", se = TRUE, color = "red")
```

### Two Variables: Discrete x and Continuous y

**Column chart** with `geom_col()` --- unlike `geom_bar()`, this requires both x and y aesthetics. Use it when you have pre-computed summary values:

```{r}
mpg_avg <- mpg %>%
  group_by(class) %>%
  summarize(avg_hwy = mean(hwy))

ggplot(data = mpg_avg, aes(x = class, y = avg_hwy)) +
  geom_col(fill = "steelblue")
```

### Annotations

**Text labels** --- add text at each data point:

```{r}
ggplot(data = mpg_avg, aes(x = class, y = avg_hwy)) +
  geom_col(fill = "steelblue") +
  geom_text(aes(label = round(avg_hwy, 1)), vjust = -0.5, size = 3.5)
```

**Labels** --- similar to text but with a background rectangle for readability:

```{r}
best_cars <- mpg %>%
  filter(hwy > 40)

ggplot(data = mpg, aes(x = displ, y = hwy)) +
  geom_point(alpha = 0.4) +
  geom_label(data = best_cars, aes(label = model), size = 3, nudge_y = 1)
```

**Reference lines** --- horizontal, vertical, and diagonal lines:

```{r}
ggplot(data = mpg, aes(x = displ, y = hwy)) +
  geom_point(alpha = 0.5) +
  geom_hline(yintercept = mean(mpg$hwy), linetype = "dashed", color = "red") +
  geom_vline(xintercept = mean(mpg$displ), linetype = "dashed", color = "blue")
```

The `geom_abline()` function draws lines defined by a slope and intercept:

```{r}
ggplot(data = mpg, aes(x = displ, y = hwy)) +
  geom_point(alpha = 0.5) +
  geom_abline(intercept = 35, slope = -3.5, color = "red", linetype = "dashed")
```

### Layering Multiple Geoms

One of the most powerful aspects of ggplot2 is the ability to layer multiple geoms. When you specify aesthetics in the `ggplot()` call, they are inherited by all subsequent geoms. You can also set aesthetics locally within individual geoms:

```{r}
ggplot(data = mpg, aes(x = displ, y = hwy)) +
  geom_point(aes(color = class)) +
  geom_smooth(color = "black", se = FALSE) +
  geom_hline(yintercept = mean(mpg$hwy), linetype = "dotted", color = "gray50")
```


## Faceting

Faceting splits your data into subsets and creates a panel (subplot) for each subset. This is one of the most effective ways to display how patterns change across categories.

### facet_wrap()

`facet_wrap()` wraps a one-dimensional sequence of panels into a 2D layout. It takes a formula with a single variable:

```{r}
ggplot(data = mpg, aes(x = displ, y = hwy)) +
  geom_point() +
  facet_wrap(~ class)
```

You can control the layout with `ncol` or `nrow`:

```{r}
ggplot(data = mpg, aes(x = displ, y = hwy)) +
  geom_point() +
  facet_wrap(~ class, ncol = 4)
```

### facet_grid()

`facet_grid()` creates a matrix of panels defined by row and column variables:

```{r}
ggplot(data = mpg, aes(x = displ, y = hwy)) +
  geom_point() +
  facet_grid(drv ~ cyl)
```

If you want to facet by only rows or only columns, use a `.` on the other side of the formula:

```{r}
ggplot(data = mpg, aes(x = displ, y = hwy)) +
  geom_point() +
  facet_grid(drv ~ .)
```

### Free scales

By default, all panels share the same x and y axis scales. You can allow scales to vary across panels using the `scales` argument:

```{r}
ggplot(data = mpg, aes(x = hwy)) +
  geom_histogram(binwidth = 2, fill = "steelblue", color = "white") +
  facet_wrap(~ class, scales = "free_y")
```

Options include `"free"` (both axes free), `"free_x"` (only x-axis free), and `"free_y"` (only y-axis free).

### Faceting with iris

```{r}
ggplot(data = iris, aes(x = Sepal.Length, y = Sepal.Width)) +
  geom_point(aes(color = Species)) +
  geom_smooth(method = "lm", se = FALSE) +
  facet_wrap(~ Species, scales = "free")
```

The relationship between sepal length and sepal width differs across species. Without faceting, these patterns would be tangled together.


## Statistical Transformations

Every geom has a default statistical transformation, and every stat has a default geom. Most of the time you do not need to think about stats because the defaults do what you want. However, understanding stats gives you more control.

For example, `geom_bar()` uses `stat_count()` by default, which counts the number of observations in each group:

```{r}
ggplot(data = diamonds, aes(x = cut)) +
  geom_bar()
```

This is equivalent to:

```{r}
ggplot(data = diamonds, aes(x = cut)) +
  stat_count()
```

If you already have a table of counts, you can override the default stat and use `stat = "identity"`:

```{r}
diamond_counts <- diamonds %>% count(cut)

ggplot(data = diamond_counts, aes(x = cut, y = n)) +
  geom_bar(stat = "identity", fill = "steelblue")
```

### stat_summary()

`stat_summary()` allows you to compute and display summary statistics directly in the plot:

```{r}
ggplot(data = diamonds, aes(x = cut, y = price)) +
  stat_summary(
    fun = median,
    fun.min = function(x) quantile(x, 0.25),
    fun.max = function(x) quantile(x, 0.75),
    geom = "pointrange",
    color = "steelblue"
  )
```

### stat_smooth()

`stat_smooth()` computes a smoothed conditional mean. It is the stat behind `geom_smooth()`:

```{r}
ggplot(data = mpg, aes(x = displ, y = hwy)) +
  geom_point(alpha = 0.3) +
  stat_smooth(method = "loess", color = "red")
```


## Position Adjustments

Position adjustments determine how overlapping objects are arranged. They are most commonly used with bar charts, but apply to other geoms as well.

### Stacked bars (default)

The default position for `geom_bar()` is `"stack"`, which places bars on top of each other:

```{r}
ggplot(data = diamonds, aes(x = cut, fill = clarity)) +
  geom_bar(position = "stack")
```

### Dodged bars

`position = "dodge"` places bars side by side:

```{r}
ggplot(data = diamonds, aes(x = cut, fill = clarity)) +
  geom_bar(position = "dodge")
```

### Filled bars

`position = "fill"` stacks bars and standardizes each bar to the same height, showing proportions:

```{r}
ggplot(data = diamonds, aes(x = cut, fill = clarity)) +
  geom_bar(position = "fill") +
  ylab("Proportion")
```

### Identity

`position = "identity"` places objects exactly where they fall. This is most useful for point geoms but can lead to overlapping bars:

```{r}
ggplot(data = diamonds, aes(x = cut, fill = clarity)) +
  geom_bar(position = "identity", alpha = 0.3)
```

### Jitter

`position = "jitter"` adds small random displacement. This is especially useful for scatterplots with overplotting:

```{r}
ggplot(data = mpg, aes(x = drv, y = hwy)) +
  geom_point(position = "jitter")
```

Here is a comparison to illustrate the difference. Without jitter, many points overlap and the true density is hidden:

```{r}
p1 <- ggplot(data = mpg, aes(x = drv, y = hwy)) +
  geom_point() +
  ggtitle("Without jitter")

p2 <- ggplot(data = mpg, aes(x = drv, y = hwy)) +
  geom_point(position = position_jitter(width = 0.2)) +
  ggtitle("With jitter")

library(patchwork)
p1 + p2
```


## Coordinate Systems

Coordinate systems map the position aesthetics (x and y) onto the plane of the plot. The default is `coord_cartesian()`, which is the standard Cartesian coordinate system.

### coord_flip()

`coord_flip()` swaps the x and y axes. This is useful for making horizontal bar charts or when category labels are long:

```{r}
ggplot(data = mpg, aes(x = class)) +
  geom_bar(fill = "steelblue") +
  coord_flip()
```

### coord_polar()

`coord_polar()` maps x and y positions to angle and radius in a polar coordinate system. A common use is creating pie or donut charts (though they are often discouraged from a perception standpoint):

```{r}
ggplot(data = diamonds, aes(x = "", fill = cut)) +
  geom_bar(width = 1) +
  coord_polar(theta = "y") +
  theme_void()
```

### coord_fixed()

`coord_fixed()` ensures that one unit on the x-axis equals one unit on the y-axis:

```{r}
ggplot(data = mpg, aes(x = cty, y = hwy)) +
  geom_point() +
  geom_abline(color = "red") +
  coord_fixed()
```

This is useful when the relationship between x and y should be interpreted on equal scales (such as comparing city vs. highway MPG).

### coord_cartesian() for zooming

You can zoom into a plot without dropping data using `coord_cartesian()`. This is different from setting axis limits with `xlim()`/`ylim()`, which removes data outside the range:

```{r}
ggplot(data = mpg, aes(x = displ, y = hwy)) +
  geom_point() +
  geom_smooth() +
  coord_cartesian(xlim = c(3, 6), ylim = c(15, 35))
```


## Labels and Annotations

Clear, informative labels are crucial for making your plots understandable. ggplot2 provides several functions for adding and customizing labels.

### labs()

The `labs()` function is the most flexible way to add labels. It can set the title, subtitle, caption, and axis labels all at once:

```{r}
ggplot(data = mpg, aes(x = displ, y = hwy, color = class)) +
  geom_point() +
  labs(
    title = "Fuel Efficiency vs. Engine Displacement",
    subtitle = "Data from EPA for 38 popular car models (1999-2008)",
    caption = "Source: mpg dataset in ggplot2",
    x = "Engine Displacement (liters)",
    y = "Highway Fuel Economy (mpg)",
    color = "Vehicle Class"
  )
```

### Individual label functions

You can also use individual functions for specific labels:

```{r}
ggplot(data = mpg, aes(x = displ, y = hwy)) +
  geom_point(color = "steelblue") +
  ggtitle("Fuel Efficiency vs. Engine Displacement") +
  xlab("Engine Displacement (liters)") +
  ylab("Highway Fuel Economy (mpg)")
```

### annotate()

The `annotate()` function adds text, rectangles, segments, or other annotations at specific positions on the plot. Unlike `geom_text()`, it does not map to data --- it places elements at fixed coordinates:

```{r}
ggplot(data = mpg, aes(x = displ, y = hwy)) +
  geom_point(alpha = 0.5) +
  annotate(
    "text",
    x = 6, y = 40,
    label = "High efficiency\nlarge engines",
    color = "red", fontface = "italic", size = 4
  ) +
  annotate(
    "rect",
    xmin = 5, xmax = 7, ymin = 35, ymax = 45,
    alpha = 0.1, fill = "red", color = "red", linetype = "dashed"
  )
```

You can annotate with segments and arrows as well:

```{r}
ggplot(data = mpg, aes(x = displ, y = hwy)) +
  geom_point(alpha = 0.5) +
  annotate(
    "segment",
    x = 5.5, xend = 5, y = 42, yend = 44,
    arrow = arrow(length = unit(0.2, "cm")),
    color = "red"
  ) +
  annotate(
    "text",
    x = 5.5, y = 41,
    label = "Corvette",
    color = "red", size = 3.5
  )
```


## Saving Plots

The `ggsave()` function saves the most recently created plot to a file. It automatically detects the format from the file extension.

```{r eval=FALSE}
# Save the last plot as a PNG
ggsave("my_plot.png")

# Specify dimensions and resolution
ggsave("my_plot.png", width = 8, height = 6, dpi = 300)

# Save as PDF (vector format, ideal for publications)
ggsave("my_plot.pdf", width = 8, height = 6)

# Save as SVG (vector format, ideal for web)
ggsave("my_plot.svg", width = 8, height = 6)

# Save a specific plot object
my_plot <- ggplot(mpg, aes(displ, hwy)) + geom_point()
ggsave("specific_plot.png", plot = my_plot, width = 10, height = 7, dpi = 150)
```

Key arguments for `ggsave()`:

| Argument  | Description                                          |
|-----------|------------------------------------------------------|
| `filename`| Name of the file to create (path and extension)      |
| `plot`    | The plot object to save (defaults to the last plot)   |
| `width`   | Width in inches (or `units` if specified)             |
| `height`  | Height in inches (or `units` if specified)            |
| `dpi`     | Resolution in dots per inch (default is 300)          |
| `device`  | Device to use (auto-detected from extension)          |
| `units`   | Units for width/height: `"in"`, `"cm"`, or `"mm"`    |


## Putting It All Together

Let us build a polished, publication-ready visualization from scratch, applying the concepts from this chapter. We will use the `diamonds` dataset to explore the relationship between carat weight and price across different diamond cuts.

### Step 1: Start with the basic scatterplot

```{r}
ggplot(data = diamonds, aes(x = carat, y = price)) +
  geom_point()
```

There are over 50,000 observations, so overplotting is severe. Let us improve this.

### Step 2: Address overplotting with transparency and size

```{r}
ggplot(data = diamonds, aes(x = carat, y = price)) +
  geom_point(alpha = 0.05, size = 0.5)
```

### Step 3: Map color to cut quality

```{r}
ggplot(data = diamonds, aes(x = carat, y = price, color = cut)) +
  geom_point(alpha = 0.1, size = 0.5)
```

### Step 4: Add smoothing lines for each cut

```{r}
ggplot(data = diamonds, aes(x = carat, y = price, color = cut)) +
  geom_point(alpha = 0.05, size = 0.3) +
  geom_smooth(se = FALSE, linewidth = 1)
```

### Step 5: Facet by cut for clarity

```{r}
ggplot(data = diamonds, aes(x = carat, y = price)) +
  geom_point(alpha = 0.05, size = 0.3, color = "steelblue") +
  geom_smooth(color = "red", se = FALSE) +
  facet_wrap(~ cut)
```

### Step 6: Add labels and refine the theme

```{r}
ggplot(data = diamonds, aes(x = carat, y = price)) +
  geom_point(alpha = 0.05, size = 0.3, color = "steelblue") +
  geom_smooth(color = "red", se = FALSE) +
  facet_wrap(~ cut) +
  scale_y_continuous(labels = scales::dollar_format()) +
  labs(
    title = "Diamond Price by Carat Weight and Cut Quality",
    subtitle = "Each panel shows one of five cut grades; red line is a LOESS smooth",
    caption = "Source: diamonds dataset in ggplot2",
    x = "Carat Weight",
    y = "Price (USD)"
  ) +
  theme_minimal(base_size = 12) +
  theme(
    plot.title = element_text(face = "bold"),
    strip.text = element_text(face = "bold"),
    panel.grid.minor = element_blank()
  )
```

This final plot demonstrates the full grammar of graphics at work:

- **Data**: the `diamonds` dataset
- **Aesthetics**: carat mapped to x, price mapped to y
- **Geoms**: points and smooth lines layered together
- **Facets**: one panel per cut quality
- **Stats**: LOESS smoothing computed by `geom_smooth()`
- **Coordinates**: default Cartesian (appropriate here)
- **Theme**: `theme_minimal()` with customizations for a clean, publication-ready look

This layered approach is the hallmark of ggplot2. By building incrementally, you can experiment at each step and diagnose issues easily.

### Another complete example: time series with economics

```{r}
ggplot(data = economics, aes(x = date, y = unemploy / 1000)) +
  geom_line(color = "steelblue", linewidth = 0.8) +
  geom_smooth(method = "loess", span = 0.1, se = FALSE, color = "red", linewidth = 0.5) +
  scale_x_date(date_labels = "%Y", date_breaks = "5 years") +
  labs(
    title = "U.S. Unemployment Over Time",
    subtitle = "Monthly number of unemployed persons (in thousands), 1967-2015",
    caption = "Source: economics dataset in ggplot2",
    x = NULL,
    y = "Unemployed (thousands)"
  ) +
  theme_minimal(base_size = 12) +
  theme(
    plot.title = element_text(face = "bold"),
    axis.text.x = element_text(angle = 45, hjust = 1),
    panel.grid.minor = element_blank()
  )
```

In the next chapter, we will go deeper into customizing ggplot2 visualizations with scales, themes, color palettes, and more advanced techniques.
