# Data Visualization with ggplot2 {#ggplot2basics}

```{r include=FALSE}
library(tidyverse)
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

You are about to learn the most popular data visualization tool in the R world. **ggplot2** is the package that turns your data into charts that actually look good -- the kind of charts that make people in meetings stop scrolling through their phones and pay attention.

Whether you are building a quick scatter plot to explore sales data or a polished visualization for a board presentation, ggplot2 is the tool. It is used by data analysts at every major company, by journalists at The New York Times and FiveThirtyEight, and by researchers across every field. By the end of this chapter, you will know how to use it too.

> **Note:** This chapter teaches you *how* to build charts in R. For the *why* behind choosing the right chart type -- when to use a bar chart vs. a line chart, how to design for your audience, visual perception principles -- check the [Visual Analytics and Data Storytelling](http://patilv.com/VizAnalytics4Biz/) course.

## The Grammar of Graphics (the 30-Second Version)

Here is the big idea behind ggplot2: every chart is built from the same three ingredients.

1. **Data** -- the dataset you want to visualize
2. **Aesthetics** (`aes`) -- how your variables map to visual properties (which variable goes on the x-axis, which one controls the color, etc.)
3. **Geoms** -- the geometric shapes that represent your data (points, bars, lines, etc.)

That is it. Data + mapping + shape. You can add more layers on top (facets, labels, themes), but those three ingredients are the foundation of every single ggplot2 chart.

The general template looks like this:

```
ggplot(data = <DATA>, mapping = aes(<MAPPINGS>)) +
  <GEOM_FUNCTION>()
```

You start with `ggplot()`, tell it what data and mappings to use, then add a geom with `+`. Let's see it in action.


## Your First ggplot

We will use the `mpg` dataset that comes built into ggplot2. It has fuel economy data for popular car models -- think of it as a product performance report for the auto industry.

```{r}
glimpse(mpg)
```

### Step 1: Create the canvas

`ggplot()` sets up a blank canvas. By itself, it draws nothing:

```{r}
ggplot(data = mpg)
```

### Step 2: Map your variables

Now we tell ggplot2 which variables go where using `aes()`. Still no visible data yet -- just the axis setup:

```{r}
ggplot(data = mpg, mapping = aes(x = displ, y = hwy))
```

### Step 3: Add the geom

Finally, we tell ggplot2 *how* to draw the data. For a scatter plot, use `geom_point()`:

```{r}
ggplot(data = mpg, mapping = aes(x = displ, y = hwy)) +
  geom_point()
```

And just like that, you have a chart. It shows a clear pattern: bigger engines (`displ`) tend to get worse highway mileage (`hwy`). The outliers in the upper right? Those turn out to be two-seater sports cars with efficient highway gearing.

> **Common mistake:** Putting the `+` at the *beginning* of a new line instead of at the *end* of the previous line. The `+` must come at the end of the line before the next layer.


## Aesthetic Mappings -- Making Your Charts Informative

Aesthetics are the visual properties of your chart. The `aes()` function maps variables in your data to things the viewer can see:

| Aesthetic  | What It Controls                  |
|------------|-----------------------------------|
| `x`        | Position on the x-axis            |
| `y`        | Position on the y-axis            |
| `color`    | Color of points, lines, outlines  |
| `fill`     | Fill color of bars and areas      |
| `size`     | Size of points or line thickness  |
| `shape`    | Shape of points                   |
| `alpha`    | Transparency (0 = invisible, 1 = solid) |

### Mapping vs. Setting -- This Trips Everyone Up

There is one distinction that confuses almost every beginner. Pay attention to this.

**Mapping** (inside `aes()`) connects a *variable* to a visual property. ggplot2 picks colors automatically and creates a legend:

```{r}
ggplot(data = mpg, aes(x = displ, y = hwy, color = class)) +
  geom_point()
```

Each car class gets its own color. The legend appears automatically.

**Setting** (outside `aes()`) applies a *fixed* value to everything. No variable involved, no legend:

```{r}
ggplot(data = mpg, aes(x = displ, y = hwy)) +
  geom_point(color = "steelblue", size = 3, alpha = 0.6)
```

All points are steel blue, size 3, 60% opaque. If you accidentally put `color = "steelblue"` inside `aes()`, you will get weird results. Keep mapped aesthetics inside `aes()` and fixed values outside.

### Combining Multiple Aesthetics

You can map several variables at once for richer charts:

```{r}
ggplot(data = mpg, aes(x = displ, y = hwy, color = class, shape = drv)) +
  geom_point(size = 2.5)
```

And when you map a continuous variable to color, ggplot2 uses a gradient instead of discrete colors:

```{r}
ggplot(data = mpg, aes(x = displ, y = hwy, color = cty)) +
  geom_point(size = 2)
```


## Geoms -- The Chart Types That Matter

Geoms are the building blocks of your chart. Each geom draws a different shape. Here are the ones you will use 90% of the time in a business context.

### Bar Chart -- Counting Categories

Want to know how many observations fall in each category? `geom_bar()` counts for you automatically:

```{r}
ggplot(data = mpg, aes(x = class)) +
  geom_bar(fill = "steelblue")
```

If you already have pre-computed values (say, average revenue by region), use `geom_col()` instead:

```{r}
mpg_avg <- mpg %>%
  group_by(class) %>%
  summarize(avg_hwy = mean(hwy))

ggplot(data = mpg_avg, aes(x = class, y = avg_hwy)) +
  geom_col(fill = "steelblue") +
  geom_text(aes(label = round(avg_hwy, 1)), vjust = -0.5, size = 3.5)
```

Adding value labels on top of your bars is a nice touch for presentations.

### Histogram -- Distribution of a Continuous Variable

How are values spread out? Histograms show you:

```{r}
ggplot(data = mpg, aes(x = hwy)) +
  geom_histogram(binwidth = 2, fill = "steelblue", color = "white")
```

Play with `binwidth` to change how fine-grained the bins are. Too few bins hides patterns; too many creates noise.

### Scatter Plot -- Relationships Between Two Numbers

The classic. Great for spotting correlations, clusters, and outliers:

```{r}
ggplot(data = mpg, aes(x = displ, y = hwy)) +
  geom_point()
```

When points overlap (overplotting), use transparency or jitter:

```{r}
ggplot(data = mpg, aes(x = displ, y = hwy)) +
  geom_jitter(width = 0.2, height = 0.5, alpha = 0.6)
```

### Line Chart -- Trends Over Time

Perfect for time series. Revenue by month, stock prices, unemployment rates:

```{r}
ggplot(data = economics, aes(x = date, y = unemploy)) +
  geom_line(color = "steelblue")
```

### Smooth Line -- Trend Lines and Regression

Add a trend line to a scatter plot with `geom_smooth()`. This is incredibly useful for showing the overall direction of a relationship:

```{r}
ggplot(data = mpg, aes(x = displ, y = hwy)) +
  geom_point(alpha = 0.4) +
  geom_smooth()
```

Want a straight regression line instead of a curved one? Specify `method = "lm"`:

```{r}
ggplot(data = mpg, aes(x = displ, y = hwy)) +
  geom_point(alpha = 0.4) +
  geom_smooth(method = "lm", se = TRUE, color = "red")
```

### Boxplot -- Comparing Distributions Across Groups

Boxplots are fantastic for comparing a numeric variable across categories -- like highway mileage by vehicle class:

```{r}
ggplot(data = mpg, aes(x = class, y = hwy)) +
  geom_boxplot(fill = "lightblue", color = "navy")
```

The box shows the middle 50% of the data, the line is the median, and the dots are outliers. Quick and effective.

### Layering Multiple Geoms

One of the best things about ggplot2 is that you can stack geoms on top of each other. Points plus a smooth line plus a reference line? No problem:

```{r}
ggplot(data = mpg, aes(x = displ, y = hwy)) +
  geom_point(aes(color = class)) +
  geom_smooth(color = "black", se = FALSE) +
  geom_hline(yintercept = mean(mpg$hwy), linetype = "dotted", color = "gray50")
```

That dotted line shows the overall average. Layering like this makes your charts tell a richer story.


## Faceting -- Small Multiples That Reveal Patterns

Faceting splits your data into subsets and creates a separate panel for each one. It is one of the most powerful tools in ggplot2 for comparing groups.

### facet_wrap()

Wraps panels in a row-by-row layout:

```{r}
ggplot(data = mpg, aes(x = displ, y = hwy)) +
  geom_point() +
  facet_wrap(~ class)
```

You can control the layout:

```{r}
ggplot(data = mpg, aes(x = displ, y = hwy)) +
  geom_point() +
  facet_wrap(~ class, ncol = 4)
```

### facet_grid()

Creates a grid defined by two variables -- rows and columns:

```{r}
ggplot(data = mpg, aes(x = displ, y = hwy)) +
  geom_point() +
  facet_grid(drv ~ cyl)
```

### Letting Axes Vary with Free Scales

By default, all panels share the same axes. Sometimes that hides patterns. Use `scales = "free_y"` (or `"free_x"` or `"free"`) to let axes adjust per panel:

```{r}
ggplot(data = mpg, aes(x = hwy)) +
  geom_histogram(binwidth = 2, fill = "steelblue", color = "white") +
  facet_wrap(~ class, scales = "free_y")
```


## Position Adjustments for Bar Charts

When you have a bar chart with a fill variable, you need to decide how the bars relate to each other. There are three options you should know.

### Stacked (the default)

Bars stacked on top of each other. Good for showing total sizes and composition:

```{r}
ggplot(data = diamonds, aes(x = cut, fill = clarity)) +
  geom_bar(position = "stack")
```

### Dodged (side by side)

Bars placed next to each other. Better for comparing individual groups:

```{r}
ggplot(data = diamonds, aes(x = cut, fill = clarity)) +
  geom_bar(position = "dodge")
```

### Filled (proportions)

All bars stretched to the same height, showing percentages. Great for "what share of each cut has this clarity level?":

```{r}
ggplot(data = diamonds, aes(x = cut, fill = clarity)) +
  geom_bar(position = "fill") +
  ylab("Proportion")
```


## Coordinate Systems

A few coordinate tweaks you should know about.

### coord_flip() -- Horizontal Bar Charts

When your category labels are long, flip the axes:

```{r}
ggplot(data = mpg, aes(x = class)) +
  geom_bar(fill = "steelblue") +
  coord_flip()
```

### coord_cartesian() -- Zooming Without Losing Data

This lets you zoom into a region of your plot without throwing away any data points:

```{r}
ggplot(data = mpg, aes(x = displ, y = hwy)) +
  geom_point() +
  geom_smooth() +
  coord_cartesian(xlim = c(3, 6), ylim = c(15, 35))
```

This is different from setting axis limits directly, which actually *removes* data outside the range and can mess up trend lines.


## Labels and Annotations -- Making Charts Presentation-Ready

A chart without labels is like a report without a title. Always label your plots.

### The labs() Function

`labs()` is your one-stop shop for titles, subtitles, captions, and axis labels:

```{r}
ggplot(data = mpg, aes(x = displ, y = hwy, color = class)) +
  geom_point() +
  labs(
    title = "Fuel Efficiency vs. Engine Displacement",
    subtitle = "Data from EPA for 38 popular car models (1999-2008)",
    caption = "Source: mpg dataset in ggplot2",
    x = "Engine Displacement (liters)",
    y = "Highway Fuel Economy (mpg)",
    color = "Vehicle Class"
  )
```

Every chart you put in front of another human being should have at minimum a title and clear axis labels. Subtitles and captions are bonus polish.

### annotate() -- Adding Notes to Specific Spots

The `annotate()` function lets you place text, rectangles, or arrows at exact positions on your plot:

```{r}
ggplot(data = mpg, aes(x = displ, y = hwy)) +
  geom_point(alpha = 0.5) +
  annotate(
    "text",
    x = 6, y = 40,
    label = "High efficiency\nlarge engines",
    color = "red", fontface = "italic", size = 4
  ) +
  annotate(
    "rect",
    xmin = 5, xmax = 7, ymin = 35, ymax = 45,
    alpha = 0.1, fill = "red", color = "red", linetype = "dashed"
  )
```

Reference lines are also great for adding context -- like showing the average:

```{r}
ggplot(data = mpg, aes(x = displ, y = hwy)) +
  geom_point(alpha = 0.5) +
  geom_hline(yintercept = mean(mpg$hwy), linetype = "dashed", color = "red") +
  geom_vline(xintercept = mean(mpg$displ), linetype = "dashed", color = "blue")
```


## Saving Your Plots

Made something great? Save it with `ggsave()`:

```{r eval=FALSE}
# Save the last plot as a PNG
ggsave("my_plot.png", width = 8, height = 6, dpi = 300)

# Save as PDF (great for print or presentations)
ggsave("my_plot.pdf", width = 8, height = 6)

# Save a specific plot object
my_plot <- ggplot(mpg, aes(displ, hwy)) + geom_point()
ggsave("specific_plot.png", plot = my_plot, width = 10, height = 7, dpi = 150)
```


## Putting It All Together

Let's build a polished visualization step by step, applying everything from this chapter. We will use the `diamonds` dataset to explore how price relates to carat weight across different cut qualities.

### Start simple

```{r}
ggplot(data = diamonds, aes(x = carat, y = price)) +
  geom_point()
```

Over 50,000 data points. Everything is overplotted into a blob. Let's fix that.

### Fix overplotting

```{r}
ggplot(data = diamonds, aes(x = carat, y = price)) +
  geom_point(alpha = 0.05, size = 0.5)
```

### Add color for cut quality

```{r}
ggplot(data = diamonds, aes(x = carat, y = price, color = cut)) +
  geom_point(alpha = 0.05, size = 0.3) +
  geom_smooth(se = FALSE, linewidth = 1)
```

### Facet for clarity and polish it

```{r}
ggplot(data = diamonds, aes(x = carat, y = price)) +
  geom_point(alpha = 0.05, size = 0.3, color = "steelblue") +
  geom_smooth(color = "red", se = FALSE) +
  facet_wrap(~ cut) +
  scale_y_continuous(labels = scales::dollar_format()) +
  labs(
    title = "Diamond Price by Carat Weight and Cut Quality",
    subtitle = "Each panel shows one of five cut grades; red line is a LOESS smooth",
    caption = "Source: diamonds dataset in ggplot2",
    x = "Carat Weight",
    y = "Price (USD)"
  ) +
  theme_minimal(base_size = 12) +
  theme(
    plot.title = element_text(face = "bold"),
    strip.text = element_text(face = "bold"),
    panel.grid.minor = element_blank()
  )
```

That is the grammar of graphics at work. We layered data, aesthetics, geoms, facets, labels, and a theme -- all built incrementally so you can experiment and debug at each step.

### One more: a time series

```{r}
ggplot(data = economics, aes(x = date, y = unemploy / 1000)) +
  geom_line(color = "steelblue", linewidth = 0.8) +
  geom_smooth(method = "loess", span = 0.1, se = FALSE, color = "red", linewidth = 0.5) +
  scale_x_date(date_labels = "%Y", date_breaks = "5 years") +
  labs(
    title = "U.S. Unemployment Over Time",
    subtitle = "Monthly number of unemployed persons (in thousands), 1967-2015",
    caption = "Source: economics dataset in ggplot2",
    x = NULL,
    y = "Unemployed (thousands)"
  ) +
  theme_minimal(base_size = 12) +
  theme(
    plot.title = element_text(face = "bold"),
    axis.text.x = element_text(angle = 45, hjust = 1),
    panel.grid.minor = element_blank()
  )
```

You now have everything you need to build charts that impress people. In the next chapter, we will cover how to customize colors, themes, and layouts to make your plots look truly professional.
