# Data Visualization with ggplot2 {#ggplot2basics}

```{r include=FALSE}
library(tidyverse)
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

If data is the new oil, then **ggplot2** is the refinery that turns crude numbers into something people actually want to look at. It is the most popular visualization package in R, and for good reason --- it makes charts that are clean enough for a board presentation and flexible enough for a quick-and-dirty exploration of last quarter's sales.

Journalists at The New York Times use it. Analysts at Fortune 500 companies use it. After this chapter, you will use it too.

> **Note:** This chapter teaches you *how* to build charts in R. For the *why* behind choosing the right chart type --- when to use a bar chart vs. a line chart, how to design for your audience --- check the [Visual Analytics and Data Storytelling](http://patilv.com/VizAnalytics4Biz/) course.

## The Grammar of Graphics (the 30-Second Version)

Here is the big idea: every chart is built from the same three ingredients, kind of like how every business presentation has data, a story, and (hopefully) a point.

1. **Data** -- the dataset you want to visualize.
2. **Aesthetics** (`aes()`) -- which columns go where. This is where you say "put Revenue on the y-axis and Quarter on the x-axis."
3. **Geoms** -- the shapes that represent your data (points, bars, lines, etc.).

Data + mapping + shape. You can layer on extras (facets, labels, themes), but those three are the foundation of every single ggplot2 chart.

The template:

```
ggplot(data = <DATA>, mapping = aes(<MAPPINGS>)) +
  <GEOM_FUNCTION>()
```

You call `ggplot()`, tell it what data and mappings to use, then add a geom with `+`. That is it. Let's see it in action.


## Your First ggplot

We will use the `mpg` dataset that comes built into ggplot2 --- fuel economy data for popular car models. Think of it as a product performance report for the auto industry.

```{r}
glimpse(mpg)
```

### Step 1: Create the canvas

`ggplot()` sets up a blank canvas. By itself, it draws nothing --- like opening PowerPoint and staring at a blank slide:

```{r}
ggplot(data = mpg)
```

### Step 2: Map your variables

Now we tell ggplot2 which variables go where using `aes()`. Still no visible data --- just the axes:

```{r}
ggplot(data = mpg, mapping = aes(x = displ, y = hwy))
```

### Step 3: Add the geom

Finally, we tell ggplot2 *how* to draw the data. For a scatter plot, use `geom_point()`:

```{r}
ggplot(data = mpg, mapping = aes(x = displ, y = hwy)) +
  geom_point()
```

Bigger engines (`displ`) tend to get worse highway mileage (`hwy`). The outliers in the upper right? Two-seater sports cars with efficient highway gearing. Even your first chart tells a story.

> **Common mistake:** Putting the `+` at the *beginning* of a new line instead of the *end* of the previous line. The `+` must come at the end. Think of it like ending a sentence with "and" before starting the next clause.


## Aesthetic Mappings -- Making Charts Informative

`aes()` is where you tell ggplot which columns from your data control which visual properties. Think of it as assigning roles in a play --- "you are the x-axis, you control the color, you determine the size."

| Aesthetic  | What It Controls                  |
|------------|-----------------------------------|
| `x`        | Position on the x-axis            |
| `y`        | Position on the y-axis            |
| `color`    | Color of points, lines, outlines  |
| `fill`     | Fill color of bars and areas      |
| `size`     | Size of points or line thickness  |
| `shape`    | Shape of points                   |
| `alpha`    | Transparency (0 = invisible, 1 = solid) |

### Mapping vs. Setting -- This Trips Everyone Up

There is one distinction that trips up nearly every beginner. Read this twice.

**Mapping** (inside `aes()`) connects a *variable* to a visual property. ggplot2 picks colors automatically and creates a legend:

```{r}
ggplot(data = mpg, aes(x = displ, y = hwy, color = class)) +
  geom_point()
```

Each car class gets its own color. Legend appears automatically. You did not pick the colors --- ggplot did.

**Setting** (outside `aes()`) applies a *fixed* value to everything. No variable, no legend:

```{r}
ggplot(data = mpg, aes(x = displ, y = hwy)) +
  geom_point(color = "steelblue", size = 3, alpha = 0.6)
```

All points are steel blue, size 3, 60% opaque. If you accidentally put `color = "steelblue"` inside `aes()`, you will get bizarre results and a useless legend. Mapped aesthetics go inside `aes()`. Fixed values go outside.

### Combining Multiple Aesthetics

You can map several variables at once for richer, more informative charts:

```{r}
ggplot(data = mpg, aes(x = displ, y = hwy, color = class, shape = drv)) +
  geom_point(size = 2.5)
```

Map a continuous variable to color and you get a gradient:

```{r}
ggplot(data = mpg, aes(x = displ, y = hwy, color = cty)) +
  geom_point(size = 2)
```


## Geoms -- The Chart Types That Matter

Geoms are the building blocks. Each one draws a different shape. Here are the ones you will use 90% of the time in any business context.

### Bar Chart -- Counting Categories

How many observations in each category? `geom_bar()` counts for you automatically --- no pre-calculation needed:

```{r}
ggplot(data = mpg, aes(x = class)) +
  geom_bar(fill = "steelblue")
```

If you already have pre-computed values (say, average revenue by region from a summary table), use `geom_col()` instead:

```{r}
mpg_avg <- mpg %>%
  group_by(class) %>%
  summarize(avg_hwy = mean(hwy))

ggplot(data = mpg_avg, aes(x = class, y = avg_hwy)) +
  geom_col(fill = "steelblue") +
  geom_text(aes(label = round(avg_hwy, 1)), vjust = -0.5, size = 3.5)
```

Adding value labels on top of bars is the kind of small touch that makes your manager think you have been doing this for years.

### Histogram -- Distribution of a Continuous Variable

"How are values spread out?" is one of the first questions any analyst should ask. Histograms answer it:

```{r}
ggplot(data = mpg, aes(x = hwy)) +
  geom_histogram(binwidth = 2, fill = "steelblue", color = "white")
```

Play with `binwidth` to change the granularity. Too few bins hides patterns; too many creates noise. It is like choosing the right zoom level on a map.

### Scatter Plot -- Relationships Between Two Numbers

The workhorse of exploratory analysis. Great for spotting correlations, clusters, and that one weird outlier your intern accidentally entered:

```{r}
ggplot(data = mpg, aes(x = displ, y = hwy)) +
  geom_point()
```

When points pile on top of each other (overplotting), use transparency or jitter:

```{r}
ggplot(data = mpg, aes(x = displ, y = hwy)) +
  geom_jitter(width = 0.2, height = 0.5, alpha = 0.6)
```

### Line Chart -- Trends Over Time

The bread and butter of business dashboards. Revenue by month, stock prices, KPIs over time:

```{r}
ggplot(data = economics, aes(x = date, y = unemploy)) +
  geom_line(color = "steelblue")
```

### Smooth Line -- Trend Lines and Regression

Add a trend line to a scatter plot with `geom_smooth()`. This is incredibly useful for showing the overall direction without getting lost in the noise:

```{r}
ggplot(data = mpg, aes(x = displ, y = hwy)) +
  geom_point(alpha = 0.4) +
  geom_smooth()
```

Want a straight regression line? Specify `method = "lm"`:

```{r}
ggplot(data = mpg, aes(x = displ, y = hwy)) +
  geom_point(alpha = 0.4) +
  geom_smooth(method = "lm", se = TRUE, color = "red")
```

### Boxplot -- Comparing Distributions Across Groups

Boxplots let you compare a numeric variable across categories at a glance --- like highway mileage by vehicle class:

```{r}
ggplot(data = mpg, aes(x = class, y = hwy)) +
  geom_boxplot(fill = "lightblue", color = "navy")
```

The box is the middle 50%, the line is the median, and the dots are outliers. It is the executive summary of a distribution.

### Layering Multiple Geoms

One of the best things about ggplot2 is that you stack geoms like layers on a cake. Points plus a trend line plus a reference line? No problem:

```{r}
ggplot(data = mpg, aes(x = displ, y = hwy)) +
  geom_point(aes(color = class)) +
  geom_smooth(color = "black", se = FALSE) +
  geom_hline(yintercept = mean(mpg$hwy), linetype = "dotted", color = "gray50")
```

That dotted line shows the overall average. Layering like this makes your charts tell a richer, more complete story.


## Faceting -- One Chart Per Category

Faceting splits your data into subsets and creates a separate panel for each one. In the data viz world, these are called "small multiples," and they are one of the most powerful ways to compare groups without cramming everything into a single cluttered chart.

### facet_wrap()

Wraps panels in a row-by-row layout:

```{r}
ggplot(data = mpg, aes(x = displ, y = hwy)) +
  geom_point() +
  facet_wrap(~ class)
```

Control the layout with `ncol`:

```{r}
ggplot(data = mpg, aes(x = displ, y = hwy)) +
  geom_point() +
  facet_wrap(~ class, ncol = 4)
```

### facet_grid()

Creates a grid defined by two variables --- rows and columns. Think of it as a pivot table, but visual:

```{r}
ggplot(data = mpg, aes(x = displ, y = hwy)) +
  geom_point() +
  facet_grid(drv ~ cyl)
```

### Letting Axes Vary with Free Scales

By default, all panels share the same axes. Sometimes that hides patterns in smaller groups. Use `scales = "free_y"` (or `"free_x"` or `"free"`) to let each panel set its own range:

```{r}
ggplot(data = mpg, aes(x = hwy)) +
  geom_histogram(binwidth = 2, fill = "steelblue", color = "white") +
  facet_wrap(~ class, scales = "free_y")
```


## Position Adjustments for Bar Charts

When your bar chart has a fill variable, you need to decide how the bars relate to each other. Three options worth knowing:

### Stacked (the default)

Bars stacked on top of each other. Good for showing total sizes and composition:

```{r}
ggplot(data = diamonds, aes(x = cut, fill = clarity)) +
  geom_bar(position = "stack")
```

### Dodged (side by side)

Bars placed next to each other. Better for comparing individual groups:

```{r}
ggplot(data = diamonds, aes(x = cut, fill = clarity)) +
  geom_bar(position = "dodge")
```

### Filled (proportions)

All bars stretched to the same height, showing percentages. Great for "what share of each category falls into each bucket?":

```{r}
ggplot(data = diamonds, aes(x = cut, fill = clarity)) +
  geom_bar(position = "fill") +
  ylab("Proportion")
```


## Coordinate Systems

A couple of coordinate tweaks that come up regularly.

### coord_flip() -- Horizontal Bar Charts

When your category labels are long and overlap, just flip the axes:

```{r}
ggplot(data = mpg, aes(x = class)) +
  geom_bar(fill = "steelblue") +
  coord_flip()
```

### coord_cartesian() -- Zooming Without Losing Data

This lets you zoom into a region of your plot without actually throwing away data points:

```{r}
ggplot(data = mpg, aes(x = displ, y = hwy)) +
  geom_point() +
  geom_smooth() +
  coord_cartesian(xlim = c(3, 6), ylim = c(15, 35))
```

This is different from setting axis limits directly, which *removes* data outside the range and can distort trend lines. `coord_cartesian()` just changes the camera angle.


## Labels and Annotations -- Making Charts Presentation-Ready

A chart without labels is like a resume without your name on it. Technically it exists, but it is not doing you any favors.

### The labs() Function

`labs()` is your one-stop shop for titles, subtitles, captions, and axis labels:

```{r}
ggplot(data = mpg, aes(x = displ, y = hwy, color = class)) +
  geom_point() +
  labs(
    title = "Fuel Efficiency vs. Engine Displacement",
    subtitle = "Data from EPA for 38 popular car models (1999-2008)",
    caption = "Source: mpg dataset in ggplot2",
    x = "Engine Displacement (liters)",
    y = "Highway Fuel Economy (mpg)",
    color = "Vehicle Class"
  )
```

Every chart you put in front of another human being should have, at minimum, a title and clear axis labels. Subtitles and captions are bonus polish that signal you actually care about your audience.

### annotate() -- Adding Notes to Specific Spots

`annotate()` lets you place text, rectangles, or arrows at exact positions on your plot --- perfect for calling out that anomaly your boss will definitely ask about:

```{r}
ggplot(data = mpg, aes(x = displ, y = hwy)) +
  geom_point(alpha = 0.5) +
  annotate(
    "text",
    x = 6, y = 40,
    label = "High efficiency\nlarge engines",
    color = "red", fontface = "italic", size = 4
  ) +
  annotate(
    "rect",
    xmin = 5, xmax = 7, ymin = 35, ymax = 45,
    alpha = 0.1, fill = "red", color = "red", linetype = "dashed"
  )
```

Reference lines are also great for adding context --- like showing the average so people know where "normal" is:

```{r}
ggplot(data = mpg, aes(x = displ, y = hwy)) +
  geom_point(alpha = 0.5) +
  geom_hline(yintercept = mean(mpg$hwy), linetype = "dashed", color = "red") +
  geom_vline(xintercept = mean(mpg$displ), linetype = "dashed", color = "blue")
```


## Saving Your Plots

Made something worth sharing? Save it with `ggsave()`:

```{r eval=FALSE}
# Save the last plot as a PNG
ggsave("my_plot.png", width = 8, height = 6, dpi = 300)

# Save as PDF (great for print or presentations)
ggsave("my_plot.pdf", width = 8, height = 6)

# Save a specific plot object
my_plot <- ggplot(mpg, aes(displ, hwy)) + geom_point()
ggsave("specific_plot.png", plot = my_plot, width = 10, height = 7, dpi = 150)
```


## Putting It All Together

Let's build a polished visualization step by step, applying everything from this chapter. We will use the `diamonds` dataset to see how price relates to carat weight across cut qualities.

### Start simple

```{r}
ggplot(data = diamonds, aes(x = carat, y = price)) +
  geom_point()
```

Over 50,000 data points. Everything is overplotted into an unreadable blob. Let's fix that.

### Fix overplotting

```{r}
ggplot(data = diamonds, aes(x = carat, y = price)) +
  geom_point(alpha = 0.05, size = 0.5)
```

### Add color for cut quality

```{r}
ggplot(data = diamonds, aes(x = carat, y = price, color = cut)) +
  geom_point(alpha = 0.05, size = 0.3) +
  geom_smooth(se = FALSE, linewidth = 1)
```

### Facet for clarity and polish it

```{r}
ggplot(data = diamonds, aes(x = carat, y = price)) +
  geom_point(alpha = 0.05, size = 0.3, color = "steelblue") +
  geom_smooth(color = "red", se = FALSE) +
  facet_wrap(~ cut) +
  scale_y_continuous(labels = scales::dollar_format()) +
  labs(
    title = "Diamond Price by Carat Weight and Cut Quality",
    subtitle = "Each panel shows one of five cut grades; red line is a LOESS smooth",
    caption = "Source: diamonds dataset in ggplot2",
    x = "Carat Weight",
    y = "Price (USD)"
  ) +
  theme_minimal(base_size = 12) +
  theme(
    plot.title = element_text(face = "bold"),
    strip.text = element_text(face = "bold"),
    panel.grid.minor = element_blank()
  )
```

That is the grammar of graphics at work. Data, aesthetics, geoms, facets, labels, theme --- all layered incrementally so you can experiment and debug at each step. It is like building a presentation slide by slide instead of trying to design the whole deck at once.

### One more: a time series

```{r}
ggplot(data = economics, aes(x = date, y = unemploy / 1000)) +
  geom_line(color = "steelblue", linewidth = 0.8) +
  geom_smooth(method = "loess", span = 0.1, se = FALSE, color = "red", linewidth = 0.5) +
  scale_x_date(date_labels = "%Y", date_breaks = "5 years") +
  labs(
    title = "U.S. Unemployment Over Time",
    subtitle = "Monthly number of unemployed persons (in thousands), 1967-2015",
    caption = "Source: economics dataset in ggplot2",
    x = NULL,
    y = "Unemployed (thousands)"
  ) +
  theme_minimal(base_size = 12) +
  theme(
    plot.title = element_text(face = "bold"),
    axis.text.x = element_text(angle = 45, hjust = 1),
    panel.grid.minor = element_blank()
  )
```

You now have everything you need to build charts that make people stop scrolling through their phones in meetings. In the next chapter, we will cover how to customize colors, themes, and layouts to make your plots look truly professional.
