# Customizing ggplot2 Visualizations {#ggplot2customize}

```{r include=FALSE}
library(tidyverse)
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

You know how to build charts. Now let's make them look *good*. The default ggplot2 appearance -- that gray background with white gridlines -- is fine for quick exploration, but it is not what you want in a slide deck, a client report, or a dashboard. This chapter is about going from "I made a chart" to "I made a chart that looks professional."

We will cover:

- **Themes** -- swap out the entire visual style in one line
- **Colors** -- use palettes that look polished and are accessible
- **Axis formatting** -- dollar signs, commas, percentages on your axes
- **Legends** -- move them, restyle them, or get rid of them
- **Annotations** -- draw attention to the things that matter
- **Combining plots** -- put multiple charts together for a complete story

We will use the built-in `mpg`, `iris`, and `diamonds` datasets throughout.

## Themes -- One-Line Style Upgrades

### Built-in Themes

Every ggplot2 plot uses a theme to control non-data elements like backgrounds, gridlines, and fonts. The default is `theme_gray()`. Here are the ones worth knowing:

```{r fig.height=8, fig.width=10}
base_plot <- ggplot(mpg, aes(x = displ, y = hwy)) +
  geom_point(aes(color = class)) +
  labs(title = "Engine Displacement vs. Highway MPG")

library(patchwork)

p1 <- base_plot + theme_gray() + ggtitle("theme_gray() (default)")
p2 <- base_plot + theme_bw() + ggtitle("theme_bw()")
p3 <- base_plot + theme_minimal() + ggtitle("theme_minimal()")
p4 <- base_plot + theme_classic() + ggtitle("theme_classic()")
p5 <- base_plot + theme_light() + ggtitle("theme_light()")

(p1 | p2 | p3) / (p4 | p5)
```

Here is the quick rundown:

- **`theme_bw()`** -- Removes the gray background. Clean, great for printed reports.
- **`theme_minimal()`** -- Even cleaner. This is probably the most popular choice for business presentations.
- **`theme_classic()`** -- Looks like a textbook chart. Axis lines, no gridlines.
- **`theme_light()`** -- Light gray gridlines. A nice middle ground.

For most business work, `theme_minimal()` or `theme_bw()` will serve you well.

### Themes from ggthemes

The **ggthemes** package provides themes inspired by well-known publications. If you want your chart to look like it belongs in The Economist or the Wall Street Journal:

```{r fig.height=8, fig.width=10}
library(ggthemes)

t1 <- base_plot + theme_economist() + ggtitle("theme_economist()")
t2 <- base_plot + theme_fivethirtyeight() + ggtitle("theme_fivethirtyeight()")
t3 <- base_plot + theme_tufte() + ggtitle("theme_tufte()")
t4 <- base_plot + theme_wsj() + ggtitle("theme_wsj()")

(t1 | t2) / (t3 | t4)
```

Fun to play with, and occasionally the right choice for a specific audience.

### Setting a Global Theme

Tired of adding `+ theme_minimal()` to every plot? Set it once at the top of your script:

```{r}
theme_set(theme_minimal())

# Now every plot uses theme_minimal() automatically
ggplot(mpg, aes(x = class)) +
  geom_bar(fill = "steelblue") +
  labs(title = "Vehicle Class Counts", x = "Class", y = "Count")
```

```{r include=FALSE}
# Reset to default for the rest of the chapter
theme_set(theme_gray())
```

## Fine-Tuning with theme()

Sometimes you want a built-in theme but with a few tweaks. The `theme()` function lets you modify individual elements using four helpers:

| Helper Function | Controls | Key Arguments |
|:---|:---|:---|
| `element_text()` | Text | `family`, `face`, `size`, `color`, `angle` |
| `element_line()` | Lines | `color`, `linewidth`, `linetype` |
| `element_rect()` | Rectangles/borders | `fill`, `color` |
| `element_blank()` | Removes the element | (none) |

### Example: Making Titles Stand Out

```{r}
ggplot(mpg, aes(x = displ, y = hwy, color = class)) +
  geom_point(size = 2) +
  labs(
    title = "Fuel Efficiency by Engine Size",
    subtitle = "Larger engines tend to have lower highway mileage",
    caption = "Source: EPA fuel economy data (mpg dataset)",
    x = "Engine Displacement (liters)",
    y = "Highway MPG",
    color = "Vehicle Class"
  ) +
  theme(
    plot.title = element_text(face = "bold", size = 16, color = "navy"),
    plot.subtitle = element_text(face = "italic", size = 12, color = "gray40"),
    plot.caption = element_text(size = 9, hjust = 0, color = "gray50"),
    axis.title = element_text(face = "bold", size = 12)
  )
```

### Example: Cleaning Up the Background

```{r}
ggplot(mpg, aes(x = cty, y = hwy)) +
  geom_point(alpha = 0.5, color = "darkorange") +
  labs(title = "City vs. Highway Fuel Economy") +
  theme(
    panel.background = element_rect(fill = "white"),
    plot.background = element_rect(fill = "gray95"),
    panel.grid.major = element_line(color = "gray80", linewidth = 0.5),
    panel.grid.minor = element_blank()
  )
```

### Example: Rotating Long Axis Labels

When your category labels overlap, angle them:

```{r}
ggplot(mpg, aes(x = manufacturer)) +
  geom_bar(fill = "coral") +
  labs(title = "Vehicles by Manufacturer", x = NULL, y = "Count") +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1, size = 9),
    plot.title = element_text(face = "bold", size = 14)
  )
```


## Colors That Look Good

Color is one of the most impactful things you can change. ggplot2 gives you several ways to control it.

### Manual Colors -- Pick Your Own

Use `scale_color_manual()` (for points and lines) or `scale_fill_manual()` (for bars and areas) to set exact colors:

```{r}
ggplot(mpg, aes(x = displ, y = hwy, color = drv)) +
  geom_point(size = 2) +
  scale_color_manual(
    values = c("f" = "#E41A1C", "r" = "#377EB8", "4" = "#4DAF4A"),
    labels = c("f" = "Front", "r" = "Rear", "4" = "Four-Wheel")
  ) +
  labs(title = "MPG by Engine Size and Drive Type", color = "Drive Type")
```

This is great when your company has brand colors or when you want precise control.

### ColorBrewer -- Pre-Made Palettes That Work

ColorBrewer palettes were designed by a cartographer to be visually clear and print-friendly. They are excellent defaults:

```{r fig.height=6, fig.width=8}
RColorBrewer::display.brewer.all()
```

Three categories:

- **Sequential** (Blues, Greens, OrRd) -- for ordered data, low to high
- **Diverging** (RdBu, RdYlGn) -- for data with a meaningful midpoint
- **Qualitative** (Set1, Set2, Dark2) -- for unordered categories

```{r}
ggplot(mpg, aes(x = class, fill = class)) +
  geom_bar() +
  scale_fill_brewer(palette = "Set2") +
  labs(title = "Vehicle Counts by Class (Set2 palette)") +
  theme(legend.position = "none")
```

```{r}
ggplot(iris, aes(x = Sepal.Length, y = Sepal.Width, color = Species)) +
  geom_point(size = 2.5) +
  scale_color_brewer(palette = "Dark2") +
  labs(title = "Iris Measurements (Dark2 palette)")
```

### Viridis -- Colorblind-Friendly and Beautiful

The viridis palettes are perceptually uniform and work for people with color vision deficiencies. They are a great default choice, especially for continuous data:

```{r fig.height=8, fig.width=10}
diamonds_sample <- diamonds[sample(nrow(diamonds), 2000), ]

v1 <- ggplot(diamonds_sample, aes(x = carat, y = price, color = depth)) +
  geom_point(alpha = 0.6) +
  scale_color_viridis_c(option = "viridis") +
  labs(title = "viridis")

v2 <- ggplot(diamonds_sample, aes(x = carat, y = price, color = depth)) +
  geom_point(alpha = 0.6) +
  scale_color_viridis_c(option = "magma") +
  labs(title = "magma")

v3 <- ggplot(diamonds_sample, aes(x = carat, y = price, color = depth)) +
  geom_point(alpha = 0.6) +
  scale_color_viridis_c(option = "plasma") +
  labs(title = "plasma")

v4 <- ggplot(diamonds_sample, aes(x = carat, y = price, color = depth)) +
  geom_point(alpha = 0.6) +
  scale_color_viridis_c(option = "inferno") +
  labs(title = "inferno")

(v1 | v2) / (v3 | v4)
```

For discrete (categorical) data, use `scale_color_viridis_d()` or `scale_fill_viridis_d()`:

```{r}
ggplot(mpg, aes(x = class, fill = class)) +
  geom_bar() +
  scale_fill_viridis_d(option = "plasma") +
  labs(title = "Vehicle Class with Viridis Discrete (Plasma)") +
  theme(legend.position = "none")
```


## Axis Formatting -- Dollars, Commas, and Percentages

Nothing says "this chart is for business people" like properly formatted axis labels.

### The scales Package

The **scales** package gives you label formatters that handle the common cases:

```{r}
diamonds_summary <- diamonds %>%
  group_by(cut) %>%
  summarize(avg_price = mean(price), count = n())

ggplot(diamonds_summary, aes(x = cut, y = avg_price)) +
  geom_col(fill = "darkgreen") +
  scale_y_continuous(labels = scales::dollar) +
  labs(title = "Average Diamond Price by Cut", x = "Cut Quality", y = "Average Price")
```

```{r}
ggplot(diamonds_summary, aes(x = cut, y = count)) +
  geom_col(fill = "steelblue") +
  scale_y_continuous(labels = scales::comma) +
  labs(title = "Diamond Count by Cut", x = "Cut Quality", y = "Number of Diamonds")
```

```{r}
mpg_pct <- mpg %>%
  count(class) %>%
  mutate(pct = n / sum(n))

ggplot(mpg_pct, aes(x = reorder(class, pct), y = pct)) +
  geom_col(fill = "tomato") +
  scale_y_continuous(labels = scales::percent) +
  coord_flip() +
  labs(title = "Percentage of Vehicles by Class", x = NULL, y = "Percentage")
```

### Log Scales

When your data spans several orders of magnitude (like revenue data that goes from $100 to $100M), log scales help:

```{r}
ggplot(diamonds, aes(x = carat, y = price)) +
  geom_point(alpha = 0.05, color = "purple") +
  scale_x_log10() +
  scale_y_log10(labels = scales::dollar) +
  labs(title = "Diamond Price vs. Carat (Log Scale)", x = "Carat (log)", y = "Price (log)")
```

### Zooming vs. Filtering -- An Important Distinction

There are two ways to limit your axes, and they behave very differently:

- **`coord_cartesian()`** zooms in *without removing data*. Trend lines and statistics stay accurate.
- **`scale_*_continuous(limits = ...)`** *removes data* outside the range, which can change trend lines.

```{r fig.width=10, fig.height=5}
p_base <- ggplot(mpg, aes(x = displ, y = hwy)) +
  geom_point() +
  geom_smooth(method = "lm")

p_scale <- p_base +
  scale_x_continuous(limits = c(3, 6)) +
  ggtitle("scale limits (data removed)")

p_coord <- p_base +
  coord_cartesian(xlim = c(3, 6)) +
  ggtitle("coord_cartesian (zoom only)")

p_scale | p_coord
```

See how the regression lines differ? Use `coord_cartesian()` when you want to zoom without distorting your analysis.


## Legend Customization

### Moving the Legend Around

```{r fig.height=8, fig.width=10}
base <- ggplot(mpg, aes(x = displ, y = hwy, color = class)) +
  geom_point()

l1 <- base + theme(legend.position = "top") + ggtitle("Top")
l2 <- base + theme(legend.position = "bottom") + ggtitle("Bottom")
l3 <- base + theme(legend.position = "left") + ggtitle("Left")
l4 <- base + theme(legend.position = "right") + ggtitle("Right (default)")

(l1 | l2) / (l3 | l4)
```

### Removing the Legend

When the information is obvious from context (like a bar chart where color matches the x-axis labels), just remove it:

```{r}
ggplot(mpg, aes(x = class, fill = class)) +
  geom_bar() +
  theme(legend.position = "none") +
  labs(title = "Bar Chart Without Legend")
```

### Customizing Legend Title and Labels

```{r}
ggplot(iris, aes(x = Sepal.Length, y = Petal.Length, color = Species)) +
  geom_point(size = 2) +
  scale_color_manual(
    values = c("setosa" = "#1b9e77", "versicolor" = "#d95f02", "virginica" = "#7570b3"),
    labels = c("Setosa", "Versicolor", "Virginica")
  ) +
  labs(color = "Iris Species", title = "Custom Legend Title and Labels")
```


## Annotations -- Drawing Attention Where It Matters

### annotate()

Add text, shapes, or arrows to specific spots on your chart:

```{r}
ggplot(mpg, aes(x = displ, y = hwy)) +
  geom_point(color = "gray50") +
  annotate("text", x = 6, y = 40, label = "Efficient outliers",
           color = "red", fontface = "bold", size = 5) +
  annotate("rect", xmin = 5.5, xmax = 7, ymin = 25, ymax = 45,
           alpha = 0.1, fill = "red", color = "red", linetype = "dashed") +
  labs(title = "Using annotate() for Text and Shapes")
```

### geom_text() and geom_label() for Data-Driven Labels

```{r}
top_classes <- mpg %>%
  count(class) %>%
  arrange(desc(n)) %>%
  head(5)

ggplot(top_classes, aes(x = reorder(class, n), y = n)) +
  geom_col(fill = "steelblue") +
  geom_text(aes(label = n), hjust = -0.3, size = 5, fontface = "bold") +
  coord_flip() +
  scale_y_continuous(expand = expansion(mult = c(0, 0.15))) +
  labs(title = "Top 5 Vehicle Classes", x = NULL, y = "Count")
```

### ggrepel for Non-Overlapping Labels

When labels pile on top of each other, the **ggrepel** package automatically spaces them out:

```{r}
library(ggrepel)

best_mpg <- mpg %>%
  group_by(manufacturer) %>%
  slice_max(hwy, n = 1) %>%
  ungroup() %>%
  slice_max(hwy, n = 8)

ggplot(mpg, aes(x = displ, y = hwy)) +
  geom_point(color = "gray70") +
  geom_point(data = best_mpg, color = "red", size = 3) +
  geom_label_repel(
    data = best_mpg,
    aes(label = paste(manufacturer, model)),
    size = 3,
    max.overlaps = 20,
    fill = "lightyellow"
  ) +
  labs(
    title = "Most Fuel-Efficient Vehicles by Manufacturer",
    subtitle = "Top 8 models highlighted",
    x = "Engine Displacement (L)",
    y = "Highway MPG"
  )
```


## Combining Plots with patchwork

In real life, you often need to present multiple charts together -- a scatter plot next to a bar chart, or a grid of related views. The **patchwork** package makes this easy with intuitive operators:

- `|` places plots side by side
- `/` stacks plots vertically
- `()` groups plots to control layout order

```{r fig.height=4, fig.width=10}
p_scatter <- ggplot(mpg, aes(x = displ, y = hwy)) +
  geom_point(color = "steelblue") +
  labs(title = "Scatter Plot")

p_box <- ggplot(mpg, aes(x = drv, y = hwy, fill = drv)) +
  geom_boxplot(show.legend = FALSE) +
  labs(title = "Box Plot")

p_bar <- ggplot(mpg, aes(x = class)) +
  geom_bar(fill = "coral") +
  labs(title = "Bar Chart") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

p_scatter | p_box | p_bar
```

### Controlling Layout

```{r fig.height=7, fig.width=10}
(p_scatter | p_box) / p_bar +
  plot_layout(heights = c(2, 1))
```

### Adding a Shared Title and Tags

Tags automatically label your panels A, B, C -- perfect for reports and papers:

```{r fig.height=7, fig.width=10}
(p_scatter | p_box) / p_bar +
  plot_annotation(
    title = "Overview of the mpg Dataset",
    subtitle = "Three complementary views of the data",
    caption = "Source: EPA fuel economy data",
    tag_levels = "A"
  ) +
  plot_layout(heights = c(2, 1))
```

### Sharing a Legend Across Plots

When multiple plots use the same color scheme, collect the legends into one:

```{r fig.height=4, fig.width=10}
p_a <- ggplot(mpg, aes(x = displ, y = hwy, color = drv)) +
  geom_point()

p_b <- ggplot(mpg, aes(x = cty, y = hwy, color = drv)) +
  geom_point()

p_a + p_b +
  plot_layout(guides = "collect") +
  plot_annotation(title = "Shared Legend Between Plots")
```


## Summary

Here is your customization toolkit at a glance:

1. **Themes** -- Use `theme_minimal()` or `theme_bw()` for clean, professional charts. Try `ggthemes` for publication-specific styles. Set a global default with `theme_set()`.

2. **theme() tweaks** -- Fine-tune titles, axes, gridlines, and backgrounds with `element_text()`, `element_line()`, `element_rect()`, and `element_blank()`.

3. **Colors** -- Use `scale_color_brewer()` for categorical data, `scale_color_viridis_c()` for continuous data, and `scale_color_manual()` when you need exact colors.

4. **Axis formatting** -- Use the `scales` package for `dollar`, `comma`, and `percent` labels. Use `coord_cartesian()` to zoom without distorting statistics.

5. **Legends** -- Move with `theme(legend.position = ...)`. Remove with `"none"`. Customize labels in the scale function.

6. **Annotations** -- Use `annotate()` for fixed labels, `geom_text()` for data-driven labels, and `ggrepel` to prevent overlaps.

7. **Multi-panel figures** -- Use patchwork with `|`, `/`, `plot_layout()`, and `plot_annotation()` to combine charts into a cohesive story.

With these tools, you can take any default ggplot2 chart and turn it into something you would be proud to put in front of a client, a manager, or an audience of 500 people.
