# Working with Text Data {#stringr}

```{r include=FALSE}
library(tidyverse)
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

## Why You Need This

Here is an uncomfortable truth: a terrifying amount of real business data is just text. Survey responses. Product names. Customer addresses. Free-form comments from your NPS survey that say things like "  GREAT product!!  " and "great   Product" --- which are obviously the same sentiment, but R treats them as completely different values.

The **stringr** package is your cleanup crew. Every function starts with `str_`, takes the string first (so it works beautifully with the pipe), and does exactly what its name suggests. Let's get into it.


## String Basics

Strings in R live inside quotes. Double quotes are the convention.

```{r}
product <- "Wireless Bluetooth Speaker"
product
```

If your string contains a quote, escape it with a backslash or wrap in single quotes:

```{r}
tagline <- 'She said, "Take my money!"'
tagline
```

`str_length()` counts characters --- handy for flagging those one-word survey responses from people who clearly wanted to leave early:

```{r}
responses <- c("Great product!", "OK", "This changed my life and I want to tell everyone", "N/A")
str_length(responses)
```


## Combining Strings with `str_c()`

`str_c()` glues strings together. Think `paste()`, but tidyverse-flavored.

```{r}
str_c("Q", 1:4)
str_c("Hello", "World", sep = " ")
```

The `collapse` argument squashes a vector into one string --- great for building labels:

```{r}
departments <- c("Marketing", "Finance", "Operations")
str_c(departments, collapse = ", ")
str_c(departments, collapse = " & ")
```


## Changing Case --- Your Most-Used Trick

When survey respondents type "NEW YORK", "new york", and "New york" --- those are three different values to R. One line fixes it:

```{r}
cities <- c("new york", "LOS ANGELES", "chicago", "SAN FRANCISCO")
str_to_title(cities)
```

The full case family:

```{r}
sample_text <- "the QUICK brown FOX"
str_to_lower(sample_text)
str_to_upper(sample_text)
str_to_title(sample_text)
```

You will use these every single time you clean survey data, product categories, or anything where humans had a keyboard and zero standards.


## Trimming Whitespace

Extra spaces are the silent killer of data joins. If one spreadsheet has "Marketing" and another has "  Marketing  ", your `left_join()` will silently drop rows and you will silently lose your mind.

```{r}
messy <- "   too  much   space   "
str_trim(messy)       # removes leading/trailing spaces
str_squish(messy)     # also collapses internal spaces to one
```

`str_squish()` handles everything. Use it by default.


## Detecting Patterns with `str_detect()`

"Does this string contain this word?" That is `str_detect()`. It returns TRUE/FALSE, which makes it perfect inside `filter()`.

```{r}
products <- c("iPhone 15 Pro", "Samsung Galaxy S24", "iPhone 15", "Google Pixel 8", "iPhone SE")
str_detect(products, "iPhone")
```

Inside a data frame --- this is where it gets useful:

```{r}
product_df <- tibble(
  product = c("iPhone 15 Pro", "Samsung Galaxy S24", "iPhone 15", "Google Pixel 8"),
  revenue = c(45000, 32000, 38000, 15000)
)

product_df %>% filter(str_detect(product, "iPhone"))
```

You just filtered to Apple products without needing a separate category column. Your manager will never know how easy that was.


## Replacing Text with `str_replace()`

`str_replace()` fixes the first match. `str_replace_all()` fixes every match.

```{r}
reviews <- c("This product is gr8", "Loved it!! Will buy again!!", "Good product, good price")
str_replace_all(reviews, "!!", "!")
```

Standardizing how people wrote "percent":

```{r}
responses <- c("50 percent", "30%", "about 20 percent", "75%")
str_replace_all(responses, " percent", "%")
```

`str_remove()` and `str_remove_all()` are shortcuts for replacing with nothing:

```{r}
prices <- c("$29.99", "$149.50", "$5.00")
str_remove(prices, "\\$")
```


## Extracting Text with `str_extract()`

Need to pull a number out of a messy string? `str_extract()` grabs the first match.

```{r}
order_notes <- c("Order 1234 placed", "Invoice 5678 pending", "Shipped item 91011")
str_extract(order_notes, "\\d+")
```

That `\\d+` is a regular expression. Do not panic --- the next section covers the only ones you will ever need.


## Regular Expressions: Just the 3 Patterns You Actually Need

Regex has a reputation for looking like a cat walked across your keyboard. The good news: for business data cleaning, you only need three patterns.

**Pattern 1: `\\d+` --- Find numbers**

```{r}
sales_notes <- c("Sold 15 units", "Returned 3 items", "Ordered 200 widgets")
str_extract(sales_notes, "\\d+")
```

**Pattern 2: `[A-Za-z]+` --- Find words (letters only)**

```{r}
codes <- c("SKU-12345", "PROD-6789", "ITEM-111")
str_extract(codes, "[A-Z]+")
```

**Pattern 3: `[\\w.]+@[\\w.]+` --- Find email-like patterns**

```{r}
messages <- c("Contact alice@example.com for details",
              "Email bob.smith@company.org for info",
              "No email here")
str_extract(messages, "[\\w.]+@[\\w.]+")
```

That is it. Three patterns. If you need something fancier, Google "regex cheat sheet." No shame in that --- everyone does it, including the people who wrote the cheat sheet.


## Putting It All Together: A Survey Data Cleaning Pipeline

Here is a realistic scenario. You received survey data and it looks like it was filled out during an earthquake:

```{r}
survey_raw <- tibble(
  respondent = c("  JOHN doe  ", "jane SMITH", " Bob  Johnson ",
                 "ALICE williams", " charlie Brown"),
  email = c("John.Doe@Gmail.COM", "jane_smith@YAHOO.com",
            "BOB@company.ORG", "alice.w@School.Edu", "charlie@EMAIL.com"),
  phone = c("(555) 123-4567", "555.234.5678", "555 345 6789",
            "5554567890", "(555)-567-8901"),
  favorite_color = c("  Blue ", "RED", "  green  ", "BLUE", " Green")
)

survey_clean <- survey_raw %>%
  mutate(
    # Clean names: squish + title case
    respondent = respondent %>% str_squish() %>% str_to_title(),

    # Emails: just lowercase everything
    email = str_to_lower(email),

    # Phone numbers: strip everything that isn't a digit, then reformat
    phone = phone %>%
      str_remove_all("[^0-9]") %>%
      str_replace("(\\d{3})(\\d{3})(\\d{4})", "\\1-\\2-\\3"),

    # Colors: squish + lowercase for consistency
    favorite_color = favorite_color %>% str_squish() %>% str_to_lower()
  )

survey_clean
```

Five lines of `mutate()` and the chaos is gone. Let's verify the colors standardized:

```{r}
survey_clean %>% count(favorite_color)
```

Blue and green each appear twice. Order restored.


## Cleaning Product Names

Pulling structured data out of messy product descriptions --- you will do this more often than you think:

```{r}
products <- tibble(
  description = c(
    "Widget A (SKU: WDG-001) - Price: $29.99",
    "Gadget B (SKU: GDG-042) - Price: $149.50",
    "Thingamajig (SKU: THG-100) - Price: $5.00"
  )
)

products %>%
  mutate(
    sku = str_extract(description, "[A-Z]{3}-\\d{3}"),
    price_raw = str_extract(description, "\\$[\\d.]+"),
    price = as.numeric(str_remove(price_raw, "\\$")),
    product_name = str_extract(description, "^[^(]+") %>% str_trim()
  ) %>%
  select(product_name, sku, price)
```

One messy text column became three clean, analysis-ready columns.


## Extracting Numbers from Sales Notes

Pulling numbers out of free-text notes so you can actually do math with them:

```{r}
sales_notes <- tibble(
  region = c("North", "North", "South", "South"),
  notes = c("Sold 15 units of product A",
            "Sold 8 units of product B",
            "Sold 22 units of product A",
            "Sold 3 units of product C")
)

sales_notes %>%
  mutate(
    units = as.integer(str_extract(notes, "\\d+")),
    product = str_extract(notes, "product [A-C]")
  ) %>%
  group_by(region, product) %>%
  summarize(total_units = sum(units), .groups = "drop")
```

Text data became a proper summary table. Put that in the quarterly deck.


## Quick Reference

The functions you will actually use:

| What you want to do | Function |
|:---------------------|:---------|
| Glue strings together | `str_c()` |
| Lowercase / uppercase / title case | `str_to_lower()`, `str_to_upper()`, `str_to_title()` |
| Remove extra whitespace | `str_trim()`, `str_squish()` |
| Does it contain this pattern? | `str_detect()` |
| Pull out matching text | `str_extract()` |
| Find and replace | `str_replace()`, `str_replace_all()` |
| Delete a pattern | `str_remove()`, `str_remove_all()` |
| How many characters? | `str_length()` |

And the three regex patterns that handle most problems:

| Pattern | What it finds | Example |
|:--------|:-------------|:--------|
| `\\d+` | Numbers | "Order **1234** placed" |
| `[A-Za-z]+` | Words (letters) | "**SKU**-12345" |
| `[\\w.]+@[\\w.]+` | Email addresses | "**alice@example.com**" |

That is stringr. Clean your text, extract what you need, get back to the analysis that actually matters.
