# Working with Strings {#stringr}

```{r include=FALSE}
library(tidyverse)
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

## Introduction

Much of the data we encounter in practice is not purely numeric. Survey responses, product reviews, addresses, names, dates stored as text, log files, and social media posts are all examples of string (character) data. Being able to manipulate and clean strings is therefore a critical skill for any data analyst.

Base R provides several functions for working with strings --- `grep()`, `gsub()`, `substr()`, `nchar()`, `paste()`, and others --- but the naming conventions are inconsistent and the argument orders vary from function to function. The **stringr** package, part of the tidyverse, solves this problem by providing a cohesive set of functions that all begin with the `str_` prefix. Every stringr function takes the string as its first argument, making it natural to use with the pipe operator (`%>%` or `|>`).

Under the hood, stringr is built on top of the **stringi** package, which provides a comprehensive and highly optimized set of string processing functions based on the ICU (International Components for Unicode) library. This means stringr gives you a clean, user-friendly interface backed by a powerful engine that handles international text and edge cases correctly.

Let us get started.

## String Basics

### Creating Strings

In R, strings are created by surrounding text with either double quotes or single quotes. Double quotes are the preferred convention.

```{r}
string1 <- "This is a string"
string2 <- 'This is also a string'
string1
string2
```

If your string itself contains a double quote, you can use single quotes on the outside, or escape the inner double quote with a backslash.

```{r}
string3 <- 'She said, "Hello!"'
string4 <- "She said, \"Hello!\""
string3
string4
```

### Special Characters

Certain characters have special meaning when preceded by a backslash. These are called escape sequences.

- `\n` --- newline
- `\t` --- tab
- `\\` --- a literal backslash
- `\"` --- a literal double quote

To see the actual content of a string (with the escape sequences rendered), use `writeLines()`.

```{r}
example <- "First line\nSecond line\tTabbed"
writeLines(example)
```

### String Length

The `str_length()` function returns the number of characters in each string.

```{r}
fruits <- c("apple", "banana", "kiwi", "strawberry", "fig")
str_length(fruits)
```

Note that `str_length()` is vectorized --- it works on an entire vector at once. Compare this with base R's `nchar()`, which does the same thing but does not follow the `str_` naming convention.

```{r}
str_length(c("", " ", NA))
```

An empty string has length 0, a string containing a single space has length 1, and `NA` returns `NA`.

### Combining Strings with `str_c()`

The `str_c()` function concatenates strings. It is the stringr equivalent of `paste()` and `paste0()`.

```{r}
str_c("Hello", "World")
str_c("Hello", "World", sep = " ")
str_c("Hello", "World", sep = ", ")
```

Like other stringr functions, `str_c()` is vectorized.

```{r}
names <- c("Alice", "Bob", "Charlie")
str_c("Hello, ", names, "!")
```

The `collapse` argument is used to combine a vector of strings into a single string.

```{r}
str_c(names, collapse = ", ")
str_c(names, collapse = " and ")
```

You can use both `sep` and `collapse` together. The `sep` is applied first (element-wise), and then `collapse` reduces the result to a single string.

```{r}
prefixes <- c("Dr.", "Mr.", "Ms.")
str_c(prefixes, names, sep = " ", collapse = "; ")
```

### Extracting Substrings with `str_sub()`

The `str_sub()` function extracts or replaces substrings by position. It takes `start` and `end` arguments.

```{r}
x <- "Strawberry"
str_sub(x, 1, 5)
str_sub(x, 6, 10)
```

Negative indices count from the end of the string.

```{r}
str_sub(x, -5, -1)
```

`str_sub()` is also vectorized and can be used on the left-hand side of an assignment to replace substrings.

```{r}
fruits_copy <- fruits
str_sub(fruits_copy, 1, 1) <- str_to_upper(str_sub(fruits_copy, 1, 1))
fruits_copy
```

## Detecting Patterns

One of the most common tasks with strings is checking whether they contain a particular pattern. stringr provides several functions for pattern detection.

### `str_detect()` --- Logical Matching

`str_detect()` returns a logical vector indicating whether each string matches the given pattern.

```{r}
colors <- c("red", "green", "blue", "redwood", "forest green", "navy blue")
str_detect(colors, "red")
str_detect(colors, "green")
```

This is especially useful for filtering data frames.

```{r}
color_df <- tibble(color = colors, id = 1:6)
color_df %>% filter(str_detect(color, "blue"))
```

### `str_which()` --- Index Positions

`str_which()` returns the integer indices of strings that match the pattern, similar to base R's `which()`.

```{r}
str_which(colors, "green")
```

### `str_count()` --- Counting Matches

`str_count()` counts the number of times a pattern appears in each string.

```{r}
sentences <- c("the cat sat on the mat",
               "the dog chased the cat",
               "no animals here")
str_count(sentences, "the")
str_count(sentences, "cat")
```

### `str_locate()` and `str_locate_all()`

These functions return the start and end positions of pattern matches. `str_locate()` returns the position of the first match; `str_locate_all()` returns all matches.

```{r}
str_locate("the cat sat on the mat", "the")
str_locate_all("the cat sat on the mat", "the")
```

## Extracting and Replacing

### Extracting Matches

`str_extract()` returns the first match of a pattern in each string, while `str_extract_all()` returns all matches.

```{r}
mixed_text <- c("Order 1234 placed", "Invoice 5678 pending", "No number here", "Items 91011 and 1213")
str_extract(mixed_text, "\\d+")
str_extract_all(mixed_text, "\\d+")
```

The pattern `\\d+` matches one or more digits. We will cover regular expressions in more detail later in this chapter.

### Replacing Patterns

`str_replace()` replaces the first match of a pattern, while `str_replace_all()` replaces all matches.

```{r}
messy <- c("hello   world", "  too   many   spaces  ")
str_replace(messy, "\\s+", " ")
str_replace_all(messy, "\\s+", " ")
```

Here, `\\s+` matches one or more whitespace characters. Notice that `str_replace()` only fixes the first occurrence, while `str_replace_all()` fixes all of them.

### Removing Patterns

`str_remove()` and `str_remove_all()` are convenient shortcuts for replacing with an empty string.

```{r}
phone_numbers <- c("(555) 123-4567", "(800) 555-0199", "(312) 867-5309")
# Remove non-digit characters to get clean phone numbers
str_remove_all(phone_numbers, "[^0-9]")
```

Here is a practical example of cleaning phone numbers into a standardized format.

```{r}
clean_phones <- phone_numbers %>%
  str_remove_all("[^0-9]") %>%
  str_replace("(\\d{3})(\\d{3})(\\d{4})", "\\1-\\2-\\3")
clean_phones
```

### Extracting Dates from Text

```{r}
log_entries <- c("Error on 2024-01-15: file not found",
                 "Warning on 2024-02-20: low memory",
                 "Info on 2024-03-05: process started")
str_extract(log_entries, "\\d{4}-\\d{2}-\\d{2}")
```

## Modifying Strings

### Changing Case

stringr provides four case-conversion functions.

```{r}
sample_text <- "the QUICK brown FOX"
str_to_lower(sample_text)
str_to_upper(sample_text)
str_to_title(sample_text)
str_to_sentence(sample_text)
```

These are useful for standardizing data. For example, ensuring all city names follow title case.

```{r}
cities <- c("new york", "LOS ANGELES", "chicago", "SAN FRANCISCO")
str_to_title(cities)
```

### Trimming and Squishing Whitespace

`str_trim()` removes whitespace from the beginning and end of a string. `str_squish()` additionally reduces internal whitespace to single spaces.

```{r}
padded <- "   too  much   space   "
str_trim(padded)
str_squish(padded)
```

`str_trim()` has a `side` argument that can be set to `"left"`, `"right"`, or `"both"` (the default).

```{r}
str_trim(padded, side = "left")
str_trim(padded, side = "right")
```

### Padding Strings

`str_pad()` adds characters (spaces by default) to a string to bring it to a specified width. This is useful for creating fixed-width output or aligning text.

```{r}
ids <- c("1", "23", "456", "7890")
str_pad(ids, width = 5, side = "left", pad = "0")
```

### Wrapping and Truncating

`str_wrap()` wraps a long string to a specified width, inserting newlines. `str_trunc()` truncates a string to a specified maximum width, adding an ellipsis.

```{r}
long_text <- "This is a very long sentence that we might want to wrap to a certain width so that it displays nicely in a narrow column or output window."
cat(str_wrap(long_text, width = 40))
```

```{r}
str_trunc(long_text, width = 50)
str_trunc(long_text, width = 50, side = "left")
str_trunc(long_text, width = 50, side = "center")
```

## Splitting and Combining

### Splitting Strings

`str_split()` splits a string into pieces based on a pattern and returns a list.

```{r}
csv_line <- "Alice,30,New York"
str_split(csv_line, ",")
```

When working with a vector of strings, `str_split()` returns a list with one element per input string.

```{r}
csv_lines <- c("Alice,30,New York", "Bob,25,Chicago", "Charlie,35,Denver")
str_split(csv_lines, ",")
```

`str_split_fixed()` returns a character matrix with a fixed number of columns, which is often more convenient.

```{r}
str_split_fixed(csv_lines, ",", n = 3)
```

You can also limit the number of splits with the `n` argument in `str_split()`.

```{r}
str_split("one-two-three-four", "-", n = 2)
```

### Flattening with `str_flatten()`

`str_flatten()` collapses a character vector into a single string. It is similar to using `str_c()` with the `collapse` argument.

```{r}
words <- c("R", "is", "great", "for", "data", "analysis")
str_flatten(words, collapse = " ")
str_flatten(words, collapse = "-")
```

`str_flatten_comma()` is a convenient shortcut for comma-separated lists.

```{r}
items <- c("apples", "bananas", "cherries")
str_flatten_comma(items)
str_flatten_comma(items, last = ", and ")
```

## Regular Expressions Basics

Regular expressions (regex) are a powerful language for describing patterns in strings. Most stringr functions accept regular expressions as their pattern argument. While a full treatment of regular expressions would fill a book on its own, here we cover the most useful building blocks.

### Literal Characters

The simplest patterns are literal strings. When you search for `"cat"`, it matches the exact sequence of characters c-a-t.

```{r}
animals <- c("cat", "concatenate", "scat", "dog", "caterpillar")
str_detect(animals, "cat")
str_extract(animals, "cat")
```

### Anchors

Anchors specify where in the string the match must occur.

- `^` matches the start of the string
- `$` matches the end of the string

```{r}
str_detect(animals, "^cat")
str_detect(animals, "cat$")
```

Using `^cat` matches only strings that begin with "cat" (cat and caterpillar). Using `cat$` matches only strings that end with "cat" (cat and scat).

### Character Classes

Character classes let you match any one of a set of characters.

- `[abc]` --- matches a, b, or c
- `[a-z]` --- matches any lowercase letter
- `[A-Z]` --- matches any uppercase letter
- `[0-9]` --- matches any digit
- `[^abc]` --- matches anything except a, b, or c

There are also shorthand character classes (note the double backslash in R strings):

- `\\d` --- any digit (equivalent to `[0-9]`)
- `\\D` --- any non-digit
- `\\w` --- any word character (letters, digits, underscore)
- `\\W` --- any non-word character
- `\\s` --- any whitespace (space, tab, newline)
- `\\S` --- any non-whitespace character

```{r}
test_strings <- c("abc123", "def456", "ghi", "789", "ab1cd2")
str_extract(test_strings, "[0-9]+")
str_extract(test_strings, "\\d+")
str_extract_all(test_strings, "[a-z]+")
```

### Quantifiers

Quantifiers specify how many times a pattern should be matched.

- `*` --- zero or more times
- `+` --- one or more times
- `?` --- zero or one time (optional)
- `{n}` --- exactly n times
- `{n,}` --- n or more times
- `{n,m}` --- between n and m times

```{r}
codes <- c("A1", "AB12", "ABC123", "ABCD1234", "A")
str_extract(codes, "[A-Z]{2,3}")
str_extract(codes, "[A-Z]+\\d+")
str_extract(codes, "[A-Z]{1,2}\\d{1,2}")
```

```{r}
# Matching optional characters
spellings <- c("color", "colour", "colored", "coloured")
str_detect(spellings, "colou?r")
```

### Alternation

The `|` operator acts as an OR, matching either the pattern on its left or the pattern on its right.

```{r}
pets <- c("I have a cat", "I have a dog", "I have a bird", "I have a cat and a dog")
str_extract(pets, "cat|dog")
str_extract_all(pets, "cat|dog")
```

### Grouping with Parentheses

Parentheses group sub-patterns together and create capturing groups that can be referenced in replacements.

```{r}
dates <- c("2024-01-15", "2024-02-20", "2024-12-25")
# Rearrange from YYYY-MM-DD to MM/DD/YYYY
str_replace(dates, "(\\d{4})-(\\d{2})-(\\d{2})", "\\2/\\3/\\1")
```

Here, `\\1`, `\\2`, and `\\3` are backreferences to the first, second, and third captured groups, respectively.

```{r}
# Extract just the year and month
str_extract(dates, "\\d{4}-\\d{2}")
```

### Practical Regex Examples

Here are a few common patterns you may find useful.

```{r}
# Email-like patterns
texts <- c("Contact me at alice@example.com for details",
           "Send to bob.smith@company.org or call",
           "No email here")
str_extract(texts, "[\\w.]+@[\\w.]+\\.[a-z]+")
```

```{r}
# US zip codes (5 digits, optionally followed by dash and 4 digits)
addresses <- c("New York, NY 10001", "Chicago, IL 60614-2345", "Denver, CO 80202")
str_extract(addresses, "\\d{5}(-\\d{4})?")
```

## Sorting Strings

### `str_sort()` and `str_order()`

`str_sort()` returns the sorted strings, while `str_order()` returns the integer indices that would sort them.

```{r}
fruits_unsorted <- c("banana", "apple", "cherry", "date", "elderberry")
str_sort(fruits_unsorted)
str_order(fruits_unsorted)
```

Both functions accept a `decreasing` argument.

```{r}
str_sort(fruits_unsorted, decreasing = TRUE)
```

They also handle locale-aware sorting with the `locale` argument, which can be important for non-English text.

```{r}
# Numeric sorting: compare default vs. numeric
versions <- c("version2", "version10", "version1", "version20", "version3")
str_sort(versions)
str_sort(versions, numeric = TRUE)
```

The `numeric = TRUE` argument ensures that numbers embedded in strings are sorted numerically rather than lexicographically.

## Practical Examples

Let us put everything together with some realistic data cleaning scenarios.

### Standardizing Names

Messy name data is one of the most common problems in practice. Here is a data frame with inconsistent formatting.

```{r}
messy_names <- tibble(
  raw_name = c("  john DOE ", "JANE   smith", "  Bob  JOHNSON  ",
               "alice   Williams", "CHARLIE  brown ", " diana PRINCE")
)

clean_names <- messy_names %>%
  mutate(
    clean_name = raw_name %>%
      str_squish() %>%
      str_to_title()
  )
clean_names
```

### Extracting Information from Unstructured Text

Suppose we have product descriptions and need to extract prices and product codes.

```{r}
products <- tibble(
  description = c(
    "Widget A (SKU: WDG-001) - Price: $29.99",
    "Gadget B (SKU: GDG-042) - Price: $149.50",
    "Thingamajig (SKU: THG-100) - Price: $5.00",
    "Doohickey XL (SKU: DHK-007) - Price: $89.95",
    "Whatchamacallit (SKU: WCL-999) - Price: $12.49"
  )
)

products_parsed <- products %>%
  mutate(
    sku = str_extract(description, "SKU: ([A-Z]{3}-\\d{3})", group = 1),
    price_text = str_extract(description, "\\$[\\d.]+"),
    price = as.numeric(str_remove(price_text, "\\$")),
    product_name = str_extract(description, "^[^(]+") %>% str_trim()
  )
products_parsed %>% select(product_name, sku, price)
```

### Parsing Addresses

Addresses are notoriously messy. Here is an example of extracting components from semi-structured address strings.

```{r}
addresses_df <- tibble(
  full_address = c(
    "123 Main St, Springfield, IL 62704",
    "456 Oak Ave, Suite 200, Chicago, IL 60614",
    "789 Pine Rd, Denver, CO 80202",
    "1010 Elm Blvd, Apt 3B, Portland, OR 97201",
    "222 Maple Dr, Seattle, WA 98101"
  )
)

addresses_parsed <- addresses_df %>%
  mutate(
    state = str_extract(full_address, "[A-Z]{2}(?=\\s+\\d{5})"),
    zip = str_extract(full_address, "\\d{5}$"),
    # Extract city: the word(s) before ", STATE ZIP"
    city = str_extract(full_address, "[A-Za-z ]+(?=,\\s+[A-Z]{2}\\s+\\d{5})") %>%
      str_trim()
  )
addresses_parsed %>% select(city, state, zip)
```

### A Complete Data Cleaning Pipeline

Here is a more comprehensive example that demonstrates how stringr integrates with dplyr in a typical data cleaning workflow. Imagine we have received survey data with various formatting issues.

```{r}
survey_raw <- tibble(
  respondent = c("  JOHN doe  ", "jane SMITH", " Bob  Johnson ",
                 "ALICE williams", " charlie Brown"),
  email = c("John.Doe@Gmail.COM", "jane_smith@YAHOO.com",
            "BOB@company.ORG", "alice.w@School.Edu", "charlie@EMAIL.com"),
  phone = c("(555) 123-4567", "555.234.5678", "555 345 6789",
            "5554567890", "(555)-567-8901"),
  favorite_color = c("  Blue ", "RED", "  green  ", "BLUE", " Green"),
  comments = c("Great survey!  Thanks!  ",
               "   Needs   more   questions   ",
               "N/A",
               "",
               "  Good job.   Very   thorough.  ")
)

survey_clean <- survey_raw %>%
  mutate(
    # Clean names
    respondent = respondent %>% str_squish() %>% str_to_title(),

    # Standardize emails to lowercase
    email = str_to_lower(email),

    # Standardize phone numbers to XXX-XXX-XXXX
    phone = phone %>%
      str_remove_all("[^0-9]") %>%
      str_replace("(\\d{3})(\\d{3})(\\d{4})", "\\1-\\2-\\3"),

    # Standardize colors
    favorite_color = favorite_color %>% str_squish() %>% str_to_lower(),

    # Clean up comments
    comments = comments %>%
      str_squish() %>%
      na_if("") %>%
      na_if("N/A")
  )

survey_clean
```

Let us verify the cleaned data looks correct by examining each column.

```{r}
# Check distinct colors after standardization
survey_clean %>% count(favorite_color)
```

```{r}
# Verify phone format
survey_clean %>% select(respondent, phone)
```

```{r}
# Check which respondents left meaningful comments
survey_clean %>%
  filter(!is.na(comments)) %>%
  select(respondent, comments)
```

### Working with stringr in Grouped Operations

stringr functions combine naturally with grouped dplyr operations.

```{r}
sales_notes <- tibble(
  region = c("North", "North", "South", "South", "East", "East", "West", "West"),
  notes = c("Sold 15 units of product A",
            "Sold 8 units of product B",
            "Sold 22 units of product A",
            "Sold 3 units of product C",
            "Sold 11 units of product B",
            "Sold 7 units of product A",
            "Sold 19 units of product C",
            "Sold 30 units of product A")
)

sales_summary <- sales_notes %>%
  mutate(
    units = as.integer(str_extract(notes, "\\d+")),
    product = str_extract(notes, "product [A-C]")
  ) %>%
  group_by(region, product) %>%
  summarize(total_units = sum(units), .groups = "drop") %>%
  arrange(product, desc(total_units))

sales_summary
```

This example demonstrates how extracting structured information from text fields enables us to perform quantitative analysis on data that originally existed only as free-form notes.

## Summary

The stringr package provides a consistent, expressive, and pipe-friendly toolkit for string manipulation in R. The key functions covered in this chapter include:

| Task | Functions |
|------|-----------|
| Measure length | `str_length()` |
| Combine | `str_c()`, `str_flatten()` |
| Extract substrings | `str_sub()` |
| Detect patterns | `str_detect()`, `str_which()`, `str_count()`, `str_locate()` |
| Extract matches | `str_extract()`, `str_extract_all()` |
| Replace / remove | `str_replace()`, `str_replace_all()`, `str_remove()`, `str_remove_all()` |
| Change case | `str_to_lower()`, `str_to_upper()`, `str_to_title()`, `str_to_sentence()` |
| Whitespace | `str_trim()`, `str_squish()`, `str_pad()` |
| Wrap / truncate | `str_wrap()`, `str_trunc()` |
| Split | `str_split()`, `str_split_fixed()` |
| Sort | `str_sort()`, `str_order()` |

Combined with a basic understanding of regular expressions --- anchors, character classes, quantifiers, alternation, and grouping --- these functions equip you to handle the vast majority of string manipulation tasks that arise in data analysis.
