# Wrangling Categories with forcats {#forcats}

```{r include=FALSE}
library(tidyverse)
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

## Factors = Categories. That's It.

If you've ever made a bar chart in R and wondered why the bars are in alphabetical order instead of something useful --- congratulations, you've discovered why factors exist.

A **factor** is R's way of handling categorical data: things like survey responses ("Agree", "Neutral", "Disagree"), product categories, regions, departments, or customer segments. Under the hood, a factor is a character variable with a defined set of allowed values (called **levels**) in a specific order.

Why should you care about the order? Because **ggplot2 uses factor order to arrange your charts**. Alphabetical bars are almost never what you want in a business presentation. The **forcats** package (part of the tidyverse, loaded automatically) gives you simple tools to control that order.

Let's see the problem first.


## The Problem: Alphabetical Everything

```{r}
# A character vector
satisfaction <- c("Low", "Medium", "High", "Low", "High")

# When you make it a factor, R defaults to alphabetical order
factor(satisfaction)
```

The levels are "High", "Low", "Medium" --- alphabetical, not logical. You want "Low", "Medium", "High". That's what factor manipulation is all about.

```{r}
# Fix it by specifying levels
factor(satisfaction, levels = c("Low", "Medium", "High"))
```

The forcats package makes this kind of thing much easier, especially when you're working with real data that has dozens of categories.


## The Dataset: General Social Survey

We'll use `gss_cat`, a dataset built into forcats with survey data from the General Social Survey. It has several factor columns --- perfect for practice.

```{r}
glimpse(gss_cat)
```


## `fct_reorder()` --- THE Most Important Function

If you learn one thing from this chapter, make it `fct_reorder()`. It reorders a factor based on the values of another variable. Translation: **it sorts your bar chart bars**.

Here's the before and after. First, without reordering:

```{r fig.cap="Religion by TV hours, unsorted. Good luck finding patterns here."}
relig_summary <- gss_cat %>%
  group_by(relig) %>%
  summarize(
    avg_tv = mean(tvhours, na.rm = TRUE),
    n = n()
  )

ggplot(relig_summary, aes(x = avg_tv, y = relig)) +
  geom_point() +
  labs(title = "TV Hours by Religion (unsorted chaos)",
       x = "Mean TV hours per day", y = NULL) +
  theme_minimal()
```

Now with `fct_reorder()`:

```{r fig.cap="Same data, but now sorted by value. The story jumps right out."}
ggplot(relig_summary, aes(x = avg_tv, y = fct_reorder(relig, avg_tv))) +
  geom_point() +
  labs(title = "TV Hours by Religion (sorted --- much better)",
       x = "Mean TV hours per day", y = NULL) +
  theme_minimal()
```

See the difference? Same data, but the sorted version tells a story instantly. This is the difference between a chart your manager glances at and one they actually understand.

Here's another example with a bar chart:

```{r fig.cap="Mean petal length by iris species, sorted by value."}
iris %>%
  group_by(Species) %>%
  summarize(mean_petal = mean(Petal.Length)) %>%
  ggplot(aes(x = mean_petal, y = fct_reorder(Species, mean_petal))) +
  geom_col(fill = "steelblue") +
  labs(title = "Mean Petal Length by Species", x = "Mean Petal Length", y = NULL) +
  theme_minimal()
```

**Rule of thumb:** If you're making a bar chart or dot plot with categories, use `fct_reorder()`. Always. Your audience will thank you.


## `fct_infreq()` --- Sort by Count

Sometimes you just want bars sorted by how many observations are in each category --- most common first. That's `fct_infreq()`.

```{r fig.cap="Marital status bars sorted by frequency. Most common categories jump out immediately."}
gss_cat %>%
  mutate(marital = fct_infreq(marital)) %>%
  ggplot(aes(x = marital)) +
  geom_bar(fill = "steelblue") +
  labs(title = "Marital Status (sorted by frequency)", x = NULL, y = "Count") +
  theme_minimal()
```

Combine it with `fct_rev()` for a horizontal bar chart that reads top-to-bottom:

```{r fig.cap="Horizontal bars with the most common status at the top."}
gss_cat %>%
  mutate(marital = fct_rev(fct_infreq(marital))) %>%
  ggplot(aes(y = marital)) +
  geom_bar(fill = "darkorange") +
  labs(title = "Marital Status (highest count at top)", x = "Count", y = NULL) +
  theme_minimal()
```


## `fct_relevel()` --- Manual Ordering

Sometimes you need a specific order that isn't alphabetical or by frequency. Maybe you want "Never married" first because that's your reference group, or you need income brackets from low to high.

`fct_relevel()` lets you move specific levels to wherever you want.

```{r}
# Default level order
levels(gss_cat$marital)
```

```{r}
# Move "Never married" to the front
gss_cat %>%
  mutate(marital = fct_relevel(marital, "Never married")) %>%
  pull(marital) %>%
  levels()
```

```{r}
# Move "Widowed" to the end
gss_cat %>%
  mutate(marital = fct_relevel(marital, "Widowed", after = Inf)) %>%
  pull(marital) %>%
  levels()
```

And the chart payoff:

```{r fig.cap="Marital status in a custom order that tells a life-stage story."}
gss_cat %>%
  mutate(marital = fct_relevel(marital,
    "Never married", "Married", "Separated",
    "Divorced", "Widowed", "No answer")) %>%
  ggplot(aes(x = marital)) +
  geom_bar(fill = "steelblue") +
  labs(title = "Marital Status (life-stage order)", x = NULL, y = "Count") +
  theme_minimal()
```

Now the bars tell a life-stage story: never married, married, separated, divorced, widowed. Much more intuitive than alphabetical.


## `fct_recode()` --- Rename Your Categories

Survey data often comes with cryptic or inconsistent labels. `fct_recode()` lets you rename levels with a simple `"new name" = "old name"` syntax.

```{r}
gss_cat %>%
  mutate(partyid = fct_recode(partyid,
    "Republican (strong)"    = "Strong republican",
    "Republican (weak)"      = "Not str republican",
    "Independent (near rep)" = "Ind,near rep",
    "Independent (near dem)" = "Ind,near dem",
    "Democrat (weak)"        = "Not str democrat",
    "Democrat (strong)"      = "Strong democrat"
  )) %>%
  count(partyid)
```

This is also great when your data has labels like "1", "2", "3" that should really say "Strongly Disagree", "Neutral", "Strongly Agree".


## `fct_collapse()` --- Merge Categories Together

Sometimes you have too many categories and need to combine them. `fct_collapse()` merges multiple levels into new groups.

```{r}
gss_cat %>%
  mutate(partyid = fct_collapse(partyid,
    "Republican"  = c("Strong republican", "Not str republican"),
    "Independent" = c("Ind,near rep", "Independent", "Ind,near dem"),
    "Democrat"    = c("Not str democrat", "Strong democrat"),
    "Other"       = c("No answer", "Don't know", "Other party")
  )) %>%
  count(partyid)
```

And the chart payoff --- a clean, simple political affiliation plot:

```{r fig.cap="Political affiliation after collapsing into four clean groups."}
gss_cat %>%
  mutate(party = fct_collapse(partyid,
    "Republican"  = c("Strong republican", "Not str republican"),
    "Independent" = c("Ind,near rep", "Independent", "Ind,near dem"),
    "Democrat"    = c("Not str democrat", "Strong democrat"),
    "Other"       = c("No answer", "Don't know", "Other party")
  )) %>%
  mutate(party = fct_infreq(party)) %>%
  ggplot(aes(x = party, fill = party)) +
  geom_bar(show.legend = FALSE) +
  scale_fill_manual(values = c(
    "Democrat" = "blue", "Independent" = "gray50",
    "Republican" = "red", "Other" = "gray80"
  )) +
  labs(title = "Political Affiliation (simplified)", x = NULL, y = "Count") +
  theme_minimal()
```

Went from 10 confusing categories to 4 clear ones. Your stakeholders will actually understand this chart.


## `fct_lump()` --- "Show Me the Top N, Lump the Rest"

This is a lifesaver when you have a variable with 30+ categories. Nobody wants a bar chart with 30 bars. `fct_lump_n()` keeps the top N categories and lumps everything else into "Other."

```{r}
# Keep the 4 most common religions, lump the rest
gss_cat %>%
  mutate(relig = fct_lump_n(relig, n = 4)) %>%
  count(relig, sort = TRUE)
```

The chart version --- before and after lumping:

```{r fig.cap="All denominations shown. Way too many categories to be useful."}
gss_cat %>%
  count(denom) %>%
  ggplot(aes(x = n, y = denom)) +
  geom_point() +
  labs(title = "All Denominations (information overload)", x = "Count", y = NULL) +
  theme_minimal()
```

```{r fig.cap="Top 8 denominations with the rest lumped into Other. Now this is readable."}
gss_cat %>%
  mutate(denom = fct_lump_n(denom, n = 8)) %>%
  count(denom) %>%
  mutate(denom = fct_reorder(denom, n)) %>%
  ggplot(aes(x = n, y = denom)) +
  geom_point(size = 3, color = "steelblue") +
  labs(title = "Top 8 Denominations (rest lumped into Other)",
       x = "Count", y = NULL) +
  theme_minimal()
```

There are a few variants depending on how you want to decide what gets lumped:

- `fct_lump_n(x, n = 5)` --- keep top 5
- `fct_lump_prop(x, prop = 0.10)` --- keep categories representing at least 10% of data
- `fct_lump_min(x, min = 100)` --- keep categories with at least 100 observations

Pick the one that matches your question.


## Bonus: Making mtcars Charts Better with Factors

The `mtcars` dataset has variables like `cyl` (cylinders) and `am` (transmission) coded as numbers. They're really categories. Converting them to factors with meaningful labels instantly makes your charts more professional.

```{r fig.cap="Mean MPG by cylinders and transmission. Labeled factors make this chart presentation-ready."}
mtcars %>%
  mutate(
    cyl = factor(cyl),
    transmission = factor(am, levels = c(0, 1), labels = c("Automatic", "Manual"))
  ) %>%
  group_by(cyl, transmission) %>%
  summarize(mean_mpg = mean(mpg), .groups = "drop") %>%
  ggplot(aes(x = cyl, y = mean_mpg, fill = transmission)) +
  geom_col(position = "dodge") +
  scale_fill_brewer(palette = "Set2") +
  labs(title = "Mean MPG by Cylinders and Transmission",
       x = "Number of Cylinders", y = "Mean MPG", fill = "Transmission") +
  theme_minimal()
```

Without factors, those bars would just say "0" and "1" for transmission. With factors, they say "Automatic" and "Manual." Small change, big impact.


## `fct_reorder2()` --- Aligning Legends with Lines

Quick bonus for line charts. When you have multiple lines, the legend should match the visual order of the lines at the right edge of the plot. `fct_reorder2()` does this automatically.

```{r fig.cap="Line chart where the legend order matches the line positions at the right edge."}
gss_cat %>%
  filter(!is.na(age)) %>%
  mutate(marital = fct_lump_n(marital, n = 4)) %>%
  count(age, marital) %>%
  group_by(age) %>%
  mutate(prop = n / sum(n)) %>%
  ungroup() %>%
  ggplot(aes(x = age, y = prop, color = fct_reorder2(marital, age, prop))) +
  geom_line(linewidth = 1) +
  scale_color_brewer(palette = "Set1") +
  labs(title = "Marital Status by Age",
       x = "Age", y = "Proportion", color = "Status") +
  theme_minimal()
```


## Quick Reference

Here's your factor toolkit at a glance:

| What you want to do | Function | When to use it |
|:---------------------|:---------|:---------------|
| Sort bars by a value | `fct_reorder()` | Almost every bar/dot chart |
| Sort bars by count | `fct_infreq()` | "Show me what's most common" |
| Reverse the order | `fct_rev()` | Horizontal bar charts (top to bottom) |
| Set a custom order | `fct_relevel()` | Life stages, survey scales, custom sorts |
| Rename categories | `fct_recode()` | Cryptic labels need fixing |
| Merge categories | `fct_collapse()` | Too many groups, need to simplify |
| Keep top N, lump rest | `fct_lump_n()` | 30 categories? Nobody wants that chart |

The single most impactful habit from this chapter: **always use `fct_reorder()` or `fct_infreq()` in your ggplot2 charts.** Alphabetical ordering is almost never the most informative ordering, and a well-ordered factor is the difference between a confusing chart and one that tells a clear story.
