# Working with Factors {#forcats}

```{r include=FALSE}
library(tidyverse)
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

Categorical data is everywhere in real-world analysis: survey responses, product categories, geographic regions, education levels, and Likert-scale ratings. In R, the factor data type is purpose-built for representing categorical variables. Factors look like character strings on the surface, but underneath they carry a fixed set of allowed values --- called **levels** --- and an explicit ordering of those levels. The **forcats** package, a core member of the tidyverse, provides a comprehensive toolkit for creating, inspecting, reordering, and modifying factors. This chapter covers forcats in depth, with extensive examples showing how factor manipulation directly improves data analysis and visualization.

The forcats package is loaded automatically when you call `library(tidyverse)`, so you do not need to load it separately.


## What Are Factors?

A **factor** is R's data type for categorical variables. At its core, a factor is an integer vector with two additional attributes: a set of **levels** (the unique categories the variable can take) and an optional ordering. When you print a factor, R shows the character labels, but internally it stores integers that map to those labels.

### Character vs. Factor

Character vectors and factors can look identical when printed, but they behave differently in important ways.

```{r}
# A character vector
char_vec <- c("low", "medium", "high", "low", "high")
class(char_vec)
```

```{r}
# A factor
fac_vec <- factor(c("low", "medium", "high", "low", "high"))
class(fac_vec)
```

```{r}
# The factor has levels
levels(fac_vec)
```

Notice that the levels default to alphabetical order: "high", "low", "medium". This is rarely the order we want --- we typically want "low", "medium", "high". Controlling level order is one of the primary reasons to use factors.

### Why Use Factors?

There are three major reasons to use factors rather than plain character vectors:

1. **Controlling order in plots.** When ggplot2 encounters a character variable, it orders categories alphabetically. With a factor, you control the exact order in which categories appear on an axis or in a legend. This is critical for creating clear, well-organized visualizations.

2. **Statistical modeling.** Many modeling functions in R (such as `lm()` and `glm()`) treat factors specially, automatically creating dummy variables and using the first level as the reference category. Controlling levels lets you choose which category serves as the reference.

3. **Data validation.** Factors remember all possible levels, even if some are not present in the current data. This is useful for ensuring that summary tables include all expected categories, including those with zero counts.

### Factor Basics

The `factor()` function is the most basic way to create a factor. The `levels()` function retrieves the levels, and `nlevels()` returns the count of levels.

```{r}
# Create a factor with explicit level order
rating <- factor(
  c("good", "excellent", "poor", "good", "excellent"),
  levels = c("poor", "good", "excellent")
)
rating
```

```{r}
levels(rating)
```

```{r}
nlevels(rating)
```


## Creating Factors

### Using factor()

The `factor()` function is R's built-in tool for creating factors. The `levels` argument controls the order, and the `labels` argument optionally renames the levels.

```{r}
# Without specifying levels, order is alphabetical
sizes <- factor(c("S", "M", "L", "XL", "M", "S"))
levels(sizes)
```

```{r}
# Specifying levels sets a meaningful order
sizes_ordered <- factor(
  c("S", "M", "L", "XL", "M", "S"),
  levels = c("S", "M", "L", "XL")
)
levels(sizes_ordered)
```

```{r}
# Using labels to rename levels
gear_factor <- factor(
  mtcars$gear,
  levels = c(3, 4, 5),
  labels = c("Three", "Four", "Five")
)
table(gear_factor)
```

### factor() vs. as.factor()

The `factor()` function gives you full control over levels and labels. The `as.factor()` function is a simpler conversion that assigns alphabetical levels. Use `factor()` when you need to control level order; use `as.factor()` for quick conversions.

```{r}
# as.factor() uses alphabetical ordering
as.factor(c("banana", "apple", "cherry"))
```

```{r}
# factor() with explicit levels
factor(c("banana", "apple", "cherry"), levels = c("apple", "banana", "cherry"))
```

### Using parse_factor() from readr

The `parse_factor()` function from the readr package is stricter than `factor()`. It generates a warning when data contains values not included in the specified levels, which helps catch data entry errors.

```{r}
# parse_factor warns about values not in the levels
parse_factor(
  c("low", "medium", "high", "unknown"),
  levels = c("low", "medium", "high")
)
```

### Factors Remember All Levels

An important property of factors is that they remember all possible levels, even if some levels are not present in the current data. This is useful for maintaining consistency across subsets.

```{r}
# Create a factor with all levels defined
all_colors <- factor(
  c("red", "blue"),
  levels = c("red", "blue", "green", "yellow")
)
all_colors
```

```{r}
# All four levels are retained, even though only two are present
levels(all_colors)
```

```{r}
# table() shows counts for all levels, including zeros
table(all_colors)
```

This behavior is valuable when creating frequency tables or bar charts where you need to show that certain categories exist but have zero observations.


## Inspecting Factors

Before manipulating factors, it is important to understand what you are working with. Several functions help you inspect factor variables.

### The gss_cat Dataset

The `gss_cat` dataset, included in the forcats package, is a sample from the General Social Survey. It contains several factor variables and will serve as our primary example throughout this chapter.

```{r}
glimpse(gss_cat)
```

### levels() and nlevels()

The `levels()` function returns the levels of a factor in order. The `nlevels()` function returns the count.

```{r}
levels(gss_cat$marital)
```

```{r}
nlevels(gss_cat$marital)
```

```{r}
levels(gss_cat$race)
```

### fct_count()

The `fct_count()` function from forcats provides a quick frequency table as a tibble, which is easier to work with than base R's `table()`.

```{r}
fct_count(gss_cat$marital)
```

```{r}
# Sort by frequency
fct_count(gss_cat$marital, sort = TRUE)
```

### fct_unique()

The `fct_unique()` function returns the unique values of a factor in the order of the levels (not in the order of first appearance, as `unique()` does for character vectors).

```{r}
fct_unique(gss_cat$race)
```

### table()

Base R's `table()` function remains useful for quick cross-tabulations.

```{r}
table(gss_cat$race)
```

```{r}
# Cross-tabulation of two factor variables
table(gss_cat$race, gss_cat$marital)
```


## Reordering Factor Levels

Reordering factor levels is the single most common factor manipulation task, especially when creating visualizations. The forcats package provides a family of `fct_*` functions for this purpose.

### fct_relevel() --- Manual Reordering

The `fct_relevel()` function lets you manually set the order of levels. You can move specific levels to the front, or place them at a specific position.

```{r}
# Default level order of marital status
levels(gss_cat$marital)
```

```{r}
# Move "Never married" to the first position
gss_cat |>
  mutate(marital = fct_relevel(marital, "Never married")) |>
  pull(marital) |>
  levels()
```

```{r}
# Move "Married" to the second position
gss_cat |>
  mutate(marital = fct_relevel(marital, "Married", after = 1)) |>
  pull(marital) |>
  levels()
```

```{r}
# Move "Widowed" to the end
gss_cat |>
  mutate(marital = fct_relevel(marital, "Widowed", after = Inf)) |>
  pull(marital) |>
  levels()
```

Here is how `fct_relevel()` affects a bar chart:

```{r fig.cap="Default alphabetical level order for marital status."}
gss_cat |>
  ggplot(aes(x = marital)) +
  geom_bar() +
  labs(title = "Marital Status (default level order)", x = NULL, y = "Count") +
  theme_minimal()
```

```{r fig.cap="Marital status reordered with fct_relevel."}
gss_cat |>
  mutate(marital = fct_relevel(marital,
    "Never married", "Married", "Separated",
    "Divorced", "Widowed", "No answer")) |>
  ggplot(aes(x = marital)) +
  geom_bar() +
  labs(title = "Marital Status (manually reordered)", x = NULL, y = "Count") +
  theme_minimal()
```

### fct_infreq() --- Order by Frequency

The `fct_infreq()` function reorders levels from most frequent to least frequent. This is one of the most commonly used reordering functions for bar charts.

```{r fig.cap="Marital status bars ordered by frequency using fct_infreq."}
gss_cat |>
  mutate(marital = fct_infreq(marital)) |>
  ggplot(aes(x = marital)) +
  geom_bar() +
  labs(title = "Marital Status (ordered by frequency)", x = NULL, y = "Count") +
  theme_minimal()
```

### fct_rev() --- Reverse Order

The `fct_rev()` function reverses the order of levels. It is often used in combination with other reordering functions, especially for horizontal bar charts where you want the highest bar at the top.

```{r fig.cap="Horizontal bar chart with levels reversed for top-to-bottom reading."}
gss_cat |>
  mutate(marital = fct_rev(fct_infreq(marital))) |>
  ggplot(aes(y = marital)) +
  geom_bar() +
  labs(title = "Marital Status (reversed frequency order)", x = "Count", y = NULL) +
  theme_minimal()
```

### fct_reorder() --- Reorder by Another Variable

The `fct_reorder()` function is critical for ggplot2 bar charts. It reorders a factor by the values of another variable, using a summary function (default is `median`).

```{r fig.cap="Religion ordered by median hours of TV watched per day."}
relig_summary <- gss_cat |>
  group_by(relig) |>
  summarize(
    tvhours = mean(tvhours, na.rm = TRUE),
    n = n()
  )

# Without reordering
ggplot(relig_summary, aes(x = tvhours, y = relig)) +
  geom_point() +
  labs(title = "TV hours by religion (unordered)", x = "Mean hours of TV per day", y = NULL) +
  theme_minimal()
```

```{r fig.cap="Religion ordered by mean TV hours, making patterns far easier to read."}
# With fct_reorder --- much easier to read
ggplot(relig_summary, aes(x = tvhours, y = fct_reorder(relig, tvhours))) +
  geom_point() +
  labs(title = "TV hours by religion (ordered by TV hours)",
       x = "Mean hours of TV per day", y = NULL) +
  theme_minimal()
```

The improvement is dramatic. Reordering the factor by the plotted value transforms a chaotic scatter of points into a clear, ranked list.

Another example using the `iris` dataset:

```{r fig.cap="Mean petal length by species, with species ordered by value."}
iris |>
  group_by(Species) |>
  summarize(mean_petal = mean(Petal.Length)) |>
  ggplot(aes(x = mean_petal, y = fct_reorder(Species, mean_petal))) +
  geom_col(fill = "steelblue") +
  labs(title = "Mean Petal Length by Species", x = "Mean Petal Length", y = NULL) +
  theme_minimal()
```

### fct_reorder2() --- Reorder for Line Plots

The `fct_reorder2()` function reorders a factor by two variables, which is useful for aligning the legend order with the endpoints of lines in a line plot.

```{r fig.cap="Line plot with legend order matching the final data points."}
by_age <- gss_cat |>
  filter(!is.na(age)) |>
  count(age, marital) |>
  group_by(age) |>
  mutate(prop = n / sum(n))

ggplot(by_age, aes(x = age, y = prop, color = fct_reorder2(marital, age, prop))) +
  geom_line(linewidth = 1) +
  labs(title = "Marital status proportion by age",
       x = "Age", y = "Proportion", color = "Marital Status") +
  theme_minimal()
```

### fct_inorder() --- Order of First Appearance

The `fct_inorder()` function sets levels in the order they first appear in the data. This is useful when your data is already sorted in a meaningful way.

```{r}
months_data <- c("Jan", "Mar", "Jan", "Feb", "Mar", "Feb")
fct_inorder(months_data)
```

### fct_shift() --- Shift Levels

The `fct_shift()` function shifts levels to the left (positive `n`) or to the right (negative `n`), wrapping around. This can be useful for rotating cyclical data like days of the week or months.

```{r}
days <- factor(c("Mon", "Tue", "Wed", "Thu", "Fri"),
               levels = c("Mon", "Tue", "Wed", "Thu", "Fri"))
levels(days)
```

```{r}
# Shift levels left by 2 positions
levels(fct_shift(days, n = 2))
```


## Modifying Factor Levels

Beyond reordering, you often need to rename, combine, or collapse factor levels. The forcats package provides several functions for these tasks.

### fct_recode() --- Rename Levels

The `fct_recode()` function renames individual levels. The syntax is `new_name = "old_name"`.

```{r}
gss_cat |>
  mutate(partyid = fct_recode(partyid,
    "Republican (strong)"    = "Strong republican",
    "Republican (weak)"      = "Not str republican",
    "Independent (near rep)" = "Ind,near rep",
    "Independent (near dem)" = "Ind,near dem",
    "Democrat (weak)"        = "Not str democrat",
    "Democrat (strong)"      = "Strong democrat"
  )) |>
  count(partyid)
```

### fct_collapse() --- Combine Groups of Levels

The `fct_collapse()` function combines multiple levels into new groups. This is more efficient than calling `fct_recode()` multiple times when you want to merge several levels into one.

```{r}
gss_cat |>
  mutate(partyid = fct_collapse(partyid,
    "Other"       = c("No answer", "Don't know", "Other party"),
    "Republican"  = c("Strong republican", "Not str republican"),
    "Independent" = c("Ind,near rep", "Independent", "Ind,near dem"),
    "Democrat"    = c("Not str democrat", "Strong democrat")
  )) |>
  count(partyid)
```

### fct_lump_*() --- Lump Rare Levels

The `fct_lump_*()` family of functions collapses infrequent levels into an "Other" category. There are several variants for different lumping strategies.

**`fct_lump_n()`** keeps the `n` most frequent levels and lumps the rest.

```{r}
# Keep the 4 most common religions, lump the rest into "Other"
gss_cat |>
  mutate(relig = fct_lump_n(relig, n = 4)) |>
  count(relig, sort = TRUE)
```

**`fct_lump_prop()`** keeps levels that appear in at least the specified proportion of observations.

```{r}
# Keep religions that represent at least 10% of respondents
gss_cat |>
  mutate(relig = fct_lump_prop(relig, prop = 0.10)) |>
  count(relig, sort = TRUE)
```

**`fct_lump_min()`** keeps levels that appear at least `min` times.

```{r}
# Keep religions with at least 2000 respondents
gss_cat |>
  mutate(relig = fct_lump_min(relig, min = 2000)) |>
  count(relig, sort = TRUE)
```

**`fct_lump_lowfreq()`** lumps together the least frequent levels, ensuring that "Other" is still the smallest category.

```{r}
gss_cat |>
  mutate(relig = fct_lump_lowfreq(relig)) |>
  count(relig, sort = TRUE)
```

### fct_other() --- Keep or Drop Specific Levels

The `fct_other()` function replaces all levels except those you specify with "Other". You can use the `keep` argument to specify which levels to retain, or the `drop` argument to specify which levels to replace.

```{r}
# Keep only "Married" and "Never married", lump everything else
gss_cat |>
  mutate(marital = fct_other(marital, keep = c("Married", "Never married"))) |>
  count(marital)
```

```{r}
# Drop "No answer" into "Other", keep everything else
gss_cat |>
  mutate(marital = fct_other(marital, drop = "No answer")) |>
  count(marital)
```


## Adding and Dropping Levels

### fct_expand() --- Add New Levels

The `fct_expand()` function adds new levels to a factor, even if no observations have those values. This is useful when you need to ensure that a factor includes all possible categories.

```{r}
# The iris Species factor has 3 levels
levels(iris$Species)
```

```{r}
# Add a hypothetical fourth species
iris_expanded <- iris |>
  mutate(Species = fct_expand(Species, "unknown"))
levels(iris_expanded$Species)
```

```{r}
# table() now shows the empty level
table(iris_expanded$Species)
```

### fct_drop() --- Drop Unused Levels

The `fct_drop()` function removes levels that have no observations. This is the inverse of `fct_expand()`.

```{r}
# Filter iris to only setosa, but the factor still has all three levels
setosa_only <- iris |> filter(Species == "setosa")
levels(setosa_only$Species)
```

```{r}
# Drop unused levels
setosa_only |>
  mutate(Species = fct_drop(Species)) |>
  pull(Species) |>
  levels()
```

### Handling NAs as Factor Levels

Sometimes `NA` values in a factor should be treated as an explicit category rather than as missing data. The forcats package provides functions to convert between `NA` values and explicit factor levels.

**`fct_na_value_to_level()`** converts `NA` values to an explicit level (default name is "(Missing)").

```{r}
# Create a factor with some NAs
responses <- factor(c("yes", "no", NA, "yes", NA, "no"))
responses
```

```{r}
# Convert NAs to an explicit level
fct_na_value_to_level(responses, level = "No Response")
```

**`fct_na_level_to_value()`** does the reverse, converting a named level back to `NA`.

```{r}
# Convert the explicit level back to NA
responses_with_level <- fct_na_value_to_level(responses, level = "No Response")
fct_na_level_to_value(responses_with_level, extra_levels = "No Response")
```

This is particularly useful when you want `NA` counts to appear in bar charts or frequency tables, and then convert them back to proper `NA` values for statistical analysis.


## Practical Examples with ggplot2

Factor manipulation and data visualization go hand in hand. In this section, we bring together the techniques from this chapter to show how thoughtful factor handling transforms visualizations from confusing to clear.

### Example 1: Reordering Bars by Value

One of the most common visualization improvements is reordering bars by their values rather than displaying them alphabetically.

```{r fig.cap="Income distribution from the GSS, with bars ordered by count."}
gss_cat |>
  filter(!is.na(rincome), rincome != "Not applicable") |>
  mutate(rincome = fct_infreq(rincome)) |>
  ggplot(aes(y = rincome)) +
  geom_bar(fill = "steelblue") +
  labs(title = "Reported Income (ordered by frequency)",
       x = "Count", y = NULL) +
  theme_minimal()
```

But for income, the natural ordering (low to high) is more meaningful than frequency ordering. We can use `fct_relevel()` or ensure the original factor order is preserved:

```{r fig.cap="Income distribution preserving the natural low-to-high ordering."}
gss_cat |>
  filter(!is.na(rincome), rincome != "Not applicable") |>
  ggplot(aes(y = rincome)) +
  geom_bar(fill = "darkorange") +
  labs(title = "Reported Income (natural order)",
       x = "Count", y = NULL) +
  theme_minimal()
```

### Example 2: Lumping Small Categories for Cleaner Plots

Plots with too many categories become unreadable. Lumping rare categories into "Other" dramatically improves clarity.

```{r fig.cap="All denominations shown, creating an overwhelming plot."}
gss_cat |>
  count(denom) |>
  ggplot(aes(x = n, y = denom)) +
  geom_point() +
  labs(title = "Denomination (all levels --- hard to read)",
       x = "Count", y = NULL) +
  theme_minimal()
```

```{r fig.cap="Denominations lumped to the top 8, with the rest grouped as Other."}
gss_cat |>
  mutate(denom = fct_lump_n(denom, n = 8)) |>
  count(denom) |>
  mutate(denom = fct_reorder(denom, n)) |>
  ggplot(aes(x = n, y = denom)) +
  geom_point(size = 3, color = "steelblue") +
  labs(title = "Denomination (top 8, rest lumped into Other)",
       x = "Count", y = NULL) +
  theme_minimal()
```

### Example 3: Controlling Legend Order in Line Plots

When a line plot has multiple series, the legend should match the visual order of the lines at the right edge of the plot. The `fct_reorder2()` function achieves this.

```{r fig.cap="Proportion of top 4 marital statuses by age with aligned legend."}
gss_cat |>
  filter(!is.na(age)) |>
  mutate(marital = fct_lump_n(marital, n = 4)) |>
  count(age, marital) |>
  group_by(age) |>
  mutate(prop = n / sum(n)) |>
  ungroup() |>
  ggplot(aes(x = age, y = prop, color = fct_reorder2(marital, age, prop))) +
  geom_line(linewidth = 1) +
  scale_color_brewer(palette = "Set1") +
  labs(title = "Marital Status by Age",
       x = "Age", y = "Proportion", color = "Status") +
  theme_minimal()
```

### Example 4: Combining fct_collapse() with Visualization

Collapsing factor levels before plotting is a common workflow for simplifying complex categorical data.

```{r fig.cap="Political party affiliation after collapsing levels into broader groups."}
gss_cat |>
  mutate(party = fct_collapse(partyid,
    "Republican"  = c("Strong republican", "Not str republican"),
    "Independent" = c("Ind,near rep", "Independent", "Ind,near dem"),
    "Democrat"    = c("Not str democrat", "Strong democrat"),
    "Other"       = c("No answer", "Don't know", "Other party")
  )) |>
  mutate(party = fct_infreq(party)) |>
  ggplot(aes(x = party, fill = party)) +
  geom_bar(show.legend = FALSE) +
  scale_fill_manual(values = c(
    "Democrat" = "blue", "Independent" = "gray50",
    "Republican" = "red", "Other" = "gray80"
  )) +
  labs(title = "Political Affiliation (collapsed categories)",
       x = NULL, y = "Count") +
  theme_minimal()
```

### Example 5: Using Factors with mtcars

The `mtcars` dataset has several variables that are coded as numbers but are really categorical: `cyl`, `am`, `gear`, and `vs`. Converting them to factors with meaningful labels makes plots more informative.

```{r fig.cap="Mean MPG by number of cylinders and transmission type with labeled factors."}
mtcars |>
  mutate(
    cyl = factor(cyl),
    transmission = factor(am, levels = c(0, 1), labels = c("Automatic", "Manual"))
  ) |>
  group_by(cyl, transmission) |>
  summarize(mean_mpg = mean(mpg), .groups = "drop") |>
  ggplot(aes(x = cyl, y = mean_mpg, fill = transmission)) +
  geom_col(position = "dodge") +
  scale_fill_brewer(palette = "Set2") +
  labs(title = "Mean MPG by Cylinders and Transmission",
       x = "Number of Cylinders", y = "Mean MPG", fill = "Transmission") +
  theme_minimal()
```

### Example 6: Showing Empty Factor Levels

When comparing groups, it can be important to show categories with zero counts rather than silently omitting them. Factors make this possible with the `.drop = FALSE` argument in ggplot2 scales.

```{r fig.cap="Bar chart showing all factor levels including those with zero count."}
# Create a subset with only some species
iris_subset <- iris |>
  filter(Species != "versicolor")

# Without .drop = FALSE, the versicolor bar disappears
ggplot(iris_subset, aes(x = Species)) +
  geom_bar(fill = "steelblue") +
  scale_x_discrete(drop = FALSE) +
  labs(title = "Iris species counts (empty level shown)",
       x = NULL, y = "Count") +
  theme_minimal()
```


## Summary

The forcats package transforms factor manipulation from a frustrating chore into a clean, predictable workflow. Its key functions fall into four categories:

| Task | Key Functions |
|:--------------------------|:----------------------------------------------|
| **Creating factors** | `factor()`, `parse_factor()`, `as.factor()` |
| **Inspecting factors** | `levels()`, `nlevels()`, `fct_count()`, `fct_unique()` |
| **Reordering levels** | `fct_relevel()`, `fct_infreq()`, `fct_rev()`, `fct_reorder()`, `fct_reorder2()`, `fct_inorder()`, `fct_shift()` |
| **Modifying levels** | `fct_recode()`, `fct_collapse()`, `fct_lump_n()`, `fct_lump_prop()`, `fct_lump_min()`, `fct_lump_lowfreq()`, `fct_other()` |
| **Adding/dropping levels** | `fct_expand()`, `fct_drop()`, `fct_na_value_to_level()`, `fct_na_level_to_value()` |

The single most impactful skill from this chapter is using `fct_reorder()` and `fct_infreq()` to control the order of categories in ggplot2 visualizations. Alphabetical ordering is almost never the most informative ordering, and a well-ordered factor can transform a confusing plot into one that tells a clear story. As you work through the visualization-focused companion course, you will find yourself reaching for these functions constantly.
