# Working with Dates and Times {#lubridate}

```{r include=FALSE}
library(tidyverse)
library(lubridate)
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

## Dates Are Weird. Lubridate Makes Them Less Weird.

Dates look simple. Everyone knows what "January 15, 2025" means. But try to do math with dates and things get strange fast. Months have different numbers of days. Years sometimes have an extra day. And don't even get me started on daylight saving time.

Base R has tools for dates, but they require you to memorize format strings like `"%m/%d/%Y"` (and yes, the capital Y matters). The **lubridate** package takes a different approach: just tell it what order the year, month, and day appear in, and it figures out the rest.

One important note: lubridate is NOT loaded automatically with `library(tidyverse)`. You need to load it separately:

```{r eval=FALSE}
library(lubridate)
```

Let's jump in.


## Parsing Dates: `ymd()`, `mdy()`, `dmy()`

This is the single best thing about lubridate. Instead of memorizing format codes, you just pick the function that matches the order of your date components.

**Year-Month-Day?** Use `ymd()`:

```{r}
ymd("2025-06-15")
ymd("2025/06/15")
ymd("20250615")
```

**Month-Day-Year (the American way)?** Use `mdy()`:

```{r}
mdy("06-15-2025")
mdy("June 15, 2025")
mdy("Jun 15 2025")
```

**Day-Month-Year (the rest-of-the-world way)?** Use `dmy()`:

```{r}
dmy("15-06-2025")
dmy("15 June 2025")
dmy("15/06/2025")
```

Notice that each function handles slashes, dashes, spaces, full month names, and abbreviations. You just tell it the order and lubridate does the rest. Compare that with base R's `as.Date("06/15/2025", format = "%m/%d/%Y")` and you'll never look back.

These functions are vectorized, so they work on entire columns:

```{r}
date_strings <- c("January 1, 2025", "February 14, 2025", "July 4, 2025", "December 25, 2025")
mdy(date_strings)
```


## Adding Time: `ymd_hms()` and Friends

Need date AND time? Just add `_hms` (hours, minutes, seconds) to the function name:

```{r}
ymd_hms("2025-06-15 14:30:45")
mdy_hm("June 15, 2025 2:30 PM")
```

You can also specify a time zone:

```{r}
ymd_hms("2025-06-15 14:30:45", tz = "America/New_York")
```


## Building Dates from Separate Columns

Sometimes your data has year, month, and day in separate columns (looking at you, Excel exports). `make_date()` puts them together:

```{r}
event_data <- tibble(
  event = c("Conference", "Workshop", "Seminar", "Retreat"),
  yr = c(2025, 2025, 2025, 2026),
  mo = c(3, 6, 9, 1),
  dy = c(10, 22, 5, 15)
)

event_data %>%
  mutate(event_date = make_date(yr, mo, dy))
```


## Today and Now

Need the current date or time? Easy:

```{r}
today()
now()
```

These are surprisingly useful for things like calculating how many days until a deadline or filtering data to "last 30 days."


## Extracting Components: Year, Month, Day, etc.

Once you have a proper date object, you can pull it apart into pieces. This is where dates become really useful for business analysis --- think "sales by month" or "orders by day of week."

```{r}
dt <- ymd_hms("2025-06-15 14:30:45")

year(dt)
month(dt)
mday(dt)      # day of the month
wday(dt)      # day of the week (1 = Sunday by default)
hour(dt)
quarter(dt)
```

Want month names instead of numbers? Set `label = TRUE`:

```{r}
month(dt, label = TRUE)
month(dt, label = TRUE, abbr = FALSE)
```

Same for day of the week:

```{r}
wday(dt, label = TRUE)
wday(dt, label = TRUE, abbr = FALSE)
```

Here's how this looks in a real workflow --- breaking a transaction date into analysis-friendly components:

```{r}
sales_data <- tibble(
  transaction_date = ymd_hms(c(
    "2025-01-10 09:15:00", "2025-02-14 11:30:00",
    "2025-03-22 16:45:00", "2025-04-01 08:00:00",
    "2025-05-18 13:20:00", "2025-06-30 17:55:00"
  )),
  amount = c(150, 280, 95, 420, 175, 310)
)

sales_data %>%
  mutate(
    sale_month = month(transaction_date, label = TRUE),
    sale_day = wday(transaction_date, label = TRUE),
    sale_quarter = quarter(transaction_date)
  )
```

Now you can group by month, day of week, or quarter. That's how you answer questions like "Do we sell more on weekdays or weekends?" or "Which quarter had the highest revenue?"


## Date Math: Adding and Subtracting Time

This is one of the best parts. You can just add time to dates. Lubridate gives you two ways to think about it:

**Durations** --- exact amounts of time measured in seconds:

```{r}
ymd("2025-01-01") + ddays(30)
ymd("2025-01-01") + dweeks(2)
ymd_hms("2025-06-15 10:00:00") + dhours(3)
```

**Periods** --- calendar-aware amounts (respects different month lengths):

```{r}
ymd("2025-01-01") + months(6)
ymd("2025-01-31") + months(1)    # February doesn't have 31 days, so you get NA
ymd("2025-03-15") + years(2) + months(3) + days(10)
```

**When to use which?** Use durations when you need an exact elapsed time ("How many seconds did this process take?"). Use periods when you need calendar math ("What's the date 3 months from now?"). For most business analysis, periods are what you want.

You can also just subtract dates to get the difference:

```{r}
as.numeric(ymd("2025-12-31") - ymd("2025-01-01"))
```

That's the number of days between two dates. Simple.


## Intervals: Did This Event Happen During This Period?

An interval is just a span of time with a start and end. The main use case is checking whether a date falls within a range.

```{r}
spring_semester <- ymd("2025-01-13") %--% ymd("2025-05-09")

ymd("2025-03-15") %within% spring_semester
ymd("2025-06-01") %within% spring_semester
```

Useful for questions like "Did this sale happen during our Q2 promotion?" or "Was this employee hired during the pandemic?"


## Rounding Dates: Grouping by Time Period

`floor_date()` rounds a date down to the nearest unit. This is incredibly useful for aggregating data --- turning daily transactions into monthly summaries, for example.

```{r}
transactions <- tibble(
  date = ymd(c(
    "2025-01-05", "2025-01-18", "2025-01-29",
    "2025-02-03", "2025-02-14", "2025-02-27",
    "2025-03-08", "2025-03-15", "2025-03-30"
  )),
  revenue = c(1200, 950, 1100, 1350, 800, 1500, 1050, 1200, 900)
)

transactions %>%
  mutate(month = floor_date(date, unit = "month")) %>%
  group_by(month) %>%
  summarize(
    total_revenue = sum(revenue),
    n_transactions = n()
  )
```

You can also round to weeks, quarters, or even custom intervals like "15 minutes" or "6 hours" if you're working with high-frequency data.


## Time Zones: The One Paragraph Version

Time zones are complicated, but here's what you need to know. `with_tz()` converts a time for display ("What time is this New York meeting in Tokyo?") without changing the actual moment. `force_tz()` corrects a mislabeled time zone ("This was recorded as UTC but it was actually Eastern"). If you're working with data from a single time zone, you probably don't need either of these.

```{r}
meeting_time <- ymd_hms("2025-06-15 10:00:00", tz = "America/New_York")
with_tz(meeting_time, tzone = "Asia/Tokyo")    # same moment, displayed in Tokyo time
```


## Practical Example: Analyzing Economic Trends

Let's use lubridate with the built-in `economics` dataset to do some real analysis. This dataset has monthly US economic data from 1967 to 2015.

```{r}
economics %>%
  mutate(decade = factor(10 * (year(date) %/% 10))) %>%
  ggplot(aes(x = date, y = unemploy, color = decade)) +
  geom_line() +
  labs(
    title = "US Unemployment Over Time",
    x = "Date", y = "Unemployment (thousands)", color = "Decade"
  ) +
  theme_minimal()
```

```{r}
economics %>%
  mutate(month = month(date, label = TRUE)) %>%
  group_by(month) %>%
  summarize(avg_unemploy = mean(unemploy)) %>%
  ggplot(aes(x = month, y = avg_unemploy)) +
  geom_col(fill = "steelblue") +
  labs(
    title = "Average US Unemployment by Month (1967-2015)",
    x = "Month", y = "Unemployment (thousands)"
  ) +
  theme_minimal()
```

```{r}
# Filter to the Great Recession and plot
economics %>%
  filter(between(date, ymd("2007-06-01"), ymd("2009-06-01"))) %>%
  ggplot(aes(x = date, y = unemploy)) +
  geom_line(color = "firebrick", linewidth = 1) +
  labs(
    title = "US Unemployment During the Great Recession",
    x = "Date", y = "Unemployment (thousands)"
  ) +
  theme_minimal()
```

See how `year()`, `month()`, `between()`, and `ymd()` all work together seamlessly with dplyr and ggplot2? That's the whole point.


## Quick Reference

| What you want to do | Function |
|:---------------------|:---------|
| Parse a date string | `ymd()`, `mdy()`, `dmy()` (match the order) |
| Parse date + time | `ymd_hms()`, `mdy_hm()`, etc. |
| Build from components | `make_date()`, `make_datetime()` |
| Current date/time | `today()`, `now()` |
| Extract year, month, etc. | `year()`, `month()`, `mday()`, `wday()`, `quarter()` |
| Add calendar time | `+ months()`, `+ years()`, `+ days()` |
| Add exact time | `+ dhours()`, `+ ddays()`, `+ dweeks()` |
| Round to period | `floor_date()`, `ceiling_date()`, `round_date()` |
| Check if date is in range | `%within%` with intervals (`%--%`) |

Dates are weird in every language. But with lubridate, they're manageable. Parse them, extract what you need, do the math, and get on with your analysis.
