# Working with Dates and Times {#lubridate}

```{r include=FALSE}
library(tidyverse)
library(lubridate)
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

Dates and times are among the most deceptively difficult data types in programming. They seem straightforward on the surface --- everyone knows what "January 15, 2025" means --- but beneath that simplicity lies a thicket of complications. Months have different numbers of days. Years can be leap years. Time zones shift boundaries. Daylight saving time creates hours that happen twice and hours that do not happen at all. Even the seemingly simple question "how many days are between these two dates?" can have a surprising answer when a daylight saving transition falls in between.

Base R provides tools for working with dates and times, but their interfaces are inconsistent and their behavior can be unintuitive. The **lubridate** package, created by Garrett Grolemund and Hadley Wickham, provides a consistent, human-friendly toolkit for parsing, manipulating, and performing arithmetic with dates and times. It is not loaded automatically with `library(tidyverse)`, so you must load it explicitly.

## Date and Time Classes in R

Before diving into lubridate's functions, it helps to understand R's three main date/time classes.

| Class | Stores | Created by | Example |
|:------------|:---------------------------------|:---------------------------|:-------------------------|
| `Date` | A date (no time) | `as.Date()`, lubridate | `2025-06-15` |
| `POSIXct` | A date and time | `as.POSIXct()`, lubridate | `2025-06-15 14:30:00` |
| `hms` | A time of day (no date) | `hms::hms()` | `14:30:00` |

Internally, R stores `Date` objects as the number of days since January 1, 1970 (the Unix epoch), and `POSIXct` objects as the number of seconds since the same epoch. This numeric representation is what makes date arithmetic possible.

```{r}
# A Date object is stored as days since 1970-01-01
d <- as.Date("2025-01-15")
d
unclass(d)  # number of days since the epoch
```

```{r}
# A POSIXct object is stored as seconds since 1970-01-01
dt <- as.POSIXct("2025-01-15 10:30:00", tz = "UTC")
dt
unclass(dt)  # number of seconds since the epoch
```

Understanding this internal representation helps explain why lubridate's arithmetic functions work the way they do.


## Creating Dates and Times

Lubridate provides several families of functions for creating date and time objects. The most convenient are the parsing functions that automatically detect the format of a date string based on the order of its components.

### Parsing Dates from Strings

The key insight behind lubridate's parsing functions is that you only need to specify the **order** of the year, month, and day components --- not the exact format. The function name tells lubridate what order to expect.

```{r}
# Year-Month-Day
ymd("2025-06-15")
ymd("2025/06/15")
ymd("20250615")
```

```{r}
# Month-Day-Year (common US format)
mdy("06-15-2025")
mdy("June 15, 2025")
mdy("Jun 15 2025")
```

```{r}
# Day-Month-Year (common European format)
dmy("15-06-2025")
dmy("15 June 2025")
dmy("15/06/2025")
```

Notice that each function handles a wide variety of separators, abbreviations, and formats --- all you need to specify is the order. This is far more convenient than base R's `as.Date()`, which requires an explicit format string like `"%m/%d/%Y"`.

### Parsing Date-Times from Strings

Adding `_hms`, `_hm`, or `_h` to the function name tells lubridate to also parse time components (hours, minutes, seconds).

```{r}
# Date-time with hours, minutes, and seconds
ymd_hms("2025-06-15 14:30:45")
ymd_hms("2025/06/15 14:30:45")
```

```{r}
# Date-time with hours and minutes only
ymd_hm("2025-06-15 14:30")
mdy_hm("June 15, 2025 2:30 PM")
```

```{r}
# Specify a time zone
ymd_hms("2025-06-15 14:30:45", tz = "America/New_York")
ymd_hms("2025-06-15 14:30:45", tz = "Europe/London")
```

By default, lubridate's parsing functions return dates in UTC. If your data represents times in a specific time zone, always specify the `tz` argument.

### Parsing Vectors of Dates

These functions are vectorized, so they work naturally on entire columns of data.

```{r}
date_strings <- c("2025-01-15", "2025-03-22", "2025-07-04", "2025-12-25")
ymd(date_strings)
```

```{r}
us_dates <- c("January 1, 2025", "February 14, 2025", "July 4, 2025", "December 25, 2025")
mdy(us_dates)
```

### Creating Dates from Individual Components

When your data has year, month, and day in separate columns, use `make_date()` or `make_datetime()` to assemble them into a proper date object.

```{r}
make_date(year = 2025, month = 6, day = 15)
```

```{r}
make_datetime(year = 2025, month = 6, day = 15, hour = 14, min = 30, sec = 0)
```

```{r}
# These functions are vectorized, making them ideal for use inside mutate()
event_data <- tibble(
  event = c("Conference", "Workshop", "Seminar", "Retreat"),
  yr = c(2025, 2025, 2025, 2026),
  mo = c(3, 6, 9, 1),
  dy = c(10, 22, 5, 15)
)

event_data |>
  mutate(event_date = make_date(yr, mo, dy))
```

### Converting from Other Types

Lubridate provides `as_date()` and `as_datetime()` for converting between date types and from numeric values.

```{r}
# Convert a datetime to a date (drops the time component)
dt <- ymd_hms("2025-06-15 14:30:00")
as_date(dt)
```

```{r}
# Convert a date to a datetime (sets time to midnight)
d <- ymd("2025-06-15")
as_datetime(d)
```

```{r}
# Convert from a numeric value (days since epoch for dates, seconds for datetimes)
as_date(20000)        # 20000 days after 1970-01-01
as_datetime(1750000000)  # 1.75 billion seconds after the epoch
```

### Getting the Current Date and Time

The `today()` and `now()` functions return the current date and date-time, respectively.

```{r}
today()
now()
```

```{r}
# Specify a time zone
now("America/Chicago")
now("Asia/Tokyo")
```


## Getting and Setting Components

Once you have a date or datetime object, you often need to extract individual components --- the year, the month, the day of the week, and so on. Lubridate provides a complete set of accessor functions for this purpose.

### Extracting Components

```{r}
dt <- ymd_hms("2025-06-15 14:30:45")

year(dt)
month(dt)
mday(dt)    # day of the month
yday(dt)    # day of the year
wday(dt)    # day of the week (1 = Sunday by default)
hour(dt)
minute(dt)
second(dt)
```

### Labels for Month and Day of Week

By default, `month()` and `wday()` return numeric values. Setting `label = TRUE` returns an ordered factor with abbreviated names, which is especially useful for plotting.

```{r}
month(dt, label = TRUE)
month(dt, label = TRUE, abbr = FALSE)  # full month name
```

```{r}
wday(dt, label = TRUE)
wday(dt, label = TRUE, abbr = FALSE)  # full day name
```

You can also change the start of the week. By default, `wday()` uses Sunday as day 1 (the US convention). Set `week_start = 1` to use Monday as day 1 (the ISO convention).

```{r}
# Sunday = 1 (default)
wday(ymd("2025-06-15"))

# Monday = 1
wday(ymd("2025-06-15"), week_start = 1)
```

### Additional Component Functions

```{r}
dt <- ymd_hms("2025-06-15 14:30:45")

# Quarter of the year
quarter(dt)

# Week of the year
week(dt)
isoweek(dt)  # ISO 8601 week number

# Whether the year is a leap year
leap_year(2024)
leap_year(2025)

# AM or PM
am(dt)
pm(dt)
```

### Setting Components

Each accessor function can also be used as a setter by assigning a value to it. This modifies the date in place.

```{r}
dt <- ymd_hms("2025-06-15 14:30:45")
dt

year(dt) <- 2026
dt

month(dt) <- 1
dt

mday(dt) <- 20
dt

hour(dt) <- 9
dt
```

### Using update() for Multiple Changes

The `update()` function lets you change multiple components at once without modifying the original object.

```{r}
dt <- ymd_hms("2025-06-15 14:30:45")

# Change the year, month, and hour simultaneously
update(dt, year = 2026, month = 12, hour = 8)

# The original object is unchanged
dt
```

### Extracting Components in Data Frames

These functions integrate naturally with dplyr's `mutate()`.

```{r}
sales_data <- tibble(
  transaction_date = ymd_hms(c(
    "2025-01-10 09:15:00", "2025-02-14 11:30:00",
    "2025-03-22 16:45:00", "2025-04-01 08:00:00",
    "2025-05-18 13:20:00", "2025-06-30 17:55:00"
  )),
  amount = c(150, 280, 95, 420, 175, 310)
)

sales_data |>
  mutate(
    sale_year = year(transaction_date),
    sale_month = month(transaction_date, label = TRUE),
    sale_day = wday(transaction_date, label = TRUE),
    sale_hour = hour(transaction_date),
    sale_quarter = quarter(transaction_date)
  )
```


## Time Spans

One of lubridate's most valuable features is its system for representing and computing with time spans. There are three distinct types, each suited to different situations.

### Durations: Exact Number of Seconds

A **duration** represents an exact number of seconds. Because it uses a fixed unit, durations are mathematically precise but do not always align with human intuition about time. For example, a "duration of one year" is always exactly 365.25 days (31,557,600 seconds), regardless of whether the actual year is a leap year.

```{r}
dseconds(30)
dminutes(5)
dhours(2)
ddays(7)
dweeks(4)
dyears(1)
```

```{r}
# Durations can be added to dates
ymd("2025-01-01") + ddays(30)
ymd("2025-01-01") + dweeks(2)
ymd_hms("2025-06-15 10:00:00") + dhours(3)
```

```{r}
# Duration arithmetic
ddays(5) + dhours(12)
dweeks(2) - ddays(3)
3 * dhours(8)
```

Because durations are stored as seconds internally, they can produce unexpected results across daylight saving time boundaries. For clock-time calculations, periods (described next) are usually more appropriate.

### Periods: Human-Friendly Time Spans

A **period** represents a time span in human units --- years, months, weeks, days, hours, minutes, and seconds. Unlike durations, periods respect the varying lengths of months and years. Adding "one month" to January 31 gives February 28 (or 29), not a fixed number of seconds later.

```{r}
seconds(30)
minutes(5)
hours(2)
days(7)
weeks(4)
months(3)
years(1)
```

```{r}
# Periods respect calendar irregularities
ymd("2025-01-31") + months(1)   # February doesn't have 31 days
ymd("2024-02-29") + years(1)    # 2025 is not a leap year
```

```{r}
# Adding periods to dates
ymd("2025-01-01") + months(6)
ymd("2025-03-15") + years(2) + months(3) + days(10)
```

```{r}
# Period arithmetic
months(6) + days(15)
years(1) - months(3)
```

### When to Use Durations vs. Periods

The choice between durations and periods depends on what question you are asking:

- Use **durations** when you need an exact physical elapsed time (e.g., "How many seconds did this process take?").
- Use **periods** when you need calendar-aware arithmetic (e.g., "What date is three months from now?").

```{r}
# Durations: exact elapsed time
start <- ymd_hms("2025-03-09 01:00:00", tz = "America/New_York")
start + dhours(2)   # exactly 7200 seconds later

# Periods: calendar-based
start + hours(2)    # two clock-hours later (may differ from duration across DST)
```

### Intervals: A Span with a Start and End

An **interval** represents the time span between two specific points in time. Unlike durations and periods, an interval knows exactly when it starts and ends, which makes it possible to convert it precisely to any unit.

```{r}
# Create an interval
start_date <- ymd("2025-01-01")
end_date <- ymd("2025-07-15")

my_interval <- interval(start_date, end_date)
my_interval
```

```{r}
# The %--% operator is a shorthand for interval()
my_interval <- ymd("2025-01-01") %--% ymd("2025-07-15")
my_interval
```

```{r}
# Get the start and end of an interval
int_start(my_interval)
int_end(my_interval)
```

```{r}
# Get the length of an interval in seconds
int_length(my_interval)
```

```{r}
# Divide an interval by a period to get the number of complete units
my_interval / months(1)    # approximately how many months
my_interval / days(1)      # how many days
my_interval / weeks(1)     # how many weeks
```

### Testing Whether a Date Falls Within an Interval

The `%within%` operator tests whether a date or datetime falls inside an interval.

```{r}
spring_semester <- ymd("2025-01-13") %--% ymd("2025-05-09")

ymd("2025-03-15") %within% spring_semester
ymd("2025-06-01") %within% spring_semester
ymd("2025-01-13") %within% spring_semester  # start date is included
```

```{r}
# Test multiple dates at once
test_dates <- ymd(c("2025-01-01", "2025-02-14", "2025-04-30", "2025-06-15"))
test_dates %within% spring_semester
```

### Comparing and Overlapping Intervals

```{r}
# Do two intervals overlap?
semester1 <- ymd("2025-01-13") %--% ymd("2025-05-09")
semester2 <- ymd("2025-08-25") %--% ymd("2025-12-12")
summer <- ymd("2025-05-10") %--% ymd("2025-08-24")

int_overlaps(semester1, summer)
int_overlaps(semester1, semester2)
```


## Rounding Dates

Lubridate provides three functions for rounding dates to a specified unit: `floor_date()`, `ceiling_date()`, and `round_date()`. These are invaluable for grouping dates into bins --- for example, aggregating daily data to monthly totals.

```{r}
dt <- ymd_hms("2025-06-15 14:35:22")

# Round down (floor) to various units
floor_date(dt, unit = "year")
floor_date(dt, unit = "month")
floor_date(dt, unit = "week")
floor_date(dt, unit = "day")
floor_date(dt, unit = "hour")
floor_date(dt, unit = "minute")
```

```{r}
# Round up (ceiling)
ceiling_date(dt, unit = "month")
ceiling_date(dt, unit = "hour")
```

```{r}
# Round to nearest
round_date(dt, unit = "hour")    # 14:35 rounds to 15:00
round_date(dt, unit = "month")   # June 15 rounds to July (past midpoint)
```

### Rounding with Custom Intervals

You can specify multi-unit intervals for rounding.

```{r}
dt <- ymd_hms("2025-06-15 14:35:22")

# Round to the nearest 15-minute interval
floor_date(dt, unit = "15 minutes")
ceiling_date(dt, unit = "15 minutes")

# Round to the nearest 6-hour block
floor_date(dt, unit = "6 hours")

# Round to 2-month intervals
floor_date(dt, unit = "2 months")
```

### Rounding in Data Pipelines

Rounding is especially useful for aggregating data by time period.

```{r}
# Daily transactions aggregated to monthly totals
transactions <- tibble(
  date = ymd(c(
    "2025-01-05", "2025-01-18", "2025-01-29",
    "2025-02-03", "2025-02-14", "2025-02-27",
    "2025-03-08", "2025-03-15", "2025-03-30"
  )),
  revenue = c(1200, 950, 1100, 1350, 800, 1500, 1050, 1200, 900)
)

transactions |>
  mutate(month = floor_date(date, unit = "month")) |>
  group_by(month) |>
  summarize(
    total_revenue = sum(revenue),
    n_transactions = n()
  )
```


## Time Zones

Time zones are one of the most confusing aspects of working with date-time data. Lubridate provides two key functions for dealing with them, and understanding the difference between them is critical.

### Viewing Time Zones

```{r}
# The current time in different zones
now("America/New_York")
now("America/Los_Angeles")
now("Europe/London")
now("Asia/Tokyo")
```

```{r}
# List a sample of available time zones (there are hundreds)
head(OlsonNames(), 20)
```

```{r}
# How many time zones are available?
length(OlsonNames())
```

### with_tz(): Change the Display

The `with_tz()` function changes how a datetime is **displayed** without changing the underlying instant in time. It answers the question: "What time is it in another time zone at this same moment?"

```{r}
meeting_time <- ymd_hms("2025-06-15 10:00:00", tz = "America/New_York")
meeting_time

# What time is this meeting in London?
with_tz(meeting_time, tzone = "Europe/London")

# What time is this meeting in Tokyo?
with_tz(meeting_time, tzone = "Asia/Tokyo")

# What time is this meeting in Los Angeles?
with_tz(meeting_time, tzone = "America/Los_Angeles")
```

The underlying moment in time is the same in all four cases --- only the display changes.

### force_tz(): Change the Clock Time

The `force_tz()` function changes the **time zone label** without changing the clock-face time. It answers the question: "This datetime was recorded without a time zone, and I know it was actually in this zone."

```{r}
# Suppose a time was recorded as UTC but was actually Eastern Time
recorded <- ymd_hms("2025-06-15 10:00:00", tz = "UTC")
recorded

# Correct the time zone: the clock reading 10:00 was actually in New York
corrected <- force_tz(recorded, tzone = "America/New_York")
corrected

# Notice the underlying instant has changed
with_tz(recorded, tzone = "America/New_York")   # with_tz: same instant, different display
with_tz(corrected, tzone = "UTC")               # force_tz changed the actual instant
```

The distinction is crucial:

- **`with_tz()`**: Same instant, different label. Use when converting for display.
- **`force_tz()`**: Different instant, same clock reading. Use when correcting a mislabeled time zone.


## Practical Examples

Let us put lubridate's tools to work in realistic data analysis scenarios, integrating with dplyr and ggplot2.

### Analyzing the Economics Dataset

The built-in `economics` dataset from ggplot2 contains monthly US economic data from 1967 to 2015. The `date` column is already a Date object, making it an excellent playground for lubridate.

```{r}
data(economics)
glimpse(economics)
```

```{r}
# Add time components for analysis
econ_enhanced <- economics |>
  mutate(
    year = year(date),
    month = month(date, label = TRUE),
    decade = floor_date(date, unit = "10 years")
  )

head(econ_enhanced)
```

```{r}
# Average unemployment rate by decade
econ_enhanced |>
  group_by(decade) |>
  summarize(
    avg_unemploy = mean(unemploy),
    avg_pop = mean(pop),
    n_months = n()
  )
```

```{r}
# Monthly unemployment trend (averaged across all years)
econ_enhanced |>
  group_by(month) |>
  summarize(avg_unemploy = mean(unemploy)) |>
  ggplot(aes(x = month, y = avg_unemploy)) +
  geom_col(fill = "steelblue") +
  labs(
    title = "Average US Unemployment by Month (1967-2015)",
    x = "Month",
    y = "Unemployment (thousands)"
  ) +
  theme_minimal()
```

```{r}
# Unemployment over time, colored by decade
economics |>
  mutate(decade = factor(10 * (year(date) %/% 10))) |>
  ggplot(aes(x = date, y = unemploy, color = decade)) +
  geom_line() +
  labs(
    title = "US Unemployment Over Time",
    x = "Date",
    y = "Unemployment (thousands)",
    color = "Decade"
  ) +
  theme_minimal()
```

### Calculating Age from a Birthdate

A common task is computing a person's age from their date of birth. This requires an interval divided by years.

```{r}
people <- tibble(
  name = c("Alice", "Bob", "Carol", "David", "Eve"),
  birthdate = ymd(c("1990-05-20", "1985-11-03", "2000-02-29", "1978-08-15", "1995-12-01"))
)

people |>
  mutate(
    age = as.numeric(interval(birthdate, today()) / years(1)),
    age_years = floor(age)
  )
```

Note the handling of Carol's leap year birthday --- the interval-based calculation correctly accounts for this edge case.

### Time Between Events

Calculating the elapsed time between successive events is a pattern that appears in many domains --- customer transactions, medical visits, machine maintenance logs.

```{r}
service_log <- tibble(
  machine_id = c("A", "A", "A", "B", "B", "B", "B"),
  service_date = ymd(c(
    "2025-01-10", "2025-03-22", "2025-06-15",
    "2025-01-05", "2025-02-18", "2025-05-01", "2025-07-20"
  )),
  service_type = c("routine", "repair", "routine",
                   "routine", "routine", "repair", "routine")
)

service_log |>
  group_by(machine_id) |>
  arrange(service_date) |>
  mutate(
    days_since_last = as.numeric(service_date - lag(service_date)),
    cumulative_days = as.numeric(service_date - first(service_date))
  )
```

### Working with Business Days and Weekends

Lubridate makes it easy to identify weekdays and weekends.

```{r}
# Generate a sequence of dates
dates_in_june <- tibble(
  date = seq(ymd("2025-06-01"), ymd("2025-06-30"), by = "day")
)

dates_in_june |>
  mutate(
    day_name = wday(date, label = TRUE, abbr = FALSE),
    is_weekend = wday(date) %in% c(1, 7),
    day_type = if_else(is_weekend, "Weekend", "Weekday")
  ) |>
  count(day_type)
```

### Grouping and Summarizing by Time Periods

```{r}
# Simulate daily website traffic
set.seed(42)
web_traffic <- tibble(
  date = seq(ymd("2025-01-01"), ymd("2025-12-31"), by = "day"),
  visits = rpois(365, lambda = 500) + round(100 * sin(2 * pi * (1:365) / 365))
)

# Weekly summary
web_traffic |>
  mutate(week = floor_date(date, unit = "week")) |>
  group_by(week) |>
  summarize(
    total_visits = sum(visits),
    avg_daily_visits = mean(visits),
    peak_day = date[which.max(visits)]
  ) |>
  head(10)
```

```{r}
# Monthly summary with visualization
web_traffic |>
  mutate(month = floor_date(date, unit = "month")) |>
  group_by(month) |>
  summarize(total_visits = sum(visits)) |>
  ggplot(aes(x = month, y = total_visits)) +
  geom_col(fill = "darkorange") +
  scale_x_date(date_labels = "%b", date_breaks = "1 month") +
  labs(
    title = "Monthly Website Traffic (2025)",
    x = "Month",
    y = "Total Visits"
  ) +
  theme_minimal()
```

### Creating Date Sequences

The base R `seq()` function works with lubridate dates to create regular sequences.

```{r}
# Daily sequence
seq(ymd("2025-01-01"), ymd("2025-01-10"), by = "day")
```

```{r}
# Weekly sequence
seq(ymd("2025-01-01"), ymd("2025-03-31"), by = "week")
```

```{r}
# Monthly sequence
seq(ymd("2025-01-01"), ymd("2025-12-01"), by = "month")
```

```{r}
# Custom interval: every 2 weeks
seq(ymd("2025-01-01"), ymd("2025-06-30"), by = "2 weeks")
```

### Filtering Data by Date Ranges

Combining lubridate with dplyr's `filter()` and `between()` makes date-range queries clean and readable.

```{r}
# Filter the economics dataset to the 2000s decade
economics |>
  filter(date >= ymd("2000-01-01"), date < ymd("2010-01-01")) |>
  head()
```

```{r}
# Filter to a specific quarter
economics |>
  filter(year(date) == 2008, quarter(date) == 4) |>
  head()
```

```{r}
# Use between() for date ranges
economics |>
  filter(between(date, ymd("2007-06-01"), ymd("2009-06-01"))) |>
  ggplot(aes(x = date, y = unemploy)) +
  geom_line(color = "firebrick", linewidth = 1) +
  labs(
    title = "US Unemployment During the Great Recession",
    x = "Date",
    y = "Unemployment (thousands)"
  ) +
  theme_minimal()
```


## Summary

The lubridate package transforms date and time manipulation from a frustrating chore into a straightforward task. Its key contributions include:

- **Parsing functions** (`ymd()`, `mdy()`, `dmy()`, and their `_hms` variants) that automatically detect date formats based on component order.
- **Component extractors** (`year()`, `month()`, `mday()`, `wday()`, `hour()`, `minute()`, `second()`) that pull apart dates and also serve as setters.
- **Three types of time spans**: durations (exact seconds), periods (human-friendly calendar units), and intervals (start-to-end spans).
- **Rounding functions** (`floor_date()`, `ceiling_date()`, `round_date()`) for binning dates into time periods.
- **Time zone tools** (`with_tz()` for display conversion, `force_tz()` for correcting mislabeled zones).

When working with dates in your data analyses, lubridate integrates seamlessly with dplyr for data wrangling and ggplot2 for visualization. The patterns demonstrated in this chapter --- extracting components, computing intervals, grouping by time period, and filtering by date ranges --- will cover the vast majority of date-related tasks you encounter in practice.
