# Advanced R Markdown {#rmarkdown-advanced}

```{r include=FALSE}
library(tidyverse)
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

You survived the R Markdown basics in Chapter \@ref(rmarkdown). You can knit HTML documents, add code chunks, and format text. Congratulations --- you now know enough to be dangerous. But R Markdown has a lot more tricks up its sleeve, and some of them are genuinely impressive in a business context.

Think of this chapter as the "bonus features" menu on a DVD (if you remember those). We are not going to master every single option here. Instead, we will walk through the coolest things R Markdown can do so that when your boss says "Can you make that report run automatically for all 12 regions?" or "Can we get a dashboard instead of a Word doc?" you can say "Yeah, give me an hour."

## Parameterized Reports {#parameterized-reports}

Here is a scenario every business student will eventually face: you write a beautiful report for the Northeast region, and your manager says, "Great! Now make one for every other region too." If you copy-paste the `.Rmd` file twelve times and change the region name in each one, you will hate your life by region number four. Parameterized reports solve this problem.

The idea is simple: you write your report once, but leave a placeholder (a "parameter") for anything that changes --- like the region name, the product category, or the time period. Then you tell R to generate the report over and over, swapping in different values each time. It is basically a mail merge, but for data analysis.

### Setting It Up

You add a `params` section to your YAML header with default values. Here is a complete, self-contained example you could save as `species-report.Rmd` and actually run:

````
---
title: "Species Report"
author: "Your Name"
date: "`r knitr::inline_expr('Sys.Date()')`"
params:
  species_name: "setosa"
output:
  html_document:
    toc: true
    theme: flatly
---

```{r setup, include=FALSE}`r ''`
library(tidyverse)
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE)
```

## Overview

This report covers the **`r knitr::inline_expr('params$species_name')`** species
from the `iris` dataset.

```{r}`r ''`
species_data <- iris %>%
  filter(Species == params$species_name)
```

The dataset contains `r knitr::inline_expr('nrow(species_data)')` observations.

## Summary Statistics

```{r}`r ''`
summary(species_data %>% select(-Species))
```

## Visualization

```{r fig.width=6, fig.height=4}`r ''`
ggplot(species_data, aes(x = Sepal.Length, y = Sepal.Width)) +
  geom_point(color = "steelblue", size = 2) +
  labs(
    title = paste("Sepal Dimensions:", params$species_name),
    x = "Sepal Length (cm)",
    y = "Sepal Width (cm)"
  ) +
  theme_minimal()
```
````

Notice how `params$species_name` shows up everywhere --- in the filtering code, in the plot title, and even in the narrative text via inline R code. When you knit this document, R fills in the default value ("setosa"). But the real power comes from overriding that default.

To render with a different species, use `rmarkdown::render()`:

```{r eval=FALSE}
rmarkdown::render(
  input = "species-report.Rmd",
  params = list(species_name = "versicolor"),
  output_file = "report-versicolor.html"
)
```

In a real business setting, swap "species_name" for "region", "product_line", "client_name", or whatever dimension your reports are organized around. Same concept, much higher stakes.

### Generating Reports in Bulk

Here is where it gets fun. Want to generate a report for every region in one shot? Just loop through them:

```{r eval=FALSE}
regions <- c("Northeast", "Southeast", "Midwest", "West")

for (r in regions) {
  rmarkdown::render(
    input = "regional-report.Rmd",
    params = list(region = r),
    output_file = paste0("report-", r, ".html")
  )
}
```

Run that, go get coffee, and come back to four finished reports. Your manager thinks you are a wizard. You are not --- you just know about parameterized reports.

## Presentations {#rmarkdown-presentations}

Yes, you can make slide decks with R Markdown. They will not replace PowerPoint for every group project, but for data-driven presentations where your charts need to match your analysis exactly, this is a game-changer.

The easiest option is **ioslides**, which gives you HTML slides that open in any browser:

````
---
title: "Q4 Sales Review"
author: "Your Name"
date: "`r knitr::inline_expr('Sys.Date()')`"
output:
  ioslides_presentation:
    widescreen: true
---
````

New slides start with `##` (level 2 headers) or `---`. Your R code chunks produce output right on the slides. You can set `widescreen: true` for a 16:9 aspect ratio (which is what most projectors use these days), and everything else works exactly like a normal R Markdown document. Same syntax, different output.

If you want fancier slides with more customization, check out the `xaringan` package. It produces gorgeous, highly customizable HTML presentations and is popular for conference talks. It supports two-column layouts, incremental reveals, presenter notes, and custom CSS themes. But honestly, ioslides will get you through any class presentation just fine --- save xaringan for when you are trying to impress someone at a conference.

## Dashboards with flexdashboard {#dashboards-flexdashboard}

Dashboards are the business world's favorite way to consume data. Executives love them. Clients love them. Your professor will love them. The `flexdashboard` package lets you build dashboards using the same R Markdown skills you already have.

```{r eval=FALSE}
install.packages("flexdashboard")
```

### The Basic Structure

A flexdashboard is just an R Markdown document with a special output format. The layout uses headers to define columns and boxes:

````
---
title: "Sales Dashboard"
output:
  flexdashboard::flex_dashboard:
    orientation: columns
    vertical_layout: fill
---

```{r setup, include=FALSE}`r ''`
library(flexdashboard)
library(tidyverse)
```

Column {data-width=650}
-----------------------------------------------------------------------

### Revenue by Region

```{r}`r ''`
ggplot(sales, aes(x = Region, y = Revenue, fill = Region)) +
  geom_col() +
  theme_minimal()
```

Column {data-width=350}
-----------------------------------------------------------------------

### Top Products

```{r}`r ''`
sales %>%
  group_by(Product) %>%
  summarize(Total = sum(Revenue)) %>%
  arrange(desc(Total)) %>%
  head(10) %>%
  knitr::kable()
```

### Total Revenue

```{r}`r ''`
sum(sales$Revenue)
```
````

That is it. Level 2 headers (`##` or the line of dashes) create columns. Level 3 headers (`###`) create boxes within those columns. The `{data-width=650}` bit controls how wide each column is relative to the others.

You can also add multiple pages with level 1 headers, and tabbed sections by adding `{.tabset}` to a column header. Flexdashboard even has special components for highlighting key metrics:

```{r eval=FALSE}
# A big number with an icon --- perfect for KPIs
valueBox(value = 150, caption = "Total Sales", icon = "fa-dollar-sign", color = "primary")

# A gauge/dial that shows progress toward a goal
gauge(value = 72, min = 0, max = 100, symbol = "%",
      gaugeSectors(success = c(80, 100), warning = c(50, 79), danger = c(0, 49)))
```

These are the kind of flashy elements that executives cannot resist. "What's our conversion rate?" --- boom, there it is in a big green number on the dashboard.

If you want to go even further, adding `runtime: shiny` to your YAML header turns a static dashboard into a fully interactive app with dropdown menus, sliders, and reactive charts. That requires a running R session (so you cannot just email the HTML file), but it is incredibly powerful for internal tools.

## Interactive Documents {#interactive-documents}

Here is where you really impress people. With just a couple of extra packages, you can make your HTML reports interactive --- sortable tables, zoomable plots, the works. No Shiny server required. These are self-contained HTML files you can email to anyone.

### Interactive Tables with DT

The `DT` package turns boring static tables into searchable, sortable, paginated tables that your audience can actually explore:

```{r eval=FALSE}
install.packages("DT")
```

```{r}
library(DT)
datatable(
  iris,
  options = list(pageLength = 10, autoWidth = TRUE),
  caption = "Go ahead --- search, sort, and filter this table.",
  filter = "top"
)
```

Your audience can search across all columns, sort by clicking headers, filter individual columns, and page through the data. It is like giving them a mini-spreadsheet inside your report. Finance people go absolutely nuts for this.

### Interactive Plots with plotly

The `plotly` package is pure magic. You take a regular ggplot, wrap it in `ggplotly()`, and suddenly it has hover tooltips, zooming, and panning:

```{r eval=FALSE}
install.packages("plotly")
```

```{r}
library(plotly)

p <- ggplot(iris, aes(x = Sepal.Length, y = Sepal.Width, color = Species)) +
  geom_point(size = 2) +
  labs(title = "Iris Sepal Dimensions",
       x = "Sepal Length (cm)",
       y = "Sepal Width (cm)") +
  theme_minimal()

ggplotly(p)
```

Hover over any point to see its exact values. Click and drag to zoom in. Use the toolbar to pan, reset the view, or download the plot as a PNG. One line of code (`ggplotly(p)`) turns a static chart into something that feels like a real data product. This is the single easiest way to make your reports look impressive.

### Other Interactive Goodies

The `leaflet` package creates interactive maps (great for any location-based business data), and there are htmlwidgets for network diagrams, time series, and more. Once you are comfortable with `DT` and `plotly`, exploring these is easy because they all follow the same basic pattern.

## A Few Other Things Worth Knowing {#other-advanced-features}

**Bookdown** is the package that built the book you are reading right now. It extends R Markdown to handle multi-chapter documents with cross-referencing across chapters. You are already using it, so you know how it works.

**Blogdown** lets you build a website or blog using R Markdown. If you ever want a personal data science portfolio site, it is worth a look.

Both are great tools, but you do not need to learn them in depth right now.

## Tips for Reproducibility {#tips-reproducible-research}

Reproducibility means someone else (or future you, six months from now, panicking before a deadline) can run your analysis and get the same results. In a business context, this is the difference between "I can back up these numbers" and "I have no idea how I got these numbers."

Two simple habits will save you:

### Set Your Seeds

Any time your analysis involves randomness --- sampling, simulations, bootstrapping --- use `set.seed()` so the results are identical every time:

```{r}
set.seed(42)
sample(1:100, 5)
```

```{r}
# Same seed, same results. Every time.
set.seed(42)
sample(1:100, 5)
```

Put `set.seed()` in your setup chunk and you are good.

### Record Your Session Info

At the end of important documents, include `sessionInfo()`. This records which version of R and which packages you used, so if something breaks later, you can figure out what changed:

```{r}
sessionInfo()
```

It is not glamorous, but the day you need it, you will be very glad it is there. Think of it as an insurance policy for your analysis.

### Keep Your Project Organized

A quick checklist for reproducible work:

- **Use RStudio Projects** (the `.Rproj` file) so your working directory is always the project root.
- **Never modify raw data.** Keep originals in a `data/raw/` folder and write code that creates cleaned versions.
- **Use relative file paths.** Write `read_csv("data/sales.csv")`, not `read_csv("C:/Users/jsmith/Desktop/stuff/sales.csv")`. Your coworkers will thank you.
- **Knit from a clean session.** Before you share a report, restart R and knit the whole thing from scratch. If it does not work in a clean session, it does not work.

If you followed the advice in earlier chapters, you are already doing most of this. The point is: future you should be able to open this project in six months and have everything just run.

## Summary {#rmarkdown-advanced-summary}

R Markdown is not just for homework reports. Here is what we covered:

- **Parameterized reports** let you generate the same report for different regions, products, or time periods automatically. Write once, render many.
- **Presentations** with ioslides (and xaringan for the adventurous) let you build slide decks straight from your analysis.
- **Flexdashboard** turns R Markdown into dashboards that look like actual business intelligence tools.
- **Interactive documents** with `DT::datatable()` and `plotly::ggplotly()` make your reports feel like apps --- searchable tables, zoomable charts, no server required.
- **Reproducibility basics** like `set.seed()` and `sessionInfo()` keep your work trustworthy and your future self sane.

None of these are things you need to master today. But now you know they exist, and when the right situation comes up --- and it will --- you will know exactly where to look.
